---
title: "Some Usual Distributions"
author: "PB"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**NOTE**: Only some basic distributions and properties are mentioned. The Wikipedia pages on these distributions are a great resource.

-------

## Gaussian Distribution

### Definition

Let $X$ be a Gaussian random variable with expectation $\mu$ and variance $\sigma^2$:
$$
X \sim \mathcal{N}(\mu, \sigma^2).
$$

It admits a probability density function (pdf):
$$
f(x) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{(x - \mu)^2}{2\sigma^2}\right)
\quad
\forall x \in \mathbb{R},
$$

And a characteristic function:
$$
\phi_{X}(t) = \mathbb{E}[e^{itX}] = e^{i\mu t - \frac{\sigma^2 t^2}{2}}
\quad
\forall t \in \mathbb{R}.
$$

### Density plots

```{r gaussian, echo=FALSE, fig.height=5.5, fig.width=8}
x <- -1500:1500 / 100
ccc <- hcl.colors(4)
# Normal
mu <- 0; sigma2 <- 1
plot(x, dnorm(x, mu, sigma2), type = "l", ylab = "f(x)", col = ccc[1], lwd = 2)
segments(x0 = 0, y0 = 0, x1 = 0, y1 = dnorm(mu, mu, sigma2), lty = 2, col = ccc[1])
# mean
mu <- 10; sigma2 <- 1
lines(x, dnorm(x, mu, sigma2), col = ccc[2], lwd = 2)
segments(x0 = mu, y0 = 0, x1 = mu, y1 = dnorm(mu, mu, sigma2), lty = 2, col = ccc[2])
# variance
mu <- 0; sigma2 <- 5
lines(x, dnorm(x, mu, sigma2), col = ccc[3], lwd = 2)
segments(x0 = mu, y0 = 0, x1 = mu, y1 = dnorm(mu, mu, sigma2), lty = 2, col = ccc[3])
# legend
legend("topleft",
       c(expression(paste(mu == 0, "   ; ", sigma^2 == 1)),
         expression(paste(mu == 10, " ; ", sigma^2 == 1)),
         expression(paste(mu == 0, "   ; ", sigma^2 == 25))),
       col = ccc,
       lwd = 2, lty = 1)
```

### Moments

$$
\mathbb{E}[X] = \mu
\quad
\mathbb{Var}[X] = \sigma^2.
$$

### Some Properties

* The Gaussian distribution with mean $0$ and variance $1$
is called the *standard normal* distribution.

* If $X \sim \mathcal{N}(\mu, \sigma^2)$, then for any $(a, b) \in \mathbb{R}^2$, then
$Y = aX + b \sim \mathcal{N}(a\mu + b, a^2\sigma^2).$

* The *mode*, *median* and *mean* of the distribution are all equal to $\mu$.

* If $X_1 \sim \mathcal{N}(\mu_1, \sigma^2_1)$ and $X_2 \sim \mathcal{N}(\mu_2, \sigma^2_2)$ are
**independent** Gaussian random variables, then their sum $Y = X_1 + X_2$  is also Gaussian, 
with mean $\mu_1 + \mu_2$ and variance $\sigma^2_1 + \sigma^2_2$:
$Y \sim \mathcal{N}(\mu_1 + \mu_2, \sigma^2_1 + \sigma^2_2)$

* ... many more (see e.g. the [wikipedia](https://en.wikipedia.org/wiki/Normal_distribution) page).

### `R` functions

The density function of the Gaussian distribution in `R` is `dnorm`.

For information on the distribution function (`pnorm`),
quantile function (`qnorm`) and random generation (`rnorm`),
see:

```{r gaussianfun}
help(dnorm)
```

-------

## Chi Squared Distribution

### Definition

Let $X_1, \dotsc, X_p$ be $p$ standard normal independent identically distributed
(iid) random variables: $X_i \sim \mathcal{N}(0, 1)$.
Then 
$$
X = \sum_{i = 1}^p X_i^2 \sim \chi^2_p
$$ 

is a Chi squared random variable with $p$ degrees of freedom.

### Density Plots

```{r chisq, echo=FALSE, fig.height=5.5, fig.width=8}
x <- 0:800 / 100
ccc <- hcl.colors(6)
# 1
p <- 1
plot(x, dchisq(x, p), type = "l", ylab = "f(x)", ylim = c(0, 0.5), col = ccc[1], lwd = 2)
segments(x0 = p, y0 = -1, x1 = p, y1 = dchisq(p, p), lty = 2, col = ccc[1])
# 2 - 5
for (p in 2:5) {
  lines(x, dchisq(x, p), col = ccc[p], lwd = 2)
  segments(x0 = p, y0 = -1, x1 = p, y1 = dchisq(p, p), lty = 2, col = ccc[p])
}
# legend
legend("topright",
       c(expression(p == 1), expression(p == 2), expression(p == 3), expression(p == 4), expression(p == 5)),
       col = ccc,
       lwd = 2, lty = 1)
```

### Moments

$$
\mathbb{E}[X] = p
\quad
\mathbb{Var}[X] = 2p.
$$

### Some Properties

* The chi squared distribution converges towards the Gaussian distribution with matching moments:  
if $X \sim \chi^2_{p}$, then the distribution of $\frac{X - p}{\sqrt{2p}}$ converges toward a standard normal distribution when $p$ goes to infinity.

```{r chisq2, echo=FALSE, fig.height=5.5, fig.width=8}
x <- 0:10000 / 100
# 50
p <- 50
plot(x, dchisq(x, p), type = "l", ylab = "f(x)", ylim = c(0, 0.05), col = ccc[1], lwd = 2)
segments(x0 = p, y0 = -1, x1 = p, y1 = dchisq(p, p), lty = 2, col = ccc[1])
# normal
mu <- p; sigma2 <- sqrt(2*p)
lines(x, dnorm(x, mu, sigma2), col = ccc[4])
# legend
legend("topright",
       c(expression(p == 50),
         expression(paste(mu == 50, "   ; ", sigma^2 == 100))),
       col = ccc[c(1, 4)],
       lwd = c(2, 1), lty = 1)
```

* If $X_1 \sim \chi^2_{p_1}$ and $X_2 \sim \chi^2_{p_2}$ are
**independent** chi squared random variables with respective degrees of freedom $p_1$ and $p_2$,
then their sum $Y = X_1 + X_2$  is also a chi squared random variable, 
with $p_1 + p_2$ degrees of freedom:
$Y \sim \chi^2_{p_1+p_2}$

* ... (see e.g. the [wikipedia](https://en.wikipedia.org/wiki/Chi-square_distribution) page).

### `R` functions

The density function of the chi squared distribution in `R` is `dchisq`.

For information on the distribution function (`pchisq`),
quantile function (`qchisq`) and random generation (`rchisq`),
see:

```{r chifun}
help(dchisq)
```

-------

## Student Distribution

### Definition

Let $Z$ be a standard normal random variable : $Z \sim \mathcal{N}(0, 1)$,
and $X$ a chi squared random variable with $p$ degrees of freedom: $Z \sim \chi^2_p$,
$Z$ and $X$ independent.
Then 
$$
T = \frac{Z}{\sqrt{X/p}} \sim \mathcal{T}_p
$$ 

is a Student (t) random variable with $p$ degrees of freedom.

### Density Plots

```{r student, echo=FALSE, fig.height=5.5, fig.width=8}
x <- -600:600 / 100
ccc <- hcl.colors(6)
# 1
p <- 1
plot(x, dt(x, p), type = "l", ylab = "f(x)", ylim = c(0, 0.4), col = ccc[1], lwd = 2)
segments(x0 = 0, y0 = -1, x1 = 0, y1 = dt(0, p), lty = 2, col = ccc[1])
# 2 - 5
k <- 1
for (p in c(3, 10)) {
  k <- k + 1
  lines(x, dt(x, p), col = ccc[k], lwd = 2)
}
# normal
mu <- 0; sigma2 <- 1
lines(x, dnorm(x, mu, sigma2), col = ccc[5], lwd = 2)
# legend
legend("topright",
       c(expression(p == 1), expression(p == 3), expression(p == 10),
         expression(paste(mu == 0, "   ; ", sigma^2 == 1))),
       col = ccc[c(1:k, 5)],
       lwd = 2, lty = 1)
```

### Moments

$$
\mathbb{E}[T] = 0 ~ (\text{for } p > 1)
\quad
\mathbb{Var}[T] = \frac{p}{p-2} ~ (\text{for } p > 2).
$$

### Some Properties

* The student distribution converges towards the standard normal distribution:  
if $T \sim \mathcal{T}_{p}$, then the distribution of $T$ converges towards
a standard normal distribution when $p$ goes to infinity.

* ... (see e.g. the [wikipedia](https://en.wikipedia.org/wiki/Student%27s_t-distribution) page).

### `R` functions

The density function of the Student distribution in `R` is `dt`.

For information on the distribution function (`pt`),
quantile function (`qt`) and random generation (`rt`),
see:

```{r studentfun}
help(dt)
```

-------

## Fisher Distribution

### Definition

Let $U_1 \sim \chi^2_{p_1}$ and $U_2 \sim \chi^2_{p_2}$, be two chi squared random variables,
$U_1$ and $U_2$ independent.
Then 
$$
F = \frac{U_1/p_1}{U_2/p_2} \sim \mathcal{F}^{p_1}_{p_2}
$$ 

is a Fisher (F) random variable with $(p_1, p_2)$ degrees of freedom.

### Density Plots

```{r fisher, echo=FALSE, fig.height=5.5, fig.width=8}
x <- 0:600 / 100
ccc <- hcl.colors(6)
# 1
p1 <- 2
p <- 3
plot(x, df(x, p1, p), type = "l", ylab = "f(x)", ylim = c(0, 1), col = ccc[1], lwd = 2)
segments(x0 = p/(p-2), y0 = -1, x1 = p/(p-2), y1 = df(p/(p-2), p1, p), lty = 2, col = ccc[1])
# 2 - 5
k <- 1
for (p in c(5, 10)) {
  k <- k + 1
  lines(x, df(x, p1, p), col = ccc[k], lwd = 2)
  segments(x0 = p/(p-2), y0 = -1, x1 = p/(p-2), y1 = df(p/(p-2), p1, p), lty = 2, col = ccc[k])
}
# normal
lines(x, 2 * dchisq(2 * x, p1), col = ccc[5], lwd = 2)
# legend
legend("topright",
       c(expression(paste(p[1] == 2, "   ; ", p[2] == 3)),
         expression(paste(p[1] == 2, "   ; ", p[2] == 5)),
         expression(paste(p[1] == 2, "   ; ", p[2] == 10)),
         expression(paste(p == 2, " [", chi[p]^2 / p, "]"))),
       col = ccc[c(1:k, 5)],
       lwd = 2, lty = 1)
```

### Moments

$$
\mathbb{E}[F] = \frac{p_2}{p_2 - 2} ~ (\text{for } p_2 > 2)
\quad
\mathbb{Var}[F] = \frac{2p_2^2(p_1 + p_2 - 2)}{p_1(p_2-2)^2(p_2-4)} ~ (\text{for } p_2 > 4)
$$

### Some Properties

* If $T \sim \mathcal{T}_{p}$, then $T^2 \sim \mathcal{F}^{1}_{p}$.

* If $F \sim \mathcal{F}^{p_1}_{p_2}$, then $F^{-1} \sim \mathcal{F}^{p_2}_{p_1}$.

* If $F \sim \mathcal{F}^{p_1}_{p_2}$, then the distribution of $p_1 F$ converges 
to the chi squared distribution $\chi^2_{p_1}$ when $p_2$ goes to infinity.

* ... (see e.g. the [wikipedia](https://en.wikipedia.org/wiki/F-distribution) page).

### `R` functions

The density function of the Student distribution in `R` is `df`.

For information on the distribution function (`pf`),
quantile function (`qf`) and random generation (`rf`),
see:

```{r fisherfun}
help(df)
```