---
title: "Some Usual Distributions"
author: "Paul Bastide"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**NOTE**: To be completed. The Wikipedia pages on these distributions are a great resource.

## Gaussian Distribution

### Definition

Let $X$ be a Gaussian random variable with expectation $\mu$ and variance $\sigma^2$:
$$
X \sim \mathcal{N}(\mu, \sigma^2).
$$

It admits a probability density function (pdf):
$$
f(x) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{(x - \mu)^2}{2\sigma^2}\right)
\quad
\forall x \in \mathbb{R},
$$

And a characteristic function:
$$
\phi_{X}(t) = \mathbb{E}[e^{itX}] = e^{i\mu t - \frac{\sigma^2 t^2}{2}}
\quad
\forall t \in \mathbb{R}.
$$

### Density plots

```{r gaussian, echo=FALSE, fig.height=5.5, fig.width=8}
x <- -1500:1500 / 100
ccc <- hcl.colors(4)
# Normal
mu <- 0; sigma2 <- 1
plot(x, dnorm(x, mu, sigma2), type = "l", ylab = "f(x)", col = ccc[1], lwd = 2)
segments(x0 = 0, y0 = 0, x1 = 0, y1 = dnorm(mu, mu, sigma2), lty = 2, col = ccc[1])
# mean
mu <- 10; sigma2 <- 1
lines(x, dnorm(x, mu, sigma2), col = ccc[2], lwd = 2)
segments(x0 = mu, y0 = 0, x1 = mu, y1 = dnorm(mu, mu, sigma2), lty = 2, col = ccc[2])
# variance
mu <- 0; sigma2 <- 5
lines(x, dnorm(x, mu, sigma2), col = ccc[3], lwd = 2)
segments(x0 = mu, y0 = 0, x1 = mu, y1 = dnorm(mu, mu, sigma2), lty = 2, col = ccc[3])
# legend
legend("topleft",
       c(expression(paste(mu == 0, "   ; ", sigma^2 == 1)),
         expression(paste(mu == 10, " ; ", sigma^2 == 1)),
         expression(paste(mu == 0, "   ; ", sigma^2 == 25))),
       col = ccc,
       lwd = 2, lty = 1)
```


### Some Properties

* If $X \sim \mathcal{N}(\mu, \sigma^2)$, then for any $(a, b) \in \mathbb{R}^2$, then
$Y = aX + b \sim \mathcal{N}(a\mu + b, a^2\sigma^2).$

## Student Distribution

### Definition

Let $Z$ be a standard normal rv : $Z \sim \mathcal{N}(0, 1)$,  
and $X$ a chi squared rv with $p$ degrees of freedom: $Z \sim \chi^2_p$,
$Z$ and $X$ independent.
Then 
$$
T = \frac{Z}{\sqrt{X/p}} \sim \mathcal{T}_p
$$ 

is a Student r.v. with $p$ degrees of freedom.

### Density Plots

```{r student, echo=FALSE, fig.height=5.5, fig.width=8}
x <- -600:600 / 100
ccc <- hcl.colors(6)
# 1
p <- 1
plot(x, dt(x, p), type = "l", ylab = "f(x)", ylim = c(0, 0.4), col = ccc[1], lwd = 2)
segments(x0 = 0, y0 = -1, x1 = 0, y1 = dt(0, p), lty = 2, col = ccc[1])
# 2 - 5
k <- 1
for (p in c(3, 10)) {
  k <- k + 1
  lines(x, dt(x, p), col = ccc[k], lwd = 2)
}
# normal
mu <- 0; sigma2 <- 1
lines(x, dnorm(x, mu, sigma2), col = ccc[5], lwd = 2)
# legend
legend("topright",
       c(expression(p == 1), expression(p == 3), expression(p == 10),
         expression(paste(mu == 0, "   ; ", sigma^2 == 1))),
       col = ccc[c(1:k, 5)],
       lwd = 2, lty = 1)
```

### Properties

* For $p \geq 3$:
$$
\mathbb{E}[T] = 0
\quad
\mathbb{Var}[T] = \frac{p}{p-2}.
$$

* It converges in distribution towards the standard Gaussian.

## Chi Squared Distribution

### Definition

Let $X_1, \dotsc, X_p$ be $p$ standard normal iid rv : $X_i \sim \mathcal{N}(0, 1)$.
Then 
$$
X = \sum_{i = 1}^p X_i \sim \chi^2_p
$$ 

is a Chi squared r.v. with $p$ degrees of freedom.

### Density Plots

```{r chisq, echo=FALSE, fig.height=5.5, fig.width=8}
x <- 0:800 / 100
ccc <- hcl.colors(6)
# 1
p <- 1
plot(x, dchisq(x, p), type = "l", ylab = "f(x)", ylim = c(0, 0.5), col = ccc[1], lwd = 2)
segments(x0 = p, y0 = -1, x1 = p, y1 = dchisq(p, p), lty = 2, col = ccc[1])
# 2 - 5
for (p in 2:5) {
  lines(x, dchisq(x, p), col = ccc[p], lwd = 2)
  segments(x0 = p, y0 = -1, x1 = p, y1 = dchisq(p, p), lty = 2, col = ccc[p])
}
# legend
legend("topright",
       c(expression(p == 1), expression(p == 2), expression(p == 3), expression(p == 4), expression(p == 5)),
       col = ccc,
       lwd = 2, lty = 1)
```

### Properties

* We have:
$$
\mathbb{E}[X] = p
\quad
\mathbb{Var}[X] = 2p.
$$

* It converges in distribution towards the Gaussian with matching moments.

```{r chisq2, echo=FALSE, fig.height=5.5, fig.width=8}
x <- 0:10000 / 100
# 50
p <- 50
plot(x, dchisq(x, p), type = "l", ylab = "f(x)", ylim = c(0, 0.05), col = ccc[1], lwd = 2)
segments(x0 = p, y0 = -1, x1 = p, y1 = dchisq(p, p), lty = 2, col = ccc[1])
# normal
mu <- p; sigma2 <- sqrt(2*p)
lines(x, dnorm(x, mu, sigma2), col = ccc[4])
# legend
legend("topright",
       c(expression(p == 50),
         expression(paste(mu == 50, "   ; ", sigma^2 == 100))),
       col = ccc[c(1, 4)],
       lwd = c(2, 1), lty = 1)
```

## Fisher Distribution

### Definition

Let $U_1 \sim \chi^2_{p_1}$ and $U_2 \sim \chi^2_{p_2}$,
$U_1$ and $U_2$ independent.
Then 
$$
F = \frac{U_1/p_1}{U_2/p_2} \sim \mathcal{F}^{p_1}_{p_2}
$$ 

is a Fisher r.v. with $(p_1, p_2)$ degrees of freedom.

### Density Plots

```{r fisher, echo=FALSE, fig.height=5.5, fig.width=8}
x <- 0:600 / 100
ccc <- hcl.colors(6)
# 1
p1 <- 2
p <- 3
plot(x, df(x, p1, p), type = "l", ylab = "f(x)", ylim = c(0, 1), col = ccc[1], lwd = 2)
segments(x0 = p/(p-2), y0 = -1, x1 = p/(p-2), y1 = df(p/(p-2), p1, p), lty = 2, col = ccc[1])
# 2 - 5
k <- 1
for (p in c(5, 10)) {
  k <- k + 1
  lines(x, df(x, p1, p), col = ccc[k], lwd = 2)
  segments(x0 = p/(p-2), y0 = -1, x1 = p/(p-2), y1 = df(p/(p-2), p1, p), lty = 2, col = ccc[k])
}
# normal
lines(x, 2 * dchisq(2 * x, p1), col = ccc[5], lwd = 2)
# legend
legend("topright",
       c(expression(paste(p[1] == 2, "   ; ", p[2] == 3)),
         expression(paste(p[1] == 2, "   ; ", p[2] == 5)),
         expression(paste(p[1] == 2, "   ; ", p[2] == 10)),
         expression(paste(p == 2, " [", chi[p]^2 / p, "]"))),
       col = ccc[c(1:k, 5)],
       lwd = 2, lty = 1)
```

### Properties

* For $p_2 \geq 3$:
$$
\mathbb{E}[T] =  \frac{p_2}{p_2 - 2}.
$$

* When $p_2$ goes to infinity, $\mathcal{F}^{p_1}_{p_2}$ converges in distribution
to $\frac{1}{p_1}\chi^2_{p_1}$.