---
title: 'Régression Multiple - Modèle Gaussien'
author: "Paul Bastide - Benjamin Charlier"
date: "10/02/2022"
output:
  ioslides_presentation:
    fig_width: 7
    fig_height: 4
  self_contained: true
---
  
<!-- ************************************************************************ -->
# What we know
<!-- ************************************************************************ -->

## Model

### Model:
$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
$$

* $\mathbf{y}$ **random** vector of $n$ responses
* $\mathbf{X}$ **non random** $n\times p$ matrix of predictors
* $\boldsymbol{\epsilon}$ **random** vector of $n$ errors
* $\boldsymbol{\beta}$ **non random, unknown** vector of $p$ coefficients

### Assumptions:
  * (H1) $rg(\mathbf{X}) = p$
  * (H2) $\mathbb{E}[\boldsymbol{\epsilon}] = 0$ and $\mathbb{Var}[\boldsymbol{\epsilon}] = \sigma^2 \mathbf{I}_n$ 

## OLS Estimators

$$ 
\hat{\boldsymbol{\beta}} 
= \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\|\mathbf{y} - \mathbf{X}\boldsymbol{\beta} \|^2
$$
$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}^{T}\mathbf{X})^{-1}\mathbf{X}^{T}\mathbf{y}
\qquad
\mathbb{E}[\hat{\boldsymbol{\beta}}] = \boldsymbol{\beta}
\qquad
\mathbb{Var}[\hat{\boldsymbol{\beta}}] = \sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}.
$$


$$ 
\hat{\mathbf{y}} = \mathbf{X}\hat{\boldsymbol{\beta}} 
= \mathbf{P}^{\mathbf{X}}\mathbf{y} 
\qquad
\hat{\boldsymbol{\epsilon}} = \mathbf{y} - \hat{\mathbf{y}}
= (\mathbf{I}_n - \mathbf{P}^{\mathbf{X}})\mathbf{y}
= \mathbf{P}^{\mathbf{X}^\bot}\mathbf{y}
$$

$$
\mathbb{E}[\hat{\boldsymbol{\epsilon}}] = \mathbf{0} 
\qquad
\mathbb{Var}[\hat{\boldsymbol{\epsilon}}] = \sigma^2 \mathbf{P}^{\mathbf{X}^\bot}
$$

$$
\mathbb{E}[\hat{\mathbf{y}}]  = \mathbf{X}\boldsymbol{\beta} 
\qquad
\mathbb{Var}[\hat{\mathbf{y}}]  = \sigma^2 \mathbf{P}^{\mathbf{X}}
$$

$$
\mathbb{Cov}[\hat{\boldsymbol{\epsilon}}; \hat{\mathbf{y}}] = \mathbf{0}.
$$

$$
\hat{\sigma}^2 = \frac{1}{n - p} \| \hat{\epsilon} \|^2
\qquad
\mathbb{E}[\hat{\sigma}^2] = \sigma^2
$$

## Questions

* Confidence interval for $\hat{\boldsymbol{\beta}}, \hat{\sigma}^2$ ?

* Prediction intervals ?

* Assumptions on the moments are not enough.  
$\hookrightarrow$ We need assumptions on the *specific distribution* of the $\epsilon_i$.

* Most common assumption: $\epsilon_i$ are **Gaussian**.

<!-- ************************************************************************ -->
# Gaussian Model
<!-- ************************************************************************ -->

## Gaussian Model

### Model:
$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
$$

* $\mathbf{y}$ **random** vector of $n$ responses
* $\mathbf{X}$ **non random** $n\times p$ matrix of predictors
* $\boldsymbol{\epsilon}$ **random** vector of $n$ errors
* $\boldsymbol{\beta}$ **non random, unknown** vector of $p$ coefficients

### Assumptions:
  * (H1) $rg(\mathbf{X}) = p$
  * (H2) $\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0}, \sigma^2 \mathbf{I}_n)$ 

## Reminder - Notations

The column and row vectors of $\mathbf{X}$ are written as:

$$
\mathbf{X} = 
\begin{pmatrix} \mathbf{x}_{1} & \cdots & \mathbf{x}_{p} \end{pmatrix} 
=
\begin{pmatrix} \mathbf{x}^{1} \\ \vdots \\ \mathbf{x}^{n} \end{pmatrix} 
$$

Hence:
$$
y_i = \mathbf{x}^{i} \boldsymbol{\beta} + \epsilon_i \quad \forall~1\leq i \leq n
$$

<!-- ************************************************************************ -->
# Maximum Likelihood Estimators
<!-- ************************************************************************ -->

## Distribution of $\mathbf{y}$

* Model:
$$
y_i = \mathbf{x}^i \boldsymbol{\beta} + \epsilon_i
\quad 
\text{with}
\quad
\epsilon_i 
\underset{\text{iid}}{\operatorname{\sim}}
\mathcal{N}(0, \sigma^2) 
\quad 
\text{for}
\quad
1 \leq i \leq n
$$

* Distribution of $y_i$:
$$
y_i 
\underset{\text{iid}}{\operatorname{\sim}}
\mathcal{N}(\mathbf{x}^i \boldsymbol{\beta}, \sigma^2) 
\quad 
\text{for}
\quad
1 \leq i \leq n
$$

## Likelihood of $\mathbf{y}$ {.build}

* Distribution of $y_i$:
$$
y_i 
\underset{\text{iid}}{\operatorname{\sim}}
\mathcal{N}(\mathbf{x}^i \boldsymbol{\beta}, \sigma^2) 
\quad 
\text{for}
\quad
1 \leq i \leq n
$$

>* Likelihood of $\mathbf{y} = (y_1, \dotsc, y_n)^T$:
$$
\begin{aligned}
&L(\boldsymbol{\beta}, \sigma^2 | y_1, \dotsc, y_n)
= \prod_{i = 1}^n L(\boldsymbol{\beta}, \sigma^2 | y_i) & [ind.]\\
\end{aligned}
$$

and
$$
\begin{aligned}
L(\boldsymbol{\beta}, \sigma^2 | y_i)
= \cdots
\end{aligned}
$$

## Likelihood of $\mathbf{y}$

* Distribution of $y_i$:
$$
y_i 
\underset{\text{iid}}{\operatorname{\sim}}
\mathcal{N}(\mathbf{x}^i \boldsymbol{\beta}, \sigma^2) 
\quad 
\text{for}
\quad
1 \leq i \leq n
$$

* Likelihood of $\mathbf{y} = (y_1, \dotsc, y_n)^T$:
$$
\begin{aligned}
&L(\boldsymbol{\beta}, \sigma^2 | y_1, \dotsc, y_n)
= \prod_{i = 1}^n L(\boldsymbol{\beta}, \sigma^2 | y_i) & [ind.]\\
\end{aligned}
$$

and

$$
L(\boldsymbol{\beta}, \sigma^2 | y_i)
= \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{(y_i - \mathbf{x}^i \boldsymbol{\beta})^2}{2\sigma^2}\right)
$$

## Likelihood of $\mathbf{y}$

* Likelihood of $\mathbf{y} = (y_1, \dotsc, y_n)^T$:
$$
\begin{aligned}
&L(\boldsymbol{\beta}, \sigma^2 | y_1, \dotsc, y_n)
= \prod_{i = 1}^n L(\boldsymbol{\beta}, \sigma^2 | y_i)\\
&\qquad = \prod_{i = 1}^n \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{(y_i - \mathbf{x}^i \boldsymbol{\beta})^2}{2\sigma^2}\right)\\
&\qquad = \frac{1}{\left(\sqrt{2\pi\sigma^2}\right)^n}\exp\left(-\frac{1}{2\sigma^2}\sum_{i = 1}^n (y_i - \mathbf{x}^i \boldsymbol{\beta})^2\right)\\
\end{aligned}
$$

$$
\log L(\boldsymbol{\beta}, \sigma^2 | y_1, \dotsc, y_n) 
= - \frac{n}{2} \log(2\pi\sigma^2) - \frac{1}{2\sigma^2}\sum_{i = 1}^n (y_i - \mathbf{x}^i \boldsymbol{\beta})^2\\
$$

## ML Estimators

* The Maximum Likelihood estimators are:
$$
(\hat{\boldsymbol{\beta}}^{ML}, \hat{\sigma}^2_{ML}) 
= \underset{(\boldsymbol{\beta}, \sigma^2) \in \mathbb{R}^p\times\mathbb{R}_+^*}{\operatorname{argmax}} 
\log L(\boldsymbol{\beta}, \sigma^2 | \mathbf{y})
$$

with:
$$
\log L(\boldsymbol{\beta}, \sigma^2 | \mathbf{y}) 
= - \frac{n}{2} \log(2\pi\sigma^2) - \frac{1}{2\sigma^2}\sum_{i = 1}^n (y_i - \mathbf{x}^i \boldsymbol{\beta})^2
$$

## ML Estimators - $\boldsymbol{\beta}$

$$
\log L(\boldsymbol{\beta}, \sigma^2 | \mathbf{y}) 
= - \frac{n}{2} \log(2\pi\sigma^2) 
- \frac{1}{2\sigma^2}
\underbrace{\sum_{i = 1}^n (y_i - \mathbf{x}^i \boldsymbol{\beta})^2}_{\text{Sum of Squares } \|\mathbf{y} - \mathbf{X}\boldsymbol{\beta} \|^2}\\
$$

* For any $\sigma^2 > 0$,
$$
\hat{\boldsymbol{\beta}}^{ML}
= \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmax}} 
\log L(\boldsymbol{\beta}, \sigma^2 | \mathbf{y})
=  \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} \|\mathbf{y} - \mathbf{X}\boldsymbol{\beta} \|^2
= \hat{\boldsymbol{\beta}}^{OLS}
$$

$\hookrightarrow$ The ML estimators are the same as the OLS estimators.

## ML Estimators - $\sigma^2$

$$
\begin{aligned}
\hat{\sigma}^2_{ML}
&= \underset{\sigma^2 \in \mathbb{R}_+^*}{\operatorname{argmax}} 
\log L(\hat{\boldsymbol{\beta}}, \sigma^2 | \mathbf{y})\\
&=  \underset{\sigma^2 \in \mathbb{R}_+^*}{\operatorname{argmax}} 
- \frac{n}{2} \log(2\pi\sigma^2) - \frac{1}{2\sigma^2}\sum_{i = 1}^n (y_i - \mathbf{x}^i\hat{\boldsymbol{\beta}})^2\\
&=  \underset{\sigma^2 \in \mathbb{R}_+^*}{\operatorname{argmin}} 
\frac{n}{2} \log(2\pi\sigma^2) + \frac{1}{2\sigma^2}\sum_{i = 1}^n \hat{\epsilon}_i^2\\
\end{aligned}
$$
We get:

$$
\hat{\sigma}^2_{ML} 
= \frac{1}{n}\sum_{i = 1}^n \hat{\epsilon}_i^2
= \frac{\|\hat{\boldsymbol{\epsilon}} \|^2}{n}
= \frac{\|\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}} \|^2}{n}
$$

## ML Estimators - $\sigma^2$ - Remarks

* The ML estimator **is different** from the unbiased estimator of the variance we saw earlier.

$$
\hat{\sigma}^2_{ML} = \frac{\|\hat{\boldsymbol{\epsilon}} \|^2}{n}
\neq
\frac{\|\hat{\boldsymbol{\epsilon}} \|^2}{n-p}= \hat{\sigma}^2
$$

* The ML estimator of the variance is **biased**:
$$
\mathbb{E}[\hat{\sigma}^2_{ML}] = \frac{n - p}{n}\sigma^2 \neq \sigma^2
$$

<!-- ************************************************************************ -->
# Reminder - <br> Gaussian Vectors
<!-- ************************************************************************ -->

## Reminder: Gaussian Distribution 1/4

Let $X$ be a Gaussian r.v. with expectation $\mu$ and variance $\sigma^2$:
$$
X \sim \mathcal{N}(\mu, \sigma^2).
$$

It admits a probability density function (pdf):
$$
f(x) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{(x - \mu)^2}{2\sigma^2}\right)
\quad
\forall x \in \mathbb{R}.
$$

Moments:
$$
\mathbb{E}[X] = \mu
\quad
\mathbb{Var}[X] = \sigma^2
$$

## Reminder: Gaussian Vector 2/4

Let $\mathbf{X}$ be a Gaussian vector with expectation $\boldsymbol{\mu}$ and variance $\mathbf{\Sigma}$:
$$
\mathbf{X} \sim \mathcal{N}(\boldsymbol{\mu}, \mathbf{\Sigma}).
$$

It admits a probability density function (pdf):
$$
f(\mathbf{x}) 
= \frac{1}{\sqrt{(2\pi)^n|\mathbf{\Sigma}|}}
\exp\left(-\frac{1}{2} (\mathbf{x} - \boldsymbol{\mu})^T \mathbf{\Sigma}^{-1} (\mathbf{x} - \boldsymbol{\mu})\right)
\quad
\forall\ \mathbf{x} \in \mathbb{R}^n.
$$

Any linear combination of its coordinates is Gaussian.

Moments:
$$
\mathbb{E}[\mathbf{X}] = \boldsymbol{\mu}
\quad
\mathbb{Var}[\mathbf{X}] = \mathbb{E}[(\mathbf{X} - \boldsymbol{\mu})(\mathbf{X} - \boldsymbol{\mu})^T] = \mathbf{\Sigma}
$$

## Reminder: Gaussian Vector 3/4

**Property**:  
If $\mathbf{X} \sim \mathcal{N}(\boldsymbol{\mu}, \mathbf{\Sigma})$,
then for any $\mathbf{a}$ matrix and $\mathbf{b}$ vector,  
$$
\mathbf{Y} = \mathbf{a}\mathbf{X} + \mathbf{b} 
\sim \mathcal{N}(\mathbf{a}\boldsymbol{\mu} + \mathbf{b}, \mathbf{a}\mathbf{\Sigma}\mathbf{a}^T).
$$


**Property**:  
Let $\mathbf{X} \sim \mathcal{N}(\boldsymbol{\mu}, \mathbf{\Sigma})$, with $\mathbf{\Sigma}$ invertible.
Then $\mathbf{\Sigma}^{-1}$ is symmetric, positive definite and it is possible to find $\mathbf{\Sigma}^{-1/2}$ invertible
such that:
$$
[\mathbf{\Sigma}^{-1/2}]^T\mathbf{\Sigma}^{-1/2} = \mathbf{\Sigma}^{-1}
$$
Then:
$$
\mathbf{Y} = \mathbf{\Sigma}^{-1/2}(\mathbf{X} - \boldsymbol{\mu})
\sim \mathcal{N}(\mathbf{0}, \mathbf{I}).
$$

## Reminder: Gaussian Vector 4/4 {.build}

**Proof**:  
$$
\mathbf{Y} = \mathbf{\Sigma}^{-1/2}(\mathbf{X} - \boldsymbol{\mu})
$$

Take
$$
\mathbf{a} = \mathbf{\Sigma}^{-1/2}
\quad \text{and} \quad
\mathbf{b} = -\mathbf{\Sigma}^{-1/2}\boldsymbol{\mu}
$$

Then:
$$
\mathbf{a}\boldsymbol{\mu} + \mathbf{b} = \mathbf{\Sigma}^{-1/2}\boldsymbol{\mu} -\mathbf{\Sigma}^{-1/2}\boldsymbol{\mu} = \mathbf{0}
$$

and:
$$
\mathbf{a}\mathbf{\Sigma}\mathbf{a}^T 
= \mathbf{\Sigma}^{-1/2}\mathbf{\Sigma}[\mathbf{\Sigma}^{-1/2}]^T
= \mathbf{I}.
$$


<!-- ************************************************************************ -->
# Distribution of the Coefficients - $\sigma^2$ known
<!-- ************************************************************************ -->

## Distribution of $\hat{\boldsymbol{\beta}}$

* We already know the moments of the estimator:

$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}^{T}\mathbf{X})^{-1}\mathbf{X}^{T}\mathbf{y}
\qquad
\mathbb{E}[\hat{\boldsymbol{\beta}}] = \boldsymbol{\beta}
\qquad
\mathbb{Var}[\hat{\boldsymbol{\beta}}] = \sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}.
$$

* As $\mathbf{y}$ is a Gaussian vector, $\hat{\boldsymbol{\beta}}$ is hence also a Gaussian vector.

* Hence, when the variance $\sigma^2$ is known, we get:

$$
\hat{\boldsymbol{\beta}}
\sim
\mathcal{N}\left(
\boldsymbol{\beta};
\sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}
\right).
$$

<!-- ************************************************************************ -->
# Cochran Theorem
<!-- ************************************************************************ -->

## Chi Squared Distribution

**Definition**:  
Let $X_1, \dotsc, X_p$ be $p$ standard normal iid rv : $X_i \sim \mathcal{N}(0, 1)$.
$$
X = \sum_{i = 1}^p X_i^2 \sim \chi^2_p
$$ 

is a Chi squared r.v. with $p$ degrees of freedom.

**Property**:  
Let $\mathbf{X}$ be a Gaussian vector of size $p$ with expectation $\boldsymbol{\mu}$ and variance $\mathbf{\Sigma}$:
$\mathbf{X} \sim \mathcal{N}(\boldsymbol{\mu}, \mathbf{\Sigma})$.
Then, if $\mathbf{\Sigma}$ is invertible,
$$
(\mathbf{X} - \boldsymbol{\mu})^T \mathbf{\Sigma}^{-1} (\mathbf{X} - \boldsymbol{\mu})
\sim \chi^2_p
$$ 

is a Chi squared r.v. with $p$ degrees of freedom.

## Chi Squared Distribution - Proof {.build}

$\mathbf{X}$ is a Gaussian vector $\mathbf{X} \sim \mathcal{N}(\boldsymbol{\mu}, \mathbf{\Sigma})$.
with $\mathbf{\Sigma}$ is invertible, so:
$$
\mathbf{Y} = \mathbf{\Sigma}^{-1/2}(\mathbf{X} - \boldsymbol{\mu})
\sim \mathcal{N}(\mathbf{0}, \mathbf{I}_p).
$$

And:
$$
(\mathbf{X} - \boldsymbol{\mu})^T \mathbf{\Sigma}^{-1} (\mathbf{X} - \boldsymbol{\mu})
= \mathbf{Y}^T \mathbf{Y}
= \sum_{i = 1}^p Y_i^2
$$
with $Y_1, \dotsc, Y_p$ be $p$ standard normal iid rv.

Hence:
$$
(\mathbf{X} - \boldsymbol{\mu})^T \mathbf{\Sigma}^{-1} (\mathbf{X} - \boldsymbol{\mu})
= \sum_{i = 1}^p Y_i^2
\sim \chi^2_p.
$$


## Cochran Theorem

**Theorem**:
Let

* $\mathbf{Y}$ a Gaussian vector $\mathbf{Y} \sim \mathcal{N}(\boldsymbol{\mu}, \sigma^2\mathbf{I}_n)$;
* $\mathcal{M}$ a subspace of $\mathbb{R}^n$ of dimension $p$;
* $\mathbf{P}$ the orthogonal projection matrix on $\mathcal{M}$;
* $\mathbf{P}^{\bot} = \mathbf{I}_n - \mathbf{P}$ the orthogonal projection matrix on $\mathcal{M}^{\bot}$.

Then we have:

(i)
$\mathbf{P}\mathbf{Y} \sim \mathcal{N}(\mathbf{P}\boldsymbol{\mu}, \sigma^2\mathbf{P})$
and
$\mathbf{P}^{\bot}\mathbf{Y} \sim \mathcal{N}(\mathbf{P}^{\bot}\boldsymbol{\mu}, \sigma^2\mathbf{P}^{\bot})$;
(ii) $\mathbf{P}\mathbf{Y}$ and $\mathbf{P}^{\bot}\mathbf{Y}$ are **independent**;
(iii)
$\frac{1}{\sigma^2}\|\mathbf{P}(\mathbf{Y} - \boldsymbol{\mu}) \|^2 \sim \chi^2_{p}$
and
$\frac{1}{\sigma^2}\|\mathbf{P}^{\bot}(\mathbf{Y} - \boldsymbol{\mu}) \|^2 \sim \chi^2_{n - p}$

## Cochran Theorem - Proof - hints {.smaller}

**Theorem**:

* $\mathbf{Y}$ a Gaussian vector $\mathbf{Y} \sim \mathcal{N}(\boldsymbol{\mu}, \sigma^2\mathbf{I}_n)$; 
* $\mathbf{P}$ the orthogonal projection matrix on $\mathcal{M}$ dim $p$;
* $\mathbf{P}^{\bot} = \mathbf{I}_n - \mathbf{P}$ the orthogonal projection matrix on $\mathcal{M}^{\bot}$.
(i)
$\mathbf{P}\mathbf{Y} \sim \mathcal{N}(\mathbf{P}\boldsymbol{\mu}, \sigma^2\mathbf{P})$
and
$\mathbf{P}^{\bot}\mathbf{Y} \sim \mathcal{N}(\mathbf{P}^{\bot}\boldsymbol{\mu}, \sigma^2\mathbf{P}^{\bot})$;
(ii) $\mathbf{P}\mathbf{Y}$ and $\mathbf{P}^{\bot}\mathbf{Y}$ are **independent**;
(iii)
$\frac{1}{\sigma^2}\|\mathbf{P}(\mathbf{Y} - \boldsymbol{\mu}) \|^2 \sim \chi^2_{p}$
and
$\frac{1}{\sigma^2}\|\mathbf{P}^{\bot}(\mathbf{Y} - \boldsymbol{\mu}) \|^2 \sim \chi^2_{n - p}$

**Hints**:
$$
\begin{aligned}
&\text{1.}
&
\mathbf{X} \sim \mathcal{N}(\boldsymbol{\mu}, \mathbf{\Sigma})
\implies
\mathbf{a}\mathbf{X} + \mathbf{b} 
\sim \mathcal{N}(\mathbf{a}\boldsymbol{\mu} + \mathbf{b}, \mathbf{a}\mathbf{\Sigma}\mathbf{a}^T).\\
&\text{2.}
&
\mathbf{P} = \mathbf{U}\mathbf{\Delta}\mathbf{U}^T
\qquad
\mathbf{U}\mathbf{U}^T = \mathbf{I}_n
\qquad
\mathbf{\Delta} = 
\begin{pmatrix}
\mathbf{I}_p & \mathbf{0}\\
\mathbf{0} & \mathbf{0}
\end{pmatrix}.\\
&\text{2. (bis)}
&
\mathbf{Z}
= \mathbf{U}^T \mathbf{Y}
\sim \mathcal{N}(\dotsc, \dotsc)
\quad
\text{and}
\quad
\mathbf{\Delta}\mathbf{Z} 
= \begin{pmatrix}
\mathbf{Z}_p\\
\mathbf{0}_{n-p}
\end{pmatrix}
\end{aligned}
$$

## Cochran Theorem - Proof - (1) {.build}

From the linear property of Gaussian vectors:
$$
\mathbf{P}\mathbf{Y}
\sim \mathcal{N}(\mathbf{P}\boldsymbol{\mu}, \mathbf{P}[\sigma^2\mathbf{I}_n]\mathbf{P}^T)
$$

but:
$$
\begin{aligned}
\mathbf{P}[\sigma^2\mathbf{I}_n]\mathbf{P}^T
& = \sigma^2 \mathbf{P}\mathbf{P}^T \\
& = \sigma^2 \mathbf{P}\mathbf{P} & [\text{orthogonal}] \\
& = \sigma^2 \mathbf{P} & [\text{projection}] \\
\end{aligned}
$$

Same for $\mathbf{P}^{\bot}\mathbf{Y}$.

## Cochran Theorem - Proof - (2) - 1/2 {.build}

As $\mathbf{P}$ is an orthogonal projection matrix:
$$
\mathbf{P} = \mathbf{U}\mathbf{\Delta}\mathbf{U}^T
\qquad
\mathbf{U}\mathbf{U}^T = \mathbf{I}_n
\qquad
\mathbf{\Delta} = 
\begin{pmatrix}
\mathbf{I}_p & \mathbf{0}\\
\mathbf{0} & \mathbf{0}
\end{pmatrix}
$$

Define:
$$
\mathbf{Z}
= \mathbf{U}^T \mathbf{Y}
\sim \mathcal{N}(\mathbf{U}^T\boldsymbol{\mu}, \mathbf{U}^T[\sigma^2\mathbf{I}_n]\mathbf{U} = \sigma^2\mathbf{I}_n).
$$

Then:
$$
\mathbf{\Delta}\mathbf{Z} 
= \begin{pmatrix}
\mathbf{Z}_p\\
\mathbf{0}_{n-p}
\end{pmatrix}
\quad
\text{independent from}
\quad
(\mathbf{I}_n - \mathbf{\Delta})\mathbf{Z} 
= \begin{pmatrix}
\mathbf{0}_p\\
\mathbf{Z}_{n-p}
\end{pmatrix}
$$

## Cochran Theorem - Proof - (2) - 2/2

$$
\mathbf{P} = \mathbf{U}\mathbf{\Delta}\mathbf{U}^T
\qquad
\mathbf{U}\mathbf{U}^T = \mathbf{I}_n
\qquad
\mathbf{\Delta} = 
\begin{pmatrix}
\mathbf{I}_p & \mathbf{0}\\
\mathbf{0} & \mathbf{0}
\end{pmatrix}
$$

$$
\mathbf{\Delta}\mathbf{Z} 
= \begin{pmatrix}
\mathbf{Z}_p\\
\mathbf{0}_{n-p}
\end{pmatrix}
\quad
\text{independent from}
\quad
(\mathbf{I}_n - \mathbf{\Delta})\mathbf{Z} 
= \begin{pmatrix}
\mathbf{0}_p\\
\mathbf{Z}_{n-p}
\end{pmatrix}
$$
Lead to:
$$
\mathbf{U}\mathbf{\Delta}\mathbf{Z} 
= \mathbf{U}\mathbf{\Delta}\mathbf{U}^T\mathbf{Y} 
= \mathbf{P}\mathbf{Y}
$$
independent from
$$
\mathbf{U}(\mathbf{I}_n - \mathbf{\Delta})\mathbf{Z} 
= \mathbf{U}(\mathbf{I}_n - \mathbf{\Delta})\mathbf{U}^T\mathbf{Y} 
= \mathbf{P}^\bot\mathbf{Y}.
$$

## Cochran Theorem - Proof - (3) - 1/3 {.build}

$$
\begin{aligned}
\frac{1}{\sigma^2}\|\mathbf{P}(\mathbf{Y} - \boldsymbol{\mu}) \|^2
&= \frac{1}{\sigma^2}\|\mathbf{U}\mathbf{\Delta}\mathbf{U}^T(\mathbf{Y} - \boldsymbol{\mu}) \|^2
= \cdots \qquad \qquad
\end{aligned}
$$

$$
\begin{aligned}
\frac{1}{\sigma^2}\|\mathbf{P}(\mathbf{Y} - \boldsymbol{\mu}) \|^2
&= \frac{1}{\sigma^2}[\mathbf{U}\mathbf{\Delta}\mathbf{U}^T(\mathbf{Y} - \boldsymbol{\mu})]^T[\mathbf{U}\mathbf{\Delta}\mathbf{U}^T(\mathbf{Y} - \boldsymbol{\mu})] \\
&= \frac{1}{\sigma^2}[\mathbf{\Delta}\mathbf{U}^T(\mathbf{Y} - \boldsymbol{\mu})]^T\mathbf{U}^T\mathbf{U}[\mathbf{\Delta}\mathbf{U}^T(\mathbf{Y} - \boldsymbol{\mu})] \\
&= \frac{1}{\sigma^2}(\mathbf{\Delta}\mathbf{U}^T\mathbf{Y} - \mathbf{\Delta}\mathbf{U}^T\boldsymbol{\mu})^T(\mathbf{\Delta}\mathbf{U}^T\mathbf{Y} - \mathbf{\Delta}\mathbf{U}^T\boldsymbol{\mu}) \\
\end{aligned}
$$

## Cochran Theorem - Proof - (3) - 2/3 {.build}

$$
\begin{aligned}
\frac{1}{\sigma^2}\|\mathbf{P}(\mathbf{Y} - \boldsymbol{\mu}) \|^2
&= \frac{1}{\sigma^2}(\mathbf{\Delta}\mathbf{U}^T\mathbf{Y} - \mathbf{\Delta}\mathbf{U}^T\boldsymbol{\mu})^T(\mathbf{\Delta}\mathbf{U}^T\mathbf{Y} - \mathbf{\Delta}\mathbf{U}^T\boldsymbol{\mu}) \\
\end{aligned}
$$

But:
$$
\mathbf{\Delta}\mathbf{U}^T \mathbf{Y}
= \mathbf{\Delta}\mathbf{Z} 
= \begin{pmatrix}
\mathbf{Z}_p\\
\mathbf{0}_{n-p}
\end{pmatrix}
\quad
\text{and}
\quad
\mathbf{\Delta}\mathbf{U}^T\boldsymbol{\mu}
= \begin{pmatrix}
(\mathbf{U}^T\boldsymbol{\mu})_p\\
\mathbf{0}_{n-p}
\end{pmatrix}
$$

Hence:
$$
\begin{aligned}
\frac{1}{\sigma^2}\|\mathbf{P}(\mathbf{Y} - \boldsymbol{\mu}) \|^2
&= \frac{1}{\sigma^2}(\mathbf{Z}_p - (\mathbf{U}^T\boldsymbol{\mu})_p)^T(\mathbf{Z}_p - (\mathbf{U}^T\boldsymbol{\mu})_p) \\
\end{aligned}
$$

## Cochran Theorem - Proof - (3) - 3/3 {.build}
$$
\begin{aligned}
\frac{1}{\sigma^2}\|\mathbf{P}(\mathbf{Y} - \boldsymbol{\mu}) \|^2
&= \frac{1}{\sigma^2}(\mathbf{Z}_p - (\mathbf{U}^T\boldsymbol{\mu})_p)^T(\mathbf{Z}_p - (\mathbf{U}^T\boldsymbol{\mu})_p) \\
&= (\mathbf{Z}_p - (\mathbf{U}^T\boldsymbol{\mu})_p)^T[\sigma^2\mathbf{I}_p]^{-1}(\mathbf{Z}_p - (\mathbf{U}^T\boldsymbol{\mu})_p) \\
\end{aligned}
$$

But:
$$
\mathbf{Z}
\sim \mathcal{N}(\mathbf{U}^T\boldsymbol{\mu}, \sigma^2\mathbf{I}_n)
\quad
\text{hence}
\quad
\mathbf{Z}_p
\sim \mathcal{N}((\mathbf{U}^T\boldsymbol{\mu})_p, \sigma^2\mathbf{I}_p)
$$

thanks to previous lemma:
$$
\frac{1}{\sigma^2}\|\mathbf{P}(\mathbf{Y} - \boldsymbol{\mu}) \|^2 \sim \chi^2_{p}
$$

Same for $\frac{1}{\sigma^2}\|\mathbf{P}^\bot(\mathbf{Y} - \boldsymbol{\mu}) \|^2 \sim \chi^2_{n-p}$

<!-- ************************************************************************ -->
# Distribution of $\hat{\sigma}^2$
<!-- ************************************************************************ -->

## Distribution of $\hat{\sigma}^2$

When the variance $\sigma^2$ is known, we get:

$$
\frac{(n-p) \hat{\sigma}^2}{\sigma^2}
\sim
\chi^2_{n-p}
$$

and

$$
\hat{\boldsymbol{\beta}} \text{ and } \hat{\sigma}^2 \text{ are independent.}
$$

## Proof: Use Cochran Theorem

We have:

* $\mathbf{Y}$ a Gaussian vector $\mathbf{Y} \sim \mathcal{N}(\mathbf{X}\boldsymbol{\beta}, \sigma^2\mathbf{I}_n)$;
* $\mathcal{M}(\mathbf{X}) = \text{span}\{\mathbf{x}_1, \dotsc, \mathbf{x}_p\}$;
* $\mathbf{P}^{\mathbf{X}} = \mathbf{X}(\mathbf{X}^{T}\mathbf{X})^{-1}\mathbf{X}^{T}$ orthogonal projection on $\mathcal{M}(\mathbf{X})$;
* $\mathbf{P}^{\mathbf{X}^\bot} = \mathbf{I}_n - \mathbf{P}^{\mathbf{X}}$

And:

* $\hat{\boldsymbol{\beta}} = (\mathbf{X}^{T}\mathbf{X})^{-1}\mathbf{X}^{T}\mathbf{y}$
* $\hat{\sigma}^2 = \frac{\|\hat{\boldsymbol{\epsilon}} \|^2}{n-p} = \frac{\|\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}} \|^2}{n-p}$

## Proof - Chi squared {.build}

$$
\begin{aligned}
\hat{\sigma}^2 
& = \frac{\|\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}} \|^2}{n-p}
= \frac{\|\mathbf{y} - \hat{\mathbf{y}} \|^2}{n-p}\\
& = \frac{\|\mathbf{y} - \mathbf{P}^{\mathbf{X}}\mathbf{y} \|^2}{n-p}
= \frac{\|\mathbf{P}^{\mathbf{X}^\bot}\mathbf{y} \|^2}{n-p}\\
\end{aligned}
$$

Hence:
$$
\begin{aligned}
\frac{n-p}{\sigma^2}\hat{\sigma}^2 
& = \frac{1}{\sigma^2}\|\mathbf{P}^{\mathbf{X}^\bot}\mathbf{y} \|^2
\end{aligned}
$$

And, from Cochran's Theorem:
$$
\frac{n-p}{\sigma^2}\hat{\sigma}^2 
= \frac{1}{\sigma^2}\|\mathbf{P}^{\mathbf{X}^\bot}\mathbf{y} \|^2
\sim \chi^2_{n-p}.
$$

## Proof - Independence

$$
\begin{aligned}
\hat{\boldsymbol{\beta}} 
&= (\mathbf{X}^{T}\mathbf{X})^{-1}\mathbf{X}^{T}\mathbf{y}\\
&= (\mathbf{X}^{T}\mathbf{X})^{-1}(\mathbf{X}^{T}\mathbf{X})(\mathbf{X}^{T}\mathbf{X})^{-1}\mathbf{X}^{T}\mathbf{y}\\
&= (\mathbf{X}^{T}\mathbf{X})^{-1}\mathbf{X}^{T}(\mathbf{X}(\mathbf{X}^{T}\mathbf{X})^{-1}\mathbf{X}^{T})\mathbf{y}\\
&= (\mathbf{X}^{T}\mathbf{X})^{-1}\mathbf{X}^{T}\mathbf{P}^{\mathbf{X}}\mathbf{y}\\
\end{aligned}
$$

and
$$
\hat{\sigma}^2 
= \frac{1}{n-p}\|\mathbf{P}^{\mathbf{X}^\bot}\mathbf{y} \|^2
$$

But $\mathbf{P}^{\mathbf{X}}\mathbf{y}$ and $\mathbf{P}^{\mathbf{X}^\bot}\mathbf{y}$ are independent
from Cochran's theorem.

Hence, $\hat{\boldsymbol{\beta}}$ and $\hat{\sigma}^2$ are independent.

<!-- ************************************************************************ -->
# Distribution of the Coefficients - $\sigma^2$ unknown
<!-- ************************************************************************ -->

## Distribution of $\hat{\boldsymbol{\beta}}$

When $\sigma^2$ is known:

$$
\hat{\boldsymbol{\beta}}
\sim
\mathcal{N}\left(
\boldsymbol{\beta};
\sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}
\right).
$$

i.e. for $1 \leq k \leq p$:

$$
\hat{\beta}_k
\sim
\mathcal{N}(\beta_k, \sigma^2 [(\mathbf{X}^T\mathbf{X})^{-1}]_{kk})
$$

i.e.

$$
\frac{\hat{\beta}_k - \beta_k}{\sqrt{\sigma^2 [(\mathbf{X}^T\mathbf{X})^{-1}]_{kk}}}
\sim
\mathcal{N}(0, 1).
$$

>* **Problem** $\sigma^2$ is generally *unknown*.

>* **Solution** Replace $\sigma^2$ by $\hat{\sigma}^2$.

## Distribution of $\hat{\boldsymbol{\beta}}$

When $\sigma^2$ is **unknown**, for any $1 \leq k \leq p$:
$$
\frac{\hat{\beta}_k - \beta_k}{\sqrt{\hat{\sigma}^2 [(\mathbf{X}^T\mathbf{X})^{-1}]_{kk}}}
\sim
\mathcal{T}_{n-p}.
$$
Notation:
$$
\hat{\sigma}^2_k = \hat{\sigma}^2_{\hat{\beta}_k} = \hat{\sigma}^2 [(\mathbf{X}^T\mathbf{X})^{-1}]_{kk}
$$

**Attention**:  
$$
[(\mathbf{X}^T\mathbf{X})^{-1}]_{kk}
\neq
[(\mathbf{X}^T\mathbf{X})_{kk}]^{-1}
$$

## Reminder: Student Distribution

* $Z \sim \mathcal{N}(0, 1)$
* $X \sim \chi^2_p$
* $Z$ and $X$ independent

Then 
$$
T = \frac{Z}{\sqrt{X/p}} \sim \mathcal{T}_p
$$ 


## Distribution of $\hat{\boldsymbol{\beta}}$ - Proof {.build}

$$
\frac{\hat{\beta}_k - \beta_k}{\sqrt{\sigma^2 [(\mathbf{X}^T\mathbf{X})^{-1}]_{kk}}}
\sim
\mathcal{N}(0, 1)
\quad
\text{and}
\quad
\frac{(n-p) \hat{\sigma}^2}{\sigma^2}
\sim
\chi^2_{n-p}
$$

$$
\frac{\hat{\beta}_k - \beta_k}{\sqrt{\sigma^2 [(\mathbf{X}^T\mathbf{X})^{-1}]_{kk}}}
\quad
\text{and}
\quad
\frac{(n-p) \hat{\sigma}^2}{\sigma^2}
\quad
\textbf{independent}
$$

Hence:
$$
\frac{\frac{\hat{\beta}_k - \beta_k}{\sqrt{\sigma^2 [(\mathbf{X}^T\mathbf{X})^{-1}]_{kk}}}}{
\sqrt{\frac{(n-p) \hat{\sigma}^2}{\sigma^2}/ (n-p)}}
=
\frac{\hat{\beta}_k - \beta_k}{\sqrt{\hat{\sigma}^2 [(\mathbf{X}^T\mathbf{X})^{-1}]_{kk}}}
\sim
\mathcal{T}_{n-p}
$$

<!-- ************************************************************************ -->
# Confidence Intervals and Tests
<!-- ************************************************************************ -->

## Confidence intervals {.build}

$$
\frac{\hat{\beta}_k - \beta_k}{\sqrt{\hat{\sigma}^2_k}}
\sim
\mathcal{T}_{n-p}
\quad
\text{and}
\quad
\frac{n-p}{\sigma^2}\hat{\sigma}^2 \sim \chi^2_{n-p}
$$

With probability $1-\alpha$:
$$
\begin{aligned}
\beta_k &\in \left[\hat{\beta}_k \pm t_{n-p}(1 - \alpha/2) \sqrt{\hat{\sigma}_k^2}\right]\\
\sigma^2 &\in \left[\frac{(n-p)\hat{\sigma}^2}{c_{n-p}(1 - \alpha/2)}; \frac{(n-p)\hat{\sigma}^2}{c_{n-p}(\alpha/2)}\right]
\end{aligned}
$$

## Tests {.build}

$$
\text{Hypothesis:}
\qquad\qquad
\mathcal{H}_0: \beta_k = 0
\quad
\text{vs}
\quad
\mathcal{H}_1: \beta_k \neq 0
$$

$$
\text{Test Statistic:}
\qquad\qquad\qquad\qquad
T_k 
= \frac{\hat{\beta}_k}{\sqrt{\hat{\sigma}^2_k}}
\underset{\mathcal{H}_0}{\sim}
\mathcal{T}_{n-p}
$$

$$
\text{Reject Region:}
\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad\\
\mathbb{P}[\text{Reject} | \mathcal{H}_0 \text{ true}] \leq \alpha
\qquad\qquad\qquad\qquad\qquad\\
\qquad\qquad\qquad \iff \mathcal{R}_\alpha = \left\{ t \in \mathbb{R}  ~|~ |t| \geq t_{n-p}(1 - \alpha/2) \right\}
$$

$$
\text{p value:}\qquad\qquad\qquad\qquad\quad
p = \mathbb{P}_{\mathcal{H}_0}[|T_k| > T_k^{obs}]
$$

## Test - Null Distribution

```{r testplot, echo=FALSE, fig.height=5, fig.width=7, fig.align='center'}
n <- 10
p <- 5
ccc <- hcl.colors(5)
### Student Distribution
x <- -5500:5500 / 1000
# 1
p <- n - p
plot(x, dt(x, p), type = "l", ylab = "density", ylim = c(0, 0.4), col = ccc[1], lwd = 2, xlab = "")
abline(h = 0)
# quantiles for 1 - alp = 95% interval
alp <- 0.1
quants <- qt(c(alp / 2, 1 - alp / 2), p)
segments(x0 = quants, y0 = -1, x1 = quants, y1 = dt(0, p), lty = 2, col = ccc[3])
axis(side = 1, at = quants, labels = c(expression(t[n-p]^(alpha/2)), expression(t[n-p]^(1 - alpha/2))), col.ticks = ccc[3], col.axis = ccc[3], las = 1, padj = 1.2)
# Color area under the curve
x_inf <- -5500:(quants[1]*1000)/1000
polygon(c(x_inf, rev(x_inf)),
        c(dt(x_inf, p), rep(0, length(x_inf))),
        col = ccc[5], border = NA, density = 20)
x_sup <- (quants[2]*1000):5500/1000
polygon(c(x_sup, rev(x_sup)),
        c(dt(x_sup, p), rep(0, length(x_sup))),
        col = ccc[5], border = NA, density = 20)
# # beta hat
# t_obs <- 3
# segments(t_obs, 5, t_obs, -1, lty = 2, col = ccc[4])
# axis(side = 1, at = t_obs, labels = expression(T[k]^obs), col.ticks = ccc[4], col.axis = ccc[4], las = 1, padj = 1.2)
# legend
legend("topright",
       c(expression(T[n-p])),
       col = ccc[1],
       lwd = 2, lty = 1)
```

## Test - Do not reject $\mathcal{H}_0$

```{r testplot2, echo=FALSE, fig.height=5, fig.width=7, fig.align='center'}
n <- 10
p <- 5
ccc <- hcl.colors(5)
### Student Distribution
x <- -5500:5500 / 1000
# 1
p <- n - p
plot(x, dt(x, p), type = "l", ylab = "density", ylim = c(0, 0.4), col = ccc[1], lwd = 2, xlab = "")
abline(h = 0)
# quantiles for 1 - alp = 95% interval
alp <- 0.1
quants <- qt(c(alp / 2, 1 - alp / 2), p)
segments(x0 = quants, y0 = -1, x1 = quants, y1 = dt(0, p), lty = 2, col = ccc[3])
axis(side = 1, at = quants, labels = c(expression(t[n-p]^(alpha/2)), expression(t[n-p]^(1 - alpha/2))), col.ticks = ccc[3], col.axis = ccc[3], las = 1, padj = 1.2)
# Color area under the curve
x_inf <- -5500:(quants[1]*1000)/1000
polygon(c(x_inf, rev(x_inf)),
        c(dt(x_inf, p), rep(0, length(x_inf))),
        col = ccc[5], border = NA, density = 20)
x_sup <- (quants[2]*1000):5500/1000
polygon(c(x_sup, rev(x_sup)),
        c(dt(x_sup, p), rep(0, length(x_sup))),
        col = ccc[5], border = NA, density = 20)
# beta hat
t_obs <- 1
segments(t_obs, 5, t_obs, -1, lty = 2, col = ccc[4])
axis(side = 1, at = t_obs, labels = expression(T[k]^obs), col.ticks = ccc[4], col.axis = ccc[4], las = 1, padj = 1.2)
# legend
legend("topright",
       c(expression(T[n-p])),
       col = ccc[1],
       lwd = 2, lty = 1)
```

## Test - Reject $\mathcal{H}_0$

```{r testplot3, echo=FALSE, fig.height=5, fig.width=7, fig.align='center'}
n <- 10
p <- 5
ccc <- hcl.colors(5)
### Student Distribution
x <- -5500:5500 / 1000
# 1
p <- n - p
plot(x, dt(x, p), type = "l", ylab = "density", ylim = c(0, 0.4), col = ccc[1], lwd = 2, xlab = "")
abline(h = 0)
# quantiles for 1 - alp = 95% interval
alp <- 0.1
quants <- qt(c(alp / 2, 1 - alp / 2), p)
segments(x0 = quants, y0 = -1, x1 = quants, y1 = dt(0, p), lty = 2, col = ccc[3])
axis(side = 1, at = quants, labels = c(expression(t[n-p]^(alpha/2)), expression(t[n-p]^(1 - alpha/2))), col.ticks = ccc[3], col.axis = ccc[3], las = 1, padj = 1.2)
# Color area under the curve
x_inf <- -5500:(quants[1]*1000)/1000
polygon(c(x_inf, rev(x_inf)),
        c(dt(x_inf, p), rep(0, length(x_inf))),
        col = ccc[5], border = NA, density = 20)
x_sup <- (quants[2]*1000):5500/1000
polygon(c(x_sup, rev(x_sup)),
        c(dt(x_sup, p), rep(0, length(x_sup))),
        col = ccc[5], border = NA, density = 20)
# beta hat
t_obs <- 3
segments(t_obs, 5, t_obs, -1, lty = 2, col = ccc[4])
axis(side = 1, at = t_obs, labels = expression(T[k]^obs), col.ticks = ccc[4], col.axis = ccc[4], las = 1, padj = 1.2)
# legend
legend("topright",
       c(expression(T[n-p])),
       col = ccc[1],
       lwd = 2, lty = 1)
```

## Test - p value

```{r testplot4, echo=FALSE, fig.height=5, fig.width=7, fig.align='center'}
n <- 10
p <- 5
ccc <- hcl.colors(5)
### Student Distribution
x <- -5500:5500 / 1000
# 1
p <- n - p
plot(x, dt(x, p), type = "l", ylab = "density", ylim = c(0, 0.4), col = ccc[1], lwd = 2, xlab = "")
abline(h = 0)
# quantiles for 1 - alp = 95% interval
alp <- 0.1
quants <- qt(c(alp / 2, 1 - alp / 2), p)
segments(x0 = quants, y0 = -1, x1 = quants, y1 = dt(0, p), lty = 2, col = ccc[3])
axis(side = 1, at = quants, labels = c(expression(t[n-p]^(alpha/2)), expression(t[n-p]^(1 - alpha/2))), col.ticks = ccc[3], col.axis = ccc[3], las = 1, padj = 1.2)
# beta hat
t_obs <- 3
segments(t_obs, 5, t_obs, -1, lty = 2, col = ccc[4])
axis(side = 1, at = t_obs, labels = expression(T[k]^obs), col.ticks = ccc[4], col.axis = ccc[4], las = 1, padj = 1.2)
segments(-t_obs, 5, -t_obs, -1, lty = 2, col = ccc[4])
axis(side = 1, at = -t_obs, labels = expression(-T[k]^obs), col.ticks = ccc[4], col.axis = ccc[4], las = 1, padj = 1.2)
# Color area under the curve
x_inf <- -5500:(-t_obs*1000)/1000
polygon(c(x_inf, rev(x_inf)),
        c(dt(x_inf, p), rep(0, length(x_inf))),
        col = ccc[4], border = NA, density = 20)
x_sup <- (t_obs*1000):5500/1000
polygon(c(x_sup, rev(x_sup)),
        c(dt(x_sup, p), rep(0, length(x_sup))),
        col = ccc[4], border = NA, density = 20)
# legend
legend("topright",
       c(expression(T[n-p])),
       col = ccc[1],
       lwd = 2, lty = 1)
```

<!-- ************************************************************************ -->
# Simulated Dataset
<!-- ************************************************************************ -->

## Simulated Dataset

$$
y_i = -1 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

```{r rsq1, echo=TRUE}
set.seed(12890926)

## Predictors
n <- 100
x_1 <- runif(n, min = -2, max = 2)
x_2 <- runif(n, min = 0, max = 4)

## Noise
eps <- rnorm(n, mean = 0, sd = 1)

## Model sim
beta_0 <- -1; beta_1 <- 3; beta_2 <- -1
y_sim <- beta_0 + beta_1 * x_1 + beta_2 * x_2 + eps
```

## Simulated Dataset - Fit {.smaller}

$$
y_i = -1 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

```{r rsq2, echo=TRUE}
fit <- lm(y_sim ~ x_1 + x_2)
summary(fit)
```

## Simulated Dataset - Wrong Fit{.smaller}

```{r rsq3, echo=TRUE}
## unrelated noise variable x_3
x_3 <- runif(n, min = -4, max = 0)
## Fit
fit <- lm(y_sim ~ x_1 + x_2 + x_3); summary(fit)
```

<!-- ************************************************************************ -->
# Prediction
<!-- ************************************************************************ -->

## Predict a new point

>* We fitted $\hat{\beta}_1, \dots, \hat{\beta}_p$ on $((y_1, \mathbf{x}^1), \dotsc, (y_n, \mathbf{x}^n))$

>* A new line of predictors $\mathbf{x}^{n+1}$ comes along. How can we guess $y_{n+1}$ ?

>* We use the same model:
$$
y_{n+1} = \mathbf{x}^{n+1}\boldsymbol{\beta} + \epsilon_{n+1}
$$
with
$\epsilon_{n+1} \sim \mathcal{N}(0, \sigma^2)$, independent from all $(\epsilon_{i})_{1\leq i \leq n}$.

>* We predict $y_{n+1}$ with:
$$
\hat{y}_{n+1} = \mathbf{x}^{n+1}\hat{\boldsymbol{\beta}}
$$

>* **Question**: What is the error 
$\hat{\epsilon}_{n+1} = y_{n+1} - \hat{y}_{n+1}$ ?

## Prediction Error - Known variance

The prediction error $\hat{\epsilon}_{n+1} = y_{n+1} - \hat{y}_{n+1}$
is such that:
$$
\hat{\epsilon}_{n+1}
\sim \mathcal{N}\left(
0,
\sigma^2 \left(1 + \mathbf{x}^{n+1} (\mathbf{X}^T\mathbf{X})^{-1}(\mathbf{x}^{n+1})^T\right)
\right)\\
$$
**Proof**:  

* $\hat{\epsilon}_{n+1}$ is Gaussian as the sum of two Gaussian variables.  
* We already know its mean and variance (see previous lesson).

**Remark**: $\mathbf{x}^{n+1} (\mathbf{X}^T\mathbf{X})^{-1}(\mathbf{x}^{n+1})^T$ is a *scalar*.

## Prediction Error - Unknown variance

We get:
$$
\frac{y_{n+1} - \hat{y}_{n+1}}{\sqrt{\hat{\sigma}^2 \left(1 + \mathbf{x}^{n+1} (\mathbf{X}^T\mathbf{X})^{-1}(\mathbf{x}^{n+1})^T\right)}}
\sim
\mathcal{T}_{n-p}
$$

**Proof**:

$$
\frac{y_{n+1} - \hat{y}_{n+1}}{\sqrt{\hat{\sigma}^2 \left(1 + \mathbf{x}^{n+1} (\mathbf{X}^T\mathbf{X})^{-1}(\mathbf{x}^{n+1})^T\right)}}
=
\frac{\frac{y_{n+1} - \hat{y}_{n+1}}{\sqrt{\sigma^2 \left(1 + \mathbf{x}^{n+1} (\mathbf{X}^T\mathbf{X})^{-1}(\mathbf{x}^{n+1})^T\right)}}}{
\sqrt{\frac{(n-p) \hat{\sigma}^2}{\sigma^2}/ (n-p)}}
$$

## Prediction - Confidence Interval

With probability $1-\alpha$:
$$
y_{n+1} \in \left[\hat{y}_{n+1} \pm t_{n-p}(1 - \alpha/2) \sqrt{\hat{\sigma}^2 \left(1 + \mathbf{x}^{n+1} (\mathbf{X}^T\mathbf{X})^{-1}(\mathbf{x}^{n+1})^T\right)}\right]\\
$$


<!-- ************************************************************************ -->
# Advertising
<!-- ************************************************************************ -->

## Advertising {.smaller}

```{r ad, echo = FALSE, message = FALSE}
library(here)
ad <- read.csv(here("data", "Advertising.csv"), row.names = "X")
```

```{r ad2}
fit_all <- lm(sales ~ TV + radio + newspaper, data = ad)
summary(fit_all)
```
