---
title: 'Régression Multiple - Sélection de Variable'
author: "Paul Bastide - Benjamin Charlier"
date: "09/03/2022"
output:
  ioslides_presentation:
    fig_width: 7
    fig_height: 4
  self_contained: true
editor_options: 
  chunk_output_type: console
---
<!-- ************************************************************************ -->
# What we know
<!-- ************************************************************************ -->

## Gaussian Model

### Model:
$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
$$

* $\mathbf{y}$ **random** vector of $n$ responses
* $\mathbf{X}$ **non random** $n\times p$ matrix of predictors
* $\boldsymbol{\epsilon}$ **random** vector of $n$ errors
* $\boldsymbol{\beta}$ **non random, unknown** vector of $p$ coefficients

### Assumptions:
* (H1) $rg(\mathbf{X}) = p$
* (H2) $\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0}, \sigma^2 \mathbf{I}_n)$ 


## Distribution of the estimators

Distribution of $\hat{\boldsymbol{\beta}}$:
$$
\hat{\boldsymbol{\beta}}
\sim
\mathcal{N}\left(
\boldsymbol{\beta};
\sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}
\right)
\quad
\text{and}
\quad
\frac{\hat{\beta}_k - \beta_k}{\sqrt{\hat{\sigma}^2 [(\mathbf{X}^T\mathbf{X})^{-1}]_{kk}}}
\sim
\mathcal{T}_{n-p}.
$$

Distribution of $\hat{\sigma}^2$:
$$
\frac{(n-p) \hat{\sigma}^2}{\sigma^2}
\sim
\chi^2_{n-p}.
$$

## Student t tests

$$
\text{Hypothesis:}
\qquad\qquad
\mathcal{H}_0: \beta_k = 0
\quad
\text{vs}
\quad
\mathcal{H}_1: \beta_k \neq 0
$$

$$
\text{Test Statistic:}
\qquad\qquad\qquad\qquad
T_k 
= \frac{\hat{\beta}_k}{\sqrt{\hat{\sigma}^2_k}}
\underset{\mathcal{H}_0}{\sim}
\mathcal{T}_{n-p}
$$

$$
\text{Reject Region:}
\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad\\
\mathbb{P}[\text{Reject} | \mathcal{H}_0 \text{ true}] \leq \alpha
\qquad\qquad\qquad\qquad\qquad\\
\qquad\qquad\qquad \iff \mathcal{R}_\alpha = \left\{ t \in \mathbb{R}  ~|~ |t| \geq t_{n-p}(1 - \alpha/2) \right\}
$$

$$
\text{p value:}\qquad\qquad\qquad\qquad\quad
p = \mathbb{P}_{\mathcal{H}_0}[|T_k| > T_k^{obs}]
$$

<!-- ************************************************************************ -->
# Problems with the $t$-test
<!-- ************************************************************************ -->

## Simulated Dataset

$$
y_i = -1 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

```{r test1, echo=TRUE}
set.seed(1289)

## Predictors
n <- 500
x_1 <- runif(n, min = -2, max = 2)
x_2 <- runif(n, min = -2, max = 2)

## Noise
eps <- rnorm(n, mean = 0, sd = 2)

## Model sim
beta_0 <- -1; beta_1 <- 3; beta_2 <- -1
y_sim <- beta_0 + beta_1 * x_1 + beta_2 * x_2 + eps

## Useless predictor
x_junk_1 <- runif(n, min = -2, max = 2)
```

## Simulated Dataset - Fit {.smaller}

$$
y_i = -1 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

```{r test2, echo=TRUE}
fit <- lm(y_sim ~ x_1 + x_2 + x_junk_1)
summary(fit)
```

## Simulated Dataset

$$
y_i = -1 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

```{r test23, echo=TRUE}
## Other Useless Predictors
p_junk <- 100
x_junk <- matrix(runif(n * p_junk, min = -2, max = 2),
                 ncol = p_junk)

## Data frame
data_junk <- data.frame(y_sim = y_sim,
                        x_junk = x_junk)
```

## Simulated Dataset - Fit - Junk {.smaller}

```{r test4, echo=TRUE}
fit <- lm(y_sim ~ -1 + ., data = data_junk)
summary(fit)
```

## Simulated Dataset - Fit - Junk {.smaller}

```{r test5, echo=TRUE}
p_values <- summary(fit)$coefficients[, 4]
hist(p_values, col = "lightblue", breaks = 20)
```

## Simulated Dataset - Fit - Junk {.smaller}

```{r test6, echo=TRUE}
sum(p_values <= 0.05)
summary(fit)$coefficients[p_values <= 0.05, ]
```


$~$

Reject Region: $\mathbb{P}[\text{Reject} | \mathcal{H}_0 \text{ true}] \leq \alpha$

**Just by chance, about 5% of false discoveries.**

<!-- ************************************************************************ -->
# Joint Distribution of the Coefficients - $\sigma^2$ unknown
<!-- ************************************************************************ -->

## Joint distribution of $\hat{\boldsymbol{\beta}}$

When $\sigma^2$ is known:

$$
\hat{\boldsymbol{\beta}}
\sim
\mathcal{N}\left(
\boldsymbol{\beta};
\sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}
\right).
$$

When $\sigma^2$ is **unknown**:
$$
\frac{1}{p\hat{\sigma}^2}\left(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}\right)^T (\mathbf{X}^T\mathbf{X})\left(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}\right)
\sim
\mathcal{F}^p_{n-p}
$$

with $\mathcal{F}^p_{n-p}$ the **Fisher** distribution.

## Reminder: Fisher Distribution

Let $U_1 \sim \chi^2_{p_1}$ and $U_2 \sim \chi^2_{p_2}$,
$U_1$ and $U_2$ independent.
Then 
$$
F = \frac{U_1/p_1}{U_2/p_2} \sim \mathcal{F}^{p_1}_{p_2}
$$ 

```{r fisher, echo=FALSE, fig.height=4, fig.width=7, fig.align='center'}
x <- 0:600 / 100
ccc <- hcl.colors(6)
par(mar = c(5, 4, 1, 2) + 0.1)
# 1
p1 <- 2
p2 <- 3
plot(x, df(x, p1, p2), type = "l", ylab = "f(x)", ylim = c(0, 1), col = ccc[1], lwd = 2)
# p2
p2 <- 10
lines(x, df(x, p1, p2), col = ccc[5], lwd = 2)
# p1
p1 <- 10
lines(x, df(x, p1, p2), col = ccc[2], lwd = 2)
# p1
p1 <- 20
lines(x, df(x, p1, p2), col = ccc[4], lwd = 2)
# legend
legend("topright",
       c(expression(paste(p[1] == 2, "   ; ", p[2] == 3)),
         expression(paste(p[1] == 2, "   ; ", p[2] == 10)),
         expression(paste(p[1] == 10, "   ; ", p[2] == 10)),
         expression(paste(p[1] == 20, "   ; ", p[2] == 10))),
       col = ccc[c(1, 5, 2, 4)],
       lwd = 2, lty = 1)
```

## Joint distribution of $\hat{\boldsymbol{\beta}}$ - Proof - Hints

Let's show:
$$
\frac{1}{p\hat{\sigma}^2}\left(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}\right)^T (\mathbf{X}^T\mathbf{X})\left(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}\right)
\sim
\mathcal{F}^p_{n-p}
$$

**Hints:**  
$$
\hat{\boldsymbol{\beta}}
\sim
\mathcal{N}\left(
\boldsymbol{\beta};
\sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}
\right)
\quad
\text{and}
\quad
\frac{(n-p) \hat{\sigma}^2}{\sigma^2}
\sim
\chi^2_{n-p}
$$

$$
\text{If } \mathbf{X} \sim \mathcal{N}(\boldsymbol{\mu}, \mathbf{\Sigma})
\quad
\text{then}
\quad
(\mathbf{X} - \boldsymbol{\mu})^T \mathbf{\Sigma}^{-1} (\mathbf{X} - \boldsymbol{\mu})
\sim \chi^2_p
$$ 

$$
\text{If }
U_1 \sim \chi^2_{p_1}
\text{ and }
U_2 \sim \chi^2_{p_2}
\text{ independent, then }
\frac{U_1/p_1}{U_2/p_2} \sim \mathcal{F}^{p_1}_{p_2}
$$ 

## Joint distribution of $\hat{\boldsymbol{\beta}}$ - Proof {.build}

$$
\hat{\boldsymbol{\beta}}
\sim
\mathcal{N}\left(
\boldsymbol{\beta};
\sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}
\right)
\qquad\qquad
\\
\qquad\qquad
\implies
\frac{1}{\sigma^2}\left(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}\right)^T (\mathbf{X}^T\mathbf{X})\left(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}\right)
\sim
\chi^2_{p}
$$

$$
\frac{(n-p) \hat{\sigma}^2}{\sigma^2}
\sim
\chi^2_{n-p}
\quad
\text{ independent from } \hat{\boldsymbol{\beta}}
$$

Hence:
$$
\begin{aligned}
\frac{
\frac{1}{\sigma^2}\left(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}\right)^T (\mathbf{X}^T\mathbf{X})\left(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}\right) / p
}{
\frac{(n-p) \hat{\sigma}^2}{\sigma^2} / (n-p)
}
\qquad\qquad\qquad
\\
\qquad\qquad
=
\frac{1}{p\hat{\sigma}^2}\left(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}\right)^T (\mathbf{X}^T\mathbf{X})\left(\hat{\boldsymbol{\beta}} - \boldsymbol{\beta}\right)
&\sim
\mathcal{F}^p_{n-p}.
\end{aligned}
$$

<!-- ************************************************************************ -->
# Global Fisher F-test
<!-- ************************************************************************ -->

## Global Fisher F-test

Hypothesis:
$$
\mathcal{H}_0: \beta_1 = \cdots = \beta_p = 0
\quad
\text{vs}
\quad
\mathcal{H}_1: \exists~k ~|~ \beta_k \neq 0
$$

Test Statistic:
$$
F
= \frac{1}{p\hat{\sigma}^2}\hat{\boldsymbol{\beta}}^T (\mathbf{X}^T\mathbf{X})\hat{\boldsymbol{\beta}}
\underset{\mathcal{H}_0}{\sim}
\mathcal{F}^p_{n-p}
$$

Reject Region:
$$
\mathbb{P}[\text{Reject} | \mathcal{H}_0 \text{ true}] \leq \alpha
\qquad\qquad\qquad\qquad\qquad\\
\qquad\qquad\qquad
\!\iff\! \mathcal{R}_\alpha = \left\{ f \in \mathbb{R}_+  ~|~ f \geq f^p_{n-p}(1 - \alpha) \right\}
$$

p value:$\qquad\qquad p = \mathbb{P}_{\mathcal{H}_0}[F > F^{obs}]$

## Simulated Dataset - Fit - Junk {.smaller}

```{r test7, echo=TRUE}
fit <- lm(y_sim ~ -1 + ., data = data_junk)
f_obs <- summary(fit)$fstatistic
f_obs
p_value_f <- 1 - pf(f_obs[1], f_obs[2], f_obs[3])
p_value_f
```

We do not reject the null hypothesis that all the (junk) coefficients are zero (phew).

## Fisher F-Test - Geometry

$$
\begin{aligned}
F
&= \frac{1}{p\hat{\sigma}^2}\hat{\boldsymbol{\beta}}^T (\mathbf{X}^T\mathbf{X})\hat{\boldsymbol{\beta}}
= \frac{(\mathbf{X}\hat{\boldsymbol{\beta}})^T (\mathbf{X}\hat{\boldsymbol{\beta}}) / p}{\|\hat{\boldsymbol{\epsilon}}\|^2 / (n - p)} \\
&= \frac{\|\hat{\mathbf{y}}\|^2 / p}{\|\hat{\boldsymbol{\epsilon}}\|^2 / (n - p)}
= \frac{\|\hat{\mathbf{y}}\|^2 / p}{\|\mathbf{y} -  \hat{\mathbf{y}}\|^2 / (n - p)} \\
\end{aligned}
$$

>* $\mathcal{H}_0: \beta_1 = \cdots = \beta_p = 0$ i.e. the model is useless.

>* If $\|\hat{\boldsymbol{\epsilon}}\|^2$ is **small**, then the error is **small**,  
and $F$ tends to be **large**, i.e. we tend to **reject** $\mathcal{H}_0$:  
the model is usefull.

>* If $\|\hat{\boldsymbol{\epsilon}}\|^2$ is **large**, then the error is **large**,  
and $F$ tends to be **small**, i.e. we tend to **not reject** $\mathcal{H}_0$:  
the model is rather useless.

## Geometrical interpretation

<div class="columns-2">
**Good model** 

```{r proj2, echo=FALSE, fig.height=2.5, fig.width=2.5, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
ybaronev <- c(0.55, 0.45)
yv <- c(0.8, 0.7)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
# ybaronev <- c(0.7, 0.2)
yhatv <- c(0.8, 0.3)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "firebrick", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(y)), col = "firebrick", pos = 2)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "dodgerblue2", lwd = 2)
text(yhatv[1], yhatv[2],
     labels = expression(hat(bold(y))),
     col = "dodgerblue2", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "dodgerblue2", lwd = 1, lty = 2)
# hat epsilon
text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
     labels = expression(hat(bold(epsilon))), col = "dodgerblue2", pos = 4)
```

* $\mathbf{y}$ **not too far** from $\mathcal{M}(\mathbf{X})$.

* $\|\hat{\boldsymbol{\epsilon}}\|^2$ **not too large** compared to $\|\hat{\mathbf{y}}\|^2$.

* $F$ tends to be large.

$~$

**Bad model** 

```{r proj1, echo=FALSE, fig.height=2.5, fig.width=2.5, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
ybaronev <- c(0.65, 0.45)
yv <- c(0.4, 1.0)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
# ybaronev <- c(0.7, 0.2)
yhatv <- c(0.4, 0.2)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "firebrick", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(y)), col = "firebrick", pos = 2)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "dodgerblue2", lwd = 2)
text(yhatv[1], yhatv[2],
     labels = expression(hat(bold(y))),
     col = "dodgerblue2", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "dodgerblue2", lwd = 1, lty = 2)
# hat epsilon
text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
     labels = expression(hat(bold(epsilon))), col = "dodgerblue2", pos = 4)
```

* $\mathbf{y}$ **almost orth.** to $\mathcal{M}(\mathbf{X})$.

* $\|\hat{\boldsymbol{\epsilon}}\|^2$ **rather large** compared to $\|\hat{\mathbf{y}}\|^2$.

* $F$ tends to be small.

</div>

## Simulated Dataset - True and Junk {.smaller}

```{r test8, echo=TRUE}
## Data frame
data_all <- data.frame(y_sim = y_sim,
                       x_1 = x_1,
                       x_2 = x_2,
                       x_junk = x_junk)
## Fit
fit <- lm(y_sim ~ ., data = data_all)
## multiple t tests
p_values_t <- summary(fit)$coefficients[, 4]
names(p_values_t[p_values_t <= 0.05])

## f test
f_obs <- summary(fit)$fstatistic
p_value_f <- 1 - pf(f_obs[1], f_obs[2], f_obs[3])
p_value_f
```

<!-- ************************************************************************ -->
# Nested Models Fisher F-test
<!-- ************************************************************************ -->

## Nested Models

Can I test ?
$$
\begin{aligned}
\mathcal{H}_0&: \beta_{p_0 + 1} = \cdots = \beta_p = 0 &\text{vs}\\
\mathcal{H}_1&: \exists~k\in \{p_0+1, \dotsc, p\} ~|~ \beta_k \neq 0
\end{aligned}
$$

i.e. decide between the full model:
$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
$$

and the nested model:
$$
\mathbf{y} = \mathbf{X}_0\boldsymbol{\beta}_0 + \boldsymbol{\epsilon}
$$

with $\mathbf{X} = (\mathbf{X}_0 ~ \mathbf{X}_1)$,
$rg(\mathbf{X}) = p$ and
$rg(\mathbf{X}_0) = p_0$.

## Geometrical interpretation

```{r proj3, echo=FALSE, fig.height=5, fig.width=5, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
ybaronev <- c(0.55, 0.45)
yv <- c(0.8, 1.0)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
# ybaronev <- c(0.7, 0.2)
yhatv <- c(0.8, 0.3)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "firebrick", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(y)), col = "firebrick", pos = 2)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "dodgerblue2", lwd = 2)
text(yhatv[1], yhatv[2],
     labels = expression(hat(bold(y))),
     col = "dodgerblue2", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "dodgerblue2", lwd = 1, lty = 2)
# hat epsilon
text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
     labels = expression(hat(bold(epsilon))), col = "dodgerblue2", pos = 4)
# M_0
down <- c(0.25, 0.15)
up <- c(0.6, 0.5)
segments(x0 = down[1], y0 = down[2], x1 = up[1], y1 = up[2],
         col = "darkblue", lwd = 3)
# yhat 0
text(ybaronev[1], ybaronev[2],
     labels = expression(hat(bold(y))[0]),
     col = "darkblue", pos = 2)
# epsilon hat 0
segments(x0 = yv[1], y0 = yv[2], x1 = ybaronev[1], y1 = ybaronev[2],
         col = "darkblue", lwd = 3, lty = 3)
text((yv[1] + ybaronev[1])/2, (yv[2] + ybaronev[2])/2,
     labels = expression(hat(bold(epsilon))[0]), col = "darkblue", pos = 4)
# yhat yhat 0
segments(x0 = ybaronev[1], y0 = ybaronev[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "limegreen", lwd = 3, lty = 3)
```

## Nested Models

**Idea**: Compare

* $\|\hat{\mathbf{y}} -  \hat{\mathbf{y}}_0\|^2$ distance
of $\hat{\mathbf{y}}$ compared to $\hat{\mathbf{y}}_0$ to

* $\|\mathbf{y} - \hat{\mathbf{y}}\|^2$ residual error.

Then:

>* If $\|\hat{\mathbf{y}} - \hat{\mathbf{y}}_0\|^2$
**small** compared to $\|\mathbf{y} -  \hat{\mathbf{y}}\|^2$,
then "$\hat{\mathbf{y}} \approx \hat{\mathbf{y}}_0$" and the null
model is enough.

>* If $\|\hat{\mathbf{y}} - \hat{\mathbf{y}}_0\|^2$
**large** compared to $\|\mathbf{y} - \hat{\mathbf{y}}\|^2$,
then the full model is useful.

## Geometrical interpretation

<div class="columns-2">
**Good model** 

```{r proj4, echo=FALSE, fig.height=2.5, fig.width=2.5, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
ybaronev <- c(0.55, 0.45)
yv <- c(0.8, 1.0)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
# ybaronev <- c(0.7, 0.2)
yhatv <- c(0.8, 0.3)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "firebrick", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(y)), col = "firebrick", pos = 2)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "dodgerblue2", lwd = 2)
text(yhatv[1], yhatv[2],
     labels = expression(hat(bold(y))),
     col = "dodgerblue2", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "dodgerblue2", lwd = 1, lty = 2)
# hat epsilon
# text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
#      labels = expression(hat(bold(epsilon))), col = "dodgerblue2", pos = 4)
# M_0
down <- c(0.25, 0.15)
up <- c(0.6, 0.5)
segments(x0 = down[1], y0 = down[2], x1 = up[1], y1 = up[2],
         col = "darkblue", lwd = 3)
# yhat 0
text(ybaronev[1], ybaronev[2],
     labels = expression(hat(bold(y))[0]),
     col = "darkblue", pos = 2)
# epsilon hat 0
segments(x0 = yv[1], y0 = yv[2], x1 = ybaronev[1], y1 = ybaronev[2],
         col = "darkblue", lwd = 3, lty = 3)
# text((yv[1] + ybaronev[1])/2, (yv[2] + ybaronev[2])/2,
#      labels = expression(hat(bold(epsilon))[0]), col = "darkblue", pos = 4)
# yhat yhat 0
segments(x0 = ybaronev[1], y0 = ybaronev[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "limegreen", lwd = 3, lty = 3)
```

* $\hat{\mathbf{y}}$ **quite different** from $\hat{\mathbf{y}}_0$.

* Full model **does add** information.

$~$

**Bad model** 

```{r proj5, echo=FALSE, fig.height=2.5, fig.width=2.5, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
ybaronev <- c(0.45, 0.35)
yv <- c(0.5, 1.0)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
# ybaronev <- c(0.7, 0.2)
yhatv <- c(0.5, 0.3)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "firebrick", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(y)), col = "firebrick", pos = 2)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "dodgerblue2", lwd = 2)
text(yhatv[1], yhatv[2],
     labels = expression(hat(bold(y))),
     col = "dodgerblue2", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "dodgerblue2", lwd = 1, lty = 2)
# hat epsilon
# text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
#      labels = expression(hat(bold(epsilon))), col = "dodgerblue2", pos = 4)
# M_0
down <- c(0.25, 0.15)
up <- c(0.6, 0.5)
segments(x0 = down[1], y0 = down[2], x1 = up[1], y1 = up[2],
         col = "darkblue", lwd = 3)
# yhat 0
text(ybaronev[1], ybaronev[2],
     labels = expression(hat(bold(y))[0]),
     col = "darkblue", pos = 2)
# epsilon hat 0
segments(x0 = yv[1], y0 = yv[2], x1 = ybaronev[1], y1 = ybaronev[2],
         col = "darkblue", lwd = 3, lty = 3)
# text((yv[1] + ybaronev[1])/2, (yv[2] + ybaronev[2])/2,
#      labels = expression(hat(bold(epsilon))[0]), col = "darkblue", pos = 4)
# yhat yhat 0
segments(x0 = ybaronev[1], y0 = ybaronev[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "limegreen", lwd = 3, lty = 3)
```

* $\hat{\mathbf{y}}$ **similar** to $\hat{\mathbf{y}}_0$.

* Full model **does not add** much information.

</div>

## Nested Models - F test

$$
F = \frac{
\|\hat{\mathbf{y}} -  \hat{\mathbf{y}}_0\|^2 / (p - p_0)
}{
\|\mathbf{y} - \hat{\mathbf{y}}\|^2 / (n - p)
}
$$

>* When $F$ is **large**, full model is **useful**.

>* When $F$ is **small**, null model is enough.

>* Distribution of $F$ ?

## Distribution of $F$ - 1/2 {.build}

$$
F = \frac{
\|\hat{\mathbf{y}} -  \hat{\mathbf{y}}_0\|^2 / (p - p_0)
}{
\|\mathbf{y} - \hat{\mathbf{y}}\|^2 / (n - p)
}
$$

Let $\mathbf{P}$ projection on $\mathcal{M}(\mathbf{X})$ and
$\mathbf{P}_0$ projection on $\mathcal{M}(\mathbf{X}_0)$

$$
\begin{aligned}
\hat{\mathbf{y}} -  \hat{\mathbf{y}}_0
&= \mathbf{P}\mathbf{y} - \mathbf{P}_0\mathbf{y}
= \mathbf{P}\mathbf{y} - \mathbf{P}_0\mathbf{P}\mathbf{y}\\
&= (\mathbf{I}_n - \mathbf{P}_0) \mathbf{P}\mathbf{y}
= \mathbf{P}_0^\bot \mathbf{P}\mathbf{y}\\
&\in \mathcal{M}(\mathbf{X}_0)^\bot \cap \mathcal{M}(\mathbf{X})
\end{aligned}
$$

$$
\mathbf{y} -  \hat{\mathbf{y}}
= (\mathbf{I}_n - \mathbf{P}) \mathbf{y}
= \mathbf{P}^\bot \mathbf{y}
\in \mathcal{M}(\mathbf{X})^\bot
$$

$\hat{\mathbf{y}} -  \hat{\mathbf{y}}_0$ and $\mathbf{y} -  \hat{\mathbf{y}}$ are **orthogonal**  
i.e. their covariance is zero  
i.e. (Gaussian assumption) they are **independent**.

## Distribution of $F$ - 2/2 {.build}

From Cochran's theorem:
$$
\frac{1}{\sigma^2}\|\mathbf{y} -  \hat{\mathbf{y}}\|^2
= \frac{1}{\sigma^2}\|\mathbf{P}^\bot\mathbf{y}\|^2
= \frac{1}{\sigma^2}\|\mathbf{P}^\bot\boldsymbol{\epsilon}\|^2
\sim \chi^2_{n-p}
$$

$$
\frac{1}{\sigma^2}\|\hat{\mathbf{y}} -  \hat{\mathbf{y}}_0 - \mathbf{P}_0^\bot \mathbf{P}\mathbf{X}\boldsymbol{\beta}\|^2
= \frac{1}{\sigma^2}\|\mathbf{P}_0^\bot \mathbf{P}(\mathbf{y} - \mathbf{X}\boldsymbol{\beta})\|^2
\sim \chi^2_{p-p_0}
$$
($\mathbf{P}_0^\bot \mathbf{P}$ is a projection on a space of dimension $p - p_0$)

If $\mathcal{H}_0$ is true, then $\mathbf{P}_0^\bot \mathbf{P}\mathbf{X}\boldsymbol{\beta} = 0$, hence:

$$
F = \frac{
\|\hat{\mathbf{y}} -  \hat{\mathbf{y}}_0\|^2 / (p - p_0)
}{
\|\mathbf{y} - \hat{\mathbf{y}}\|^2 / (n - p)
}
\underset{\mathcal{H}_0}{\sim}
\mathcal{F}^{p - p_0}_{n - p}.
$$

## Nested F test {.build}

From Phythagoras's theorem:
$$
\begin{aligned}
\|\mathbf{y} -  \hat{\mathbf{y}}_0\|^2
&= \|\mathbf{y} - \mathbf{P}\mathbf{y} + \mathbf{P}\mathbf{y} -  \mathbf{P}_0\mathbf{y}\|^2
= \|\mathbf{P}^\bot\mathbf{y} + \mathbf{P}_0^\bot\mathbf{P}\mathbf{y}\|^2\\
&= \|\mathbf{P}^\bot\mathbf{y}\|^2 + \|\mathbf{P}_0^\bot\mathbf{P}\mathbf{y}\|^2
= \|\mathbf{y} -  \hat{\mathbf{y}}\|^2 + \|\hat{\mathbf{y}} -  \hat{\mathbf{y}}_0\|^2\\
\end{aligned}
$$

i.e.
$$
\|\hat{\mathbf{y}} -  \hat{\mathbf{y}}_0\|^2 
= \|\mathbf{y} -  \hat{\mathbf{y}}_0\|^2 - \|\mathbf{y} -  \hat{\mathbf{y}}\|^2
= RSS_0 - RSS
$$

Hence:
$$
F = \frac{
\|\hat{\mathbf{y}} -  \hat{\mathbf{y}}_0\|^2 / (p - p_0)
}{
\|\mathbf{y} - \hat{\mathbf{y}}\|^2 / (n - p)
}
= \frac{n - p}{p - p_0}\frac{RSS_0 - RSS}{RSS}
$$

## Nested F test

To test:
$$
\begin{aligned}
\mathcal{H}_0&: \beta_{p_0 + 1} = \cdots = \beta_p = 0 &\text{vs}\\
\mathcal{H}_1&: \exists~k\in \{p_0+1, \dotsc, p\} ~|~ \beta_k \neq 0
\end{aligned}
$$

we use the $F$ statistics:
$$
F = \frac{
\|\hat{\mathbf{y}} -  \hat{\mathbf{y}}_0\|^2 / (p - p_0)
}{
\|\mathbf{y} - \hat{\mathbf{y}}\|^2 / (n - p)
}
= \frac{n - p}{p - p_0}\frac{RSS_0 - RSS}{RSS}
\underset{\mathcal{H}_0}{\sim}
\mathcal{F}^{p - p_0}_{n - p}.
$$

## Simulated Dataset - True vs Junk {.smaller}

$$
y_i = -1 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

```{r test9, echo=TRUE}
## Data frame
data_all <- data.frame(y_sim = y_sim,
                       x_1 = x_1,
                       x_2 = x_2,
                       x_junk = x_junk)
## Fit true
fit_small <- lm(y_sim ~ x_1 + x_2, data = data_all)
## Fit true and junk
fit_all <- lm(y_sim ~ x_1 + x_2 + x_junk.1 + x_junk.2 + x_junk.3, data = data_all)
## Comparison
anova(fit_small, fit_all)
```

## Simulated Dataset - True vs Junk {.smaller}

$$
y_i = -1 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

```{r test10, echo=TRUE}
## Fit true
fit_small <- lm(y_sim ~ x_1 + x_2, data = data_all)
## Fit all
fit_all <- lm(y_sim ~ ., data = data_all)
## Comparison
anova(fit_small, fit_all)$`Pr(>F)`[2]
```

We do not reject the null hypothesis that the first two variables are enough
to explain the data.

## Full $F$ test

Assuming that we have an intercept, we often test:
$$
\begin{aligned}
\mathcal{H}_0&: \beta_{2} = \cdots = \beta_p = 0 &\text{vs}\\
\mathcal{H}_1&: \exists~k\in \{2, \dotsc, p\} ~|~ \beta_k \neq 0
\end{aligned}
$$

The associated $F$ statistics is ($p_0 = 1$):
$$
F = \frac{
\|\hat{\mathbf{y}} -  \bar{y}\mathbb{1}\|^2 / (p - 1)
}{
\|\mathbf{y} - \hat{\mathbf{y}}\|^2 / (n - p)
}
\underset{\mathcal{H}_0}{\sim}
\mathcal{F}^{p - 1}_{n - p}.
$$

That is the default test returned by `R`.

## Full $F$ test {.smaller}

$$
y_i = -1 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

```{r test11, echo=TRUE}
fit <- lm(y_sim ~ x_1 + x_2)
summary(fit)
```

## One coefficient $F$ test

If $p_0 = p-1$, we test:
$$
\begin{aligned}
\mathcal{H}_0&: \beta_p = 0 &\text{vs} \qquad
\mathcal{H}_1&: \beta_p \neq 0
\end{aligned}
$$

The associated $F$ statistics is ($p_0 = p-1$):
$$
F = \frac{
\|\hat{\mathbf{y}} -  \hat{\mathbf{y}}_{-p}\|^2
}{
\|\mathbf{y} - \hat{\mathbf{y}}\|^2 / (n - p)
}
\underset{\mathcal{H}_0}{\sim}
\mathcal{F}^{1}_{n - p}.
$$

It can be shown that:
$$
F = T^2 \quad \text{with} \quad T = \frac{\hat{\beta}_p}{\hat{\sigma}_p}
$$
so that it is equivalent to the $t$-test.

<!-- ************************************************************************ -->
# Variable Selection
<!-- ************************************************************************ -->

## How to choose the "best" model ? {.build}

**Model**:
$$
y_i = \beta_1 x_{i1} + \dotsb + \beta_p x_{ip} + \epsilon_i
$$

**Problem**:  
Can we choose the "best" subset of non-zero $\beta_k$ ?  

I.e. Can we find the predictors that really have an impact on $\mathbf{y}$ ?

$F$-test on all subsets ? Not practical, multiple testing.

**Idea**:  
Find a "score" to assess the quality of a given fit.

## Variable selection - $R^2$

$$
R^2 = 1 - \frac{RSS}{TSS}
$$

```{r test12, echo=TRUE}
## Function to get statistics for one fit
get_stats_fit <- function(fit) {
  sumfit <- summary(fit)
  res <- data.frame(R2 = sumfit$r.squared)
  return(res)
}
```

$$
y_i = -1 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

## Variable selection - $R^2$ {.smaller}

```{r test13, echo=TRUE}
## Fit true
get_stats_fit(lm(y_sim ~ x_1 + x_2, data = data_all))
## Fit bigger
get_stats_fit(lm(y_sim ~ x_1 + x_2 + x_junk.1, data = data_all))
## Fit wrong
get_stats_fit(lm(y_sim ~ x_1 + x_junk.1, data = data_all))
```


```{r test132, echo=FALSE}
## Function to get statistics for one set of predictors
get_stats_pred <- function(preds, y, data) {
  fit <- lm(reformulate(preds, y), data = data)
  res <- get_stats_fit(fit)
  res$preds <- paste(preds, collapse = "+")
  return(res)
}

## Function to get all statistics
get_all_stats <- function(data_sub) {
  all_preds <- colnames(data_sub)[-1]
  y <- colnames(data_sub)[1]
  all_combs <- c(combn(all_preds, 1, get_stats_pred, simplify = FALSE, y = y, data = data_sub),
                 combn(all_preds, 2, get_stats_pred, simplify = FALSE, y = y, data = data_sub),
                 combn(all_preds, 3, get_stats_pred, simplify = FALSE, y = y, data = data_sub))
  res <- do.call(rbind, all_combs)
  return(res)
}
```

## Variable selection - $R^2$

```{r test133, echo=TRUE}
## Only one "junk" predictor
data_sub <- data.frame(y_sim = y_sim,
                       x_1 = x_1,
                       x_2 = x_2,
                       x_junk = x_junk_1)
## Fit all possible models
get_all_stats(data_sub)
```

**$R^2$ is not enough.**

## Variable selection - $R_a^2$

$$
R_a^2 = 1 - \frac{RSS / (n-p)}{TSS / (n-1)} 
$$

```{r test14, echo=TRUE}
## Function to get statistics
get_stats_fit <- function(fit) {
  sumfit <- summary(fit)
  res <- data.frame(R2 = sumfit$r.squared,
                    R2a = sumfit$adj.r.squared)
  return(res)
}
```

## Variable selection - $R_a^2$

```{r test15, echo=TRUE}
get_all_stats(data_sub)
```

**$R_a^2$: adjust for the number of predictors.**

## Adjusted $R_a^2$

$$
R_a^2 = 1 - \frac{RSS / (n-p)}{TSS / (n-1)} 
$$

* The larger the better.

* When $p$ is bigger, the fit must be really better for $R_a^2$ to raise.

* Intuitive, but not much theoretical justifications.

<!-- ************************************************************************ -->
# Predictive Risk
<!-- ************************************************************************ -->

## Simulation

```{r predrisk1}
set.seed(1289)

## Predictors
n <- 50
x_1 <- runif(n, min = -2, max = 2)
x_2 <- runif(n, min = -2, max = 2)

## Model
beta_0 <- -1; beta_1 <- 3; beta_2 <- -1

## sim
eps <- rnorm(n, mean = 0, sd = 2)
y_sim <- beta_0 + beta_1 * x_1 + beta_2 * x_2 + eps
```

## Simulation - Fit

```{r predrisk2}
## Useless predictors
p_junk <- 48
x_junk <- matrix(runif(n * p_junk, min = -2, max = 2),
                 ncol = p_junk)

## Good and Bad dataset
data_all <- data.frame(y_sim = y_sim,
                       x_1 = x_1,
                       x_2 = x_2,
                       x_junk = x_junk)

## Bad overfitting fit
fit_bad <- lm(y_sim ~ ., data = data_all)
pred_bad <- predict(fit_bad)

## Good fit
fit_good <- lm(y_sim ~ x_1 + x_2, data = data_all)
pred_good <- predict(fit_good)
```

## Simulation - Fit {.smaller}

```{r predrisk23, fig.height=4, fig.width=7, fig.align='center'}
mu_y <- beta_0 + beta_1 * x_1 + beta_2 * x_2
plot(mu_y, y_sim, pch = 3,
     ylab = "y", xlab = expression(beta[0] + beta[1]*x[1] + beta[2]*x[2]))
## Bad prediction (big model)
points(mu_y, pred_bad, col = "firebrick", pch = 4)
## Good prediction (small model)
points(mu_y, pred_good, col = "lightblue", pch = 4)
```

## RSS: "fitted risk"

The bad (bigger) model is closer to the observed data.

```{r predrisk3}
## Distance between bad prediction and data (RSS)
sum((y_sim - pred_bad)^2)

## Distance between good prediction and data (RSS)
sum((y_sim - pred_good)^2)
```

**Overfitting**: the RSS is smaller for the bad, bigger model.

## Predictive Risk

Generate new data according to the same generative model:

```{r predrisk4}
## New dataset from the same model
eps_new <- rnorm(n, mean = 0, sd = 2)
y_new <- beta_0 + beta_1 * x_1 + beta_2 * x_2 + eps_new
```

**Questions**:  
Are the predictions from the bad (bigger) model  
*better or worse*  
than the predictions from the good (smaller) model,  
compared to the new data ?

## Simulation - Fit {.smaller}

```{r predrisk51, fig.height=3.8, fig.width=7, fig.align='center'}
plot(mu_y, y_sim, pch = 3,
     ylab = "y", xlab = expression(beta[0] + beta[1]*x[1] + beta[2]*x[2]))
## Bad prediction (big model)
points(mu_y, pred_bad, col = "firebrick", pch = 4)
## Good prediction (small model)
points(mu_y, pred_good, col = "lightblue", pch = 4)
## New points
points(mu_y, y_new, col = "darkgreen", pch = 3)
```

## Predictive Risk

Compute the risk between predictions and new data:

```{r predrisk5}
## Distance between bad prediction and true
sum((y_new - pred_bad)^2)

## Distance between good prediction and true
sum((y_new - pred_good)^2)
```

The good (smaller) model behaves better on new data.

## Predictive Risk {.build}

**RSS**:
$\|\mathbf{y} - \hat{\mathbf{y}}_p\|^2$ with $\hat{\mathbf{y}}_p$ learnt on $\mathbf{y}$  
$\to$ gets smaller when model gets bigger (overfit).

**Predictive Risk**:
$\| \mathbf{y}' - \hat{\mathbf{y}}_p\|^2$ with $\mathbf{y}'$ same distribution as $\mathbf{y}$.  
$\to$ becomes larger if $\hat{\mathbf{y}}_p$ is overfit.

**Idea**:  
We want to select a model that has a good **predictive** risk.  
i.e. that represents the underlying model best.

**Problem**:  
In general, we don't have means to generate $\mathbf{y}'$.  
$\to$ We use an *estimate* of the (theoretical) predictive risk.

<!-- ************************************************************************ -->
# Theoretical Risk
<!-- ************************************************************************ -->

## Models {.build}

**Ideal model**:
$$
\mathbf{y} = \boldsymbol{\mu} + \boldsymbol{\epsilon}
\quad
\boldsymbol{\mu} \in \mathbb{R}^n
\quad
\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0}, \sigma^2\mathbf{I}_n).
$$

$\boldsymbol{\mu}$ the "true" expectation of the observations $\mathbf{y}$.

**Full linear regression model**:
$$
\mathbf{y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\epsilon}
\quad
rg(\mathbf{X}) = p,
\quad
\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0}, \sigma^2\mathbf{I}_n).
$$

Assume that $\boldsymbol{\mu}$ can be approximated as $\mathbf{X} \boldsymbol{\beta}$,
i.e. as an element of $\mathcal{M}(\mathbf{X})$.

## Models

**Ideal model**:
$$
\mathbf{y} = \boldsymbol{\mu} + \boldsymbol{\epsilon}
\quad
\boldsymbol{\mu} \in \mathbb{R}^n
\quad
\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0}, \sigma^2\mathbf{I}_n).
$$


**Sub-models**:  
keep only predictors $\eta \subset \{1, \dotsc, p\}$
$$
\mathbf{y} = \mathbf{X}_{\eta} \boldsymbol{\beta}_{\eta} + \boldsymbol{\epsilon}_{\eta}
\quad
rg(\mathbf{X}_{\eta}) = |\eta|,
\quad
\boldsymbol{\epsilon}_{\eta} \sim \mathcal{N}(\mathbf{0}, \sigma^2\mathbf{I}_n).
$$

with $\mathbf{X}_{\eta}$ the sub-matrix of $\mathbf{X}$:
$$
\mathbf{X}_{\eta} = (\mathbf{x}_{\eta_1} ~ \mathbf{x}_{\eta_2} \cdots \mathbf{x}_{\eta_{|\eta|}})
$$

Assume that $\boldsymbol{\mu}$ can be approximated as $\mathbf{X}_{\eta} \boldsymbol{\beta}_{\eta}$,
i.e. as an element of $\mathcal{M}(\mathbf{X}_{\eta})$.

## Models

**"Truth"**:
$\mathbf{y} = \boldsymbol{\mu} + \boldsymbol{\epsilon}$, with 
$\boldsymbol{\mu} \in \mathbb{R}^n$ and 
$\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0}, \sigma^2\mathbf{I}_n)$.

**Models:**
$$
\mathbf{y} = \mathbf{X}_{\eta} \boldsymbol{\beta}_{\eta} + \boldsymbol{\epsilon}_{\eta}
\quad
rg(\mathbf{X}_{\eta}) = |\eta|,
\quad
\boldsymbol{\epsilon}_{\eta} \sim \mathcal{N}(\mathbf{0}, \sigma^2\mathbf{I}_n).
$$

**Estimators:**
$$ 
\begin{aligned}
\hat{\boldsymbol{\beta}}_\eta 
& = \underset{\boldsymbol{\beta} \in \mathbb{R}^{|\eta|}}{\operatorname{argmin}} 
\|\mathbf{y} - \mathbf{X}_\eta\boldsymbol{\beta}_\eta \|^2 
= (\mathbf{X}_\eta^{T}\mathbf{X}_\eta)^{-1}\mathbf{X}_\eta^{T}\mathbf{y}\\
\hat{\mathbf{y}}_\eta
& = \mathbf{X}_\eta\hat{\boldsymbol{\beta}}_\eta 
= \mathbf{P}^{\mathbf{X}_\eta}\mathbf{y} 
\end{aligned}
$$

**True expectation**:
$$ 
\begin{aligned}
\mathbf{y}_\eta
= \mathbf{P}^{\mathbf{X}_\eta}\boldsymbol{\mu} 
\end{aligned}
$$

## Models

```{r fullproj, echo=FALSE, fig.height=5.5, fig.width=5.5, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
ybaronev <- c(0.55, 0.45)
yv <- c(0.8, 0.7)
muv <- c(0.6, 0.8)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
# ybaronev <- c(0.7, 0.2)
yhatv <- c(0.8, 0.3)
muetav <- c(0.6, 0.4)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "dodgerblue2", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(y)), col = "dodgerblue2", pos = 4)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "dodgerblue4", lwd = 2)
text(yhatv[1], yhatv[2],
     labels = expression(hat(bold(y))[eta]),
     col = "dodgerblue4", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "dodgerblue4", lwd = 1, lty = 2)
# text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
#      labels = expression(hat(bold(epsilon))[eta]), col = "dodgerblue4", pos = 4)
# mu
arrows(x0 = or[1], y0 = or[2], x1 = muv[1], y1 = muv[2],
       length = 0.1, col = "firebrick2", lwd = 2)
text(muv[1], muv[2], labels = expression(bold(mu)), col = "firebrick2", pos = 2)
# yeta
arrows(x0 = or[1], y0 = or[2], x1 = muetav[1], y1 = muetav[2],
       length = 0.1, col = "firebrick4", lwd = 2)
text(muetav[1], muetav[2],
     labels = expression(bold(y)[eta]),
     col = "firebrick4", pos = 4)
# mu - mueta
segments(x0 = muv[1], y0 = muv[2], x1 = muetav[1], y1 = muetav[2],
         col = "firebrick4", lwd = 1, lty = 2)
# text((muv[1] + muetav[1])/2, (muv[2] + muetav[2])/2,
#      labels = expression(hat(bold(epsilon))[eta]), col = "firebrick4", pos = 4)
# mu - y
segments(x0 = muv[1], y0 = muv[2], x1 = yv[1], y1 = yv[2],
         col = "goldenrod2", lwd = 1, lty = 5)
text((muv[1] + yv[1])/2, (muv[2] + yv[2])/2,
     labels =  expression(bold(mu) - bold(y)), col = "goldenrod2", pos = 3)
```

## Risk

```{r fullproj2, echo=FALSE, fig.height=5.5, fig.width=5.5, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
ybaronev <- c(0.55, 0.45)
yv <- c(0.8, 0.7)
muv <- c(0.6, 0.8)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
# ybaronev <- c(0.7, 0.2)
yhatv <- c(0.8, 0.3)
muetav <- c(0.6, 0.4)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "dodgerblue2", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(y)), col = "dodgerblue2", pos = 4)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "dodgerblue4", lwd = 2)
text(yhatv[1], yhatv[2],
     labels = expression(hat(bold(y))[eta]),
     col = "dodgerblue4", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "dodgerblue4", lwd = 1, lty = 2)
# text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
#      labels = expression(hat(bold(epsilon))[eta]), col = "dodgerblue4", pos = 4)
# mu
arrows(x0 = or[1], y0 = or[2], x1 = muv[1], y1 = muv[2],
       length = 0.1, col = "firebrick2", lwd = 2)
text(muv[1], muv[2], labels = expression(bold(mu)), col = "firebrick2", pos = 2)
# yeta
arrows(x0 = or[1], y0 = or[2], x1 = muetav[1], y1 = muetav[2],
       length = 0.1, col = "firebrick4", lwd = 2)
text(muetav[1], muetav[2],
     labels = expression(bold(y)[eta]),
     col = "firebrick4", pos = 4)
# mu - mueta
segments(x0 = muv[1], y0 = muv[2], x1 = muetav[1], y1 = muetav[2],
         col = "firebrick4", lwd = 1, lty = 2)
# text((muv[1] + muetav[1])/2, (muv[2] + muetav[2])/2,
#      labels = expression(hat(bold(epsilon))[eta]), col = "firebrick4", pos = 4)
# mu - y
# segments(x0 = muv[1], y0 = muv[2], x1 = yv[1], y1 = yv[2],
#          col = "goldenrod2", lwd = 1, lty = 5)
# text((muv[1] + yv[1])/2, (muv[2] + yv[2])/2,
#      labels =  expression(bold(mu) - bold(y)), col = "goldenrod2", pos = 3)
# mu - yhat
segments(x0 = muv[1], y0 = muv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "goldenrod4", lwd = 1, lty = 5)
text((muv[1] + 2*yhatv[1])/3, (muv[2] + 2*yhatv[2])/3,
     labels =  expression(bold(mu) - hat(bold(y))[eta]), col = "goldenrod4", pos = 2)
```

## Risk of an estimator (theoretical)

Risk of an estimator:
$$
R(\hat{\mathbf{y}}_\eta)
= 
\mathbb{E}\left[
\|\boldsymbol{\mu} - \hat{\mathbf{y}}_\eta \|^2
\right]
$$

**Notes**:

>* $\boldsymbol{\mu}$ is unknown $\to$ cannot be computed in practice.

>* Expectation of the difference between the true mean and the estimated projection.

>* Not subject to overfitting.


<!-- ************************************************************************ -->
# Bias-Variance Decomposition
<!-- ************************************************************************ -->

## Bias-Variance Decomposition

$$
R(\hat{\mathbf{y}}_\eta)
= 
\mathbb{E}\left[
\|\boldsymbol{\mu} - \hat{\mathbf{y}}_\eta \|^2
\right]
= 
\|\boldsymbol{\mu} - \mathbf{y}_\eta \|^2 + |\eta| \sigma^2
$$

>* **Bias**: $\|\boldsymbol{\mu} - \mathbf{y}_\eta \|^2$  
>    - Difference between true and projection on the model
>    - Best that we could do if we knew the truth
>    - **Decreases** when model is more complex, i.e. $\eta$ gets larger.
    
>* **Variance**: $|\eta| \sigma^2$  
>    - Variance of the estimation
>    - **Increases** when model is more complex, i.e. $\eta$ gets larger.

>* **Risk is a compromize between bias and variance.**


## Bias-Variance - No overfitting {.build}

$$
R(\hat{\mathbf{y}}_\eta)
= 
\mathbb{E}\left[
\|\boldsymbol{\mu} - \hat{\mathbf{y}}_\eta \|^2
\right]
= 
\|\boldsymbol{\mu} - \mathbf{y}_\eta \|^2 + |\eta| \sigma^2
$$

**No overfitting**: 

Assume:

* $\boldsymbol{\mu} \in \mathcal{M}(\mathbf{X}_{\eta})$ so that 
$\mathbf{y}_{\eta} = \mathbf{P}^{\mathbf{X}_{\eta}}\boldsymbol{\mu} = \boldsymbol{\mu}$.

* $\eta \subset \eta'$ so that 
$\mathbf{y}_{\eta'} = \mathbf{P}^{\mathbf{X}_{\eta'}}\boldsymbol{\mu} = \boldsymbol{\mu}$.

Then:
$$
\begin{aligned}
R(\hat{\mathbf{y}}_{\eta'})
& = \|\boldsymbol{\mu} - \mathbf{y}_{\eta'} \|^2 + |\eta'| \sigma^2 \\
& \geq \|\boldsymbol{\mu} - \mathbf{y}_{\eta'} \|^2 + |\eta| \sigma^2 
 = \|\boldsymbol{\mu} - \mathbf{y}_{\eta} \|^2 + |\eta| \sigma^2 \\
& \geq R(\hat{\mathbf{y}}_\eta)
\end{aligned}
$$

## Bias-Variance - Proof - 1/2 {.build}

$$
R(\hat{\mathbf{y}}_\eta)
= 
\mathbb{E}\left[
\|\boldsymbol{\mu} - \hat{\mathbf{y}}_\eta \|^2
\right]
= 
\mathbb{E}\left[
\|\boldsymbol{\mu} - \mathbf{y}_\eta + \mathbf{y}_\eta - \hat{\mathbf{y}}_\eta \|^2
\right]
$$

$$
\begin{aligned}
R(\hat{\mathbf{y}}_\eta)
&= 
\mathbb{E}\left[
\|\boldsymbol{\mu} - \mathbf{P}^{\mathbf{X}_{\eta}}\boldsymbol{\mu} 
+ \mathbf{P}^{\mathbf{X}_{\eta}}\boldsymbol{\mu} - \mathbf{P}^{\mathbf{X}_{\eta}}\mathbf{y} \|^2
\right] \\
&= 
\mathbb{E}\left[
\|(\mathbf{I}_n - \mathbf{P}^{\mathbf{X}_{\eta}})\boldsymbol{\mu} 
+ \mathbf{P}^{\mathbf{X}_{\eta}}(\boldsymbol{\mu} - \mathbf{y}) \|^2
\right]\\
&= 
\mathbb{E}\left[
\|\mathbf{P}^{\mathbf{X}_{\eta}^\bot}\boldsymbol{\mu} 
+ \mathbf{P}^{\mathbf{X}_{\eta}}(\boldsymbol{\mu} - \mathbf{y}) \|^2
\right]
\end{aligned}
$$

$$
\begin{aligned}
R(\hat{\mathbf{y}}_\eta)
&= 
\mathbb{E}\left[
\|\mathbf{P}^{\mathbf{X}_{\eta}^\bot}\boldsymbol{\mu} \|^2
+ \|\mathbf{P}^{\mathbf{X}_{\eta}}(\boldsymbol{\mu} - \mathbf{y}) \|^2
\right] \\
&= 
\|\mathbf{P}^{\mathbf{X}_{\eta}^\bot}\boldsymbol{\mu} \|^2
+ 
\mathbb{E}\left[
\|\mathbf{P}^{\mathbf{X}_{\eta}}(\boldsymbol{\mu} - \mathbf{y}) \|^2
\right] \\
&= 
\|\boldsymbol{\mu} - \mathbf{y}_\eta \|^2
+ 
\mathbb{E}\left[
\|\mathbf{P}^{\mathbf{X}_{\eta}}(\boldsymbol{\mu} - \mathbf{y}) \|^2
\right]
\end{aligned}
$$

## Bias-Variance - Proof - 2/2 {.build}

$$
R(\hat{\mathbf{y}}_\eta)
= 
\|\boldsymbol{\mu} - \mathbf{y}_\eta \|^2
+ 
\mathbb{E}\left[
\|\mathbf{P}^{\mathbf{X}_{\eta}}(\boldsymbol{\mu} - \mathbf{y}) \|^2
\right]
$$

From Cochran's Theorem:
$$
\frac{1}{\sigma^2}\|\mathbf{P}^{\mathbf{X}_{\eta}}(\boldsymbol{\mu} - \mathbf{y}) \|^2
\sim
\chi^2_{|\eta|}
$$

Hence:
$$
\mathbb{E}\left[
\frac{1}{\sigma^2}\|\mathbf{P}^{\mathbf{X}_{\eta}}(\boldsymbol{\mu} - \mathbf{y}) \|^2
\right]
=
|\eta|
$$

And:
$$
R(\hat{\mathbf{y}}_\eta)
= 
\|\boldsymbol{\mu} - \hat{\mathbf{y}}_\eta \|^2
+ 
|\eta|\sigma^2.
$$

<!-- ************************************************************************ -->
# Penalized Least Squares 
<!-- ************************************************************************ -->

## Oracle Estimator

We would like:
$$
\eta_0 = \underset{\eta \subset \{1, \cdots, n\}}{\operatorname{argmin}} R(\hat{\mathbf{y}}_\eta)
$$

>* Minimize the risk over all possible models.

>* **Problem**: $R(\hat{\mathbf{y}}_\eta)$ is theoretical $\to$ cannot be computed.

>* **Idea**: Minimize an *estimator* $\hat{R}(\hat{\mathbf{y}}_\eta)$ of $R(\hat{\mathbf{y}}_\eta)$:
$$
\hat{\eta} = \underset{\eta \subset \{1, \cdots, n\}}{\operatorname{argmin}} \hat{R}(\hat{\mathbf{y}}_\eta).
$$

## Expectation of the RSS

For any model $\eta$, the expectation of the RSS is:
$$
\mathbb{E}[\|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2]
=
R(\hat{\mathbf{y}}_\eta) + n\sigma^2 - 2|\eta|\sigma^2
$$

## Expectation of the RSS - Proof {.build}

$$
\begin{aligned}
\mathbb{E}[\|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2]
&= \mathbb{E}[\|\mathbf{y} - \boldsymbol{\mu} + \boldsymbol{\mu} - \hat{\mathbf{y}}_\eta \|^2] \\
&= 
\mathbb{E}[\|\mathbf{y} - \boldsymbol{\mu}\|^2 
+ \|\boldsymbol{\mu} - \hat{\mathbf{y}}_\eta \|^2 
+ 2 \langle \mathbf{y} - \boldsymbol{\mu}; \boldsymbol{\mu} - \hat{\mathbf{y}}_\eta \rangle]
\end{aligned}
$$

$$
\begin{aligned}
\mathbb{E}[\|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2]
&= 
\mathbb{E}[\|\boldsymbol{\epsilon}\|^2]
+ R(\hat{\mathbf{y}}_\eta)
+ 2 \mathbb{E}[\langle \boldsymbol{\epsilon}; \boldsymbol{\mu} - \mathbf{P}^{\mathbf{X}_{\eta}}\mathbf{y} \rangle]
\qquad
\end{aligned}
$$

$$
\begin{aligned}
\mathbb{E}[\langle \boldsymbol{\epsilon}; \boldsymbol{\mu} - \mathbf{P}^{\mathbf{X}_{\eta}}\mathbf{y} \rangle]
&=
\mathbb{E}[\langle \boldsymbol{\epsilon}; \boldsymbol{\mu} - \mathbf{P}^{\mathbf{X}_{\eta}}(\boldsymbol{\mu} + \boldsymbol{\epsilon}) \rangle] \\
&= 
\mathbb{E}[\langle \boldsymbol{\epsilon}; \boldsymbol{\mu} - \mathbf{P}^{\mathbf{X}_{\eta}}\boldsymbol{\mu}) \rangle]
- \mathbb{E}[\langle \boldsymbol{\epsilon}; \mathbf{P}^{\mathbf{X}_{\eta}}\boldsymbol{\epsilon} \rangle]
\end{aligned}
$$

$$
\begin{aligned}
\mathbb{E}[\langle \boldsymbol{\epsilon}; \boldsymbol{\mu} - \mathbf{P}^{\mathbf{X}_{\eta}}\mathbf{y} \rangle]
&= 
0
- \mathbb{E}[\langle \mathbf{P}^{\mathbf{X}_{\eta}}\boldsymbol{\epsilon}; \mathbf{P}^{\mathbf{X}_{\eta}}\boldsymbol{\epsilon} \rangle] \qquad\qquad~~ \\
&= - \mathbb{E}[\| \mathbf{P}^{\mathbf{X}_{\eta}}\boldsymbol{\epsilon}\|^2] \\
&= - |\eta| \sigma^2
\end{aligned}
$$

$$
\begin{aligned}
\mathbb{E}[\|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2]
&= 
n \sigma^2
+ R(\hat{\mathbf{y}}_\eta)
- 2 |\eta| \sigma^2
\qquad\qquad\qquad\qquad
\end{aligned}
$$


## Mallow’s $C_p$

We have:
$$
\mathbb{E}[\|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2]
=
R(\hat{\mathbf{y}}_\eta) + n\sigma^2 - 2|\eta|\sigma^2
$$

Mallow's $C_p$:
$$
C_p(\hat{\mathbf{y}}_\eta) = \frac{1}{n} \left( \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2 + 2 |\eta| \hat{\sigma}^2 \right)
$$

>* If $\hat{\sigma}^2$ is unbiased, then:
$$
\mathbb{E}[C_p(\hat{\mathbf{y}}_\eta)] 
= \frac{1}{n}R(\hat{\mathbf{y}}_\eta) + \sigma^2
$$

## Mallow’s $C_p$ {.build}

We have:
$$
C_p(\hat{\mathbf{y}}_\eta) = \frac{1}{n} \left( \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2 + 2 |\eta| \hat{\sigma}^2 \right)
$$
And:
$$
\mathbb{E}[C_p(\hat{\mathbf{y}}_\eta)] 
= \frac{1}{n}R(\hat{\mathbf{y}}_\eta) + \sigma^2
$$

Hence:
$$
\hat{\eta} 
= \underset{\eta \subset \{1, \cdots, n\}}{\operatorname{argmin}} C_p(\hat{\mathbf{y}}_\eta)
= \underset{\eta \subset \{1, \cdots, n\}}{\operatorname{argmin}} \hat{R}(\hat{\mathbf{y}}_\eta).
$$

$\to$ Minimizing Mallow's $C_p$ is the same as minimizing an unbiased estimate of the risk.

## Mallow's $C_p$

$$
C_p(\hat{\mathbf{y}}_\eta) = \frac{1}{n} \left( \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2 + 2 |\eta| \hat{\sigma}^2 \right)
$$

* Penalize for the number of parameters $|\eta|$.

* The **smaller** the better.

## Simulated Dataset

```{r test161, echo=TRUE}
## Mallow's Cp
Cp <- function(fit) {
  n <- nobs(fit)
  p <- n - df.residual(fit)
  RSS <- deviance(fit)
  return( (RSS + p * RSS / (n - p)) / n)
}
## Function to get statistics
get_stats_fit <- function(fit) {
  sumfit <- summary(fit)
  res <- data.frame(Cp = Cp(fit))
  return(res)
}
```

## Variable selection

```{r test162, echo=TRUE}
## Apply stats to all possible combinations
all_stats <- get_all_stats(data_sub)
all_stats
```

Mallow's $C_p$ is **smallest** for the true model.

<!-- ************************************************************************ -->
# Penalized Log Likelihood 
<!-- ************************************************************************ -->

## Maximized Log Likelihood {.build}

Recall:
$$
\begin{aligned}
(\hat{\boldsymbol{\beta}}_\eta, \hat{\sigma}^2_{\eta,ML}) 
&= \underset{(\boldsymbol{\beta}, \sigma^2) \in \mathbb{R}^{|\eta|}\times\mathbb{R}_+^*}{\operatorname{argmax}} 
\log L(\boldsymbol{\beta}, \sigma^2 | \mathbf{y})\\
&= \underset{(\boldsymbol{\beta}, \sigma^2) \in \mathbb{R}^{|\eta|}\times\mathbb{R}_+^*}{\operatorname{argmax}} 
- \frac{n}{2} \log(2\pi\sigma^2) 
- \frac{1}{2\sigma^2}\|\mathbf{y} - \mathbf{X}_\eta\boldsymbol{\beta} \|^2
\end{aligned}
$$

And: 
$$
\hat{\sigma}^2_{\eta,ML} = \frac{ \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2} {n}
\quad;\quad
\hat{\mathbf{y}}_\eta = \mathbf{X}_\eta\hat{\boldsymbol{\beta}}_\eta
$$

Thus:
$$
L(\hat{\boldsymbol{\beta}}_\eta, \hat{\sigma}^2_{ML} | \mathbf{y})
= 
\cdots
$$

## Maximized Log Likelihood {.build}

The maximized likelihood is equal to:
$$
L(\hat{\boldsymbol{\beta}}_\eta, \hat{\sigma}^2_{ML} | \mathbf{y})
= 
- \frac{n}{2} \log\left(2\pi\frac{ \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2}{n}\right) 
- \frac{1}{2\frac{ \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2}{n}}\|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2
$$

i.e.
$$
\log \hat{L}_\eta
= 
- \frac{n}{2} \log\left(\frac{ \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2}{n}\right) 
- \frac{n}{2} \log\left(2\pi\right) 
- \frac{n}{2}
$$

## Maximized Log Likelihood is increasing {.build}

Remember, for any $\eta \subset \eta'$:
$$
\|\mathbf{y} - \hat{\mathbf{y}}_{\eta'}\|^2 \leq \|\mathbf{y} - \hat{\mathbf{y}}_\eta\|^2
$$

Hence
$R^2_\eta = 1 - \frac{\|\mathbf{y} - \hat{\mathbf{y}}_\eta\|^2}{\|\mathbf{y} - \bar{y} \mathbb{1}\|^2}$
always **increases** when we add a predictor.

Similarly:
$$
\log \hat{L}_\eta
= 
- \frac{n}{2} \log\left(\frac{ \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2}{n}\right) 
- \frac{n}{2} \log\left(2\pi\right) 
- \frac{n}{2}
$$
always **increases** when we add some predictors.

## Penalized Maximized Log Likelihood {.build}

**Problem:** $\log \hat{L}_\eta$ always **increases** when we add some predictors.

$\to$ it can not be used for model selection (same as $R^2$).

**Idea**: add a penalty term that depends on the size of the model

$\to$ minimize:
$$
- \log \hat{L}_\eta + f(|\eta|) \qquad \text{with f increasing}
$$

## Penalized Maximized Log Likelihood {.build}

Minimize:
$$
- \log \hat{L}_\eta + f(|\eta|) \qquad \text{with f increasing}
$$

For $\eta \subset \eta'$:

* $- \log \hat{L}_\eta$ decreases

* $f(|\eta|)$ increases

**Goal**: find $f$ that compensates for the "overfit", i.e. the noisy and insignificant
increase of the likelihood when we add some unrelated predictors.

## Simulated Dataset

```{r test171, echo=TRUE}
## Function to get statistics
get_stats_fit <- function(fit) {
  sumfit <- summary(fit)
  res <- data.frame(minusLogLik = -logLik(fit))
  return(res)
}
```

## Variable selection

```{r test181, echo=TRUE}
## Apply stats to all possible combinations
all_stats <- get_all_stats(data_sub)
all_stats
```

When we add "junk" variable, $- \log \hat{L}_\eta$ decreases, but not a lot.

**Questions**: can we find a good penalty that compensates for this noisy increasing ?

<!-- ************************************************************************ -->
# AIC and BIC
<!-- ************************************************************************ -->

## Akaike Information Criterion - AIC

$$
AIC = - 2\log \hat{L}_\eta + 2 |\eta|
$$

$~$

* $\hat{L}_\eta$ is the **maximized likelihood** of the model.

* $|\eta|$ is the **dimension** of the model.

* The **smaller** the better.

* Asymptotic theoretical guaranties.

* Easy to use and **widely spread**

## Bayesian Information Criterion - BIC

$$
BIC = - 2\log \hat{L}_\eta + |\eta| \log(n)
$$

$~$

* $\hat{L}_\eta$ is the **maximized likelihood** of the model.

* $|\eta|$ is the **dimension** of the model.

* The **smaller** the better.

* Based on an approximation of the marginal likelihood.

* Easy to use and **widely spread**

See: Lebarbier, Mary-Huard. *Le critère BIC : fondements théoriques et interprétation.* 2004. [[inria-00070685]](https://hal.archives-ouvertes.fr/inria-00070685/)


## AIC vs BIC

$$
AIC = - 2\log \hat{L}_\eta + 2 |\eta|
$$

$$
BIC = - 2\log \hat{L}_\eta + |\eta| \log(n)
$$

* Both have asymptotic theoretical justifications.

* BIC penalizes more than AIC $\to$ chooses smaller models.

* Both **widely used**.

* There are some (better) non-asymptotic criteria, see  
Giraud C. 2014. *Introduction to high-dimensional statistics.*

## Simulated Dataset

```{r test17, echo=TRUE}
## Function to get statistics
get_stats_fit <- function(fit) {
  sumfit <- summary(fit)
  res <- data.frame(minusLogLik = -logLik(fit),
                    AIC = AIC(fit),
                    BIC = BIC(fit))
  return(res)
}
```

## Variable selection {.smaller}

```{r test18, echo=TRUE}
## Apply stats to all possible combinations
all_stats <- get_all_stats(data_sub)
all_stats

## Select model
apply(all_stats[, -ncol(all_stats)], 2,
      function(x) all_stats[which.min(x), ncol(all_stats)])
```

## Variable selection - Advertising data {.smaller}

```{r test19, message = FALSE}
## Advertising data
library(here)
ad <- read.csv(here("data", "Advertising.csv"), row.names = "X")

## Apply stats to all possible combinations
all_stats <- get_all_stats(ad[, c(4, 1:3)])
all_stats

## Select model
apply(all_stats[, -ncol(all_stats)], 2,
      function(x) all_stats[which.min(x), ncol(all_stats)])
```


<!-- ************************************************************************ -->
# Forward and Backward Search
<!-- ************************************************************************ -->

## Combinatorial problem

$$
p \text{ predictors } \to 2^p \text{ possible models.}
$$

>* Cannot test all possible models.

>* "Manual" solution: forward or backward searches.

## Forward search

>* Start with the null model (no predictor)

>* Try to **add one** predictor ($p$ models to fit)

>* Choose the best fitting among the $p$ models.

>* Repeat until no more predictors to add.

>* Choose best fit with $C_p$, AIC or BIC.

## Forward search - Init

```{r forward, message = FALSE}
## Full formula
full_formula <- formula(y_sim ~ x_1 + x_2 
                        + x_junk.1 + x_junk.2 + x_junk.3 
                        + x_junk.4 + x_junk.5 + x_junk.6)

## Complete search
2^8

## No predictors
fit0 <- lm(y_sim ~ 1, data = data_all)
```

## Forward search - Add term

```{r forward1, message = FALSE}
## Add one
library(MASS)
addterm(fit0, full_formula)
```

## Forward search - Add term

```{r forward2, message = FALSE}
## Add x_1
fit1 <- lm(y_sim ~ 1 + x_1, data = data_all)
## Add another
addterm(fit1, full_formula)
```

## Forward search - Add term

```{r forward3, message = FALSE}
## Add x_1
fit2 <- lm(y_sim ~ 1 + x_1 + x_2, data = data_all)
## Add another
addterm(fit2, full_formula)
```

## Backward search

>* Start with the full model (all predictors)

>* Try to **remove one** predictor ($p$ models to fit)

>* Choose the best fitting among the $p$ models.

>* Repeat until no more predictors to remove.

>* Choose best fit with $C_p$, AIC or BIC.

## Backward search - Init {.smaller}

```{r backward, message = FALSE}
## Full formula
fit0 <- lm(y_sim ~ x_1 + x_2 
                        + x_junk.1 + x_junk.2 + x_junk.3 
                        + x_junk.4 + x_junk.5 + x_junk.6,
           data = data_all)
```

```{r backward1, message = FALSE}
## drop one
dropterm(fit0)
```

## Backward search - Drop term {.smaller}

```{r backward2, message = FALSE}
## drop x_junk.4
fit1 <- lm(y_sim ~ x_1 + x_2 
           + x_junk.1 + x_junk.2 + x_junk.3 
           + x_junk.5 + x_junk.6, data = data_all)
## Drop another
dropterm(fit1)
```

## Backward search - Drop term {.smaller}

```{r backward3, message = FALSE}
## drop x_junk.3
fit2 <- lm(y_sim ~ x_1 + x_2 
           + x_junk.1 + x_junk.3 
           + x_junk.5 + x_junk.6, data = data_all)
## Drop another
dropterm(fit2)
```

## Backward search - Drop term {.smaller}

```{r backward4, message = FALSE}
## drop x_junk.5
fit3 <- lm(y_sim ~ x_1 + x_2 
           + x_junk.1 + x_junk.3 
           + x_junk.6, data = data_all)
## Drop another
dropterm(fit3)
```

## Backward search - Drop term {.smaller}

```{r backward5, message = FALSE}
## drop x_junk.6
fit4 <- lm(y_sim ~ x_1 + x_2 
           + x_junk.1 + x_junk.3 ,
           data = data_all)
## Drop another
dropterm(fit4)
```

## Backward search - Drop term {.smaller}

```{r backward6, message = FALSE}
## drop x_junk.5
fit5 <- lm(y_sim ~ x_1 + x_2 + x_junk.3 ,
           data = data_all)
## Drop another
dropterm(fit5)
```

## Backward search - Drop term {.smaller}

```{r backward7, message = FALSE}
## drop x_junk.3
fit6 <- lm(y_sim ~ x_1 + x_2,
           data = data_all)
## Drop another
dropterm(fit6)
```


## Forward and backward searches

>* Both are **heuristics**.

>* There are no guaranties that both converge to the same solution.

>* No theoretical guaranties on the selected model.

>* Model selection is still an active area of research.
>   - See e.g: LASSO and other penalized criteria (M2)

<!-- ************************************************************************ -->
# AIC vs log FPE
<!-- ************************************************************************ -->

## Final Prediction Error - FPE

Mallow's $C_p$:
$$
C_p(\hat{\mathbf{y}}_\eta) = \frac{1}{n} \left( \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2 + 2 |\eta| \hat{\sigma}^2_{\text{full}} \right)
$$

FPE:
$$
FPE(\hat{\mathbf{y}}_\eta) = \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2 + 2 |\eta| \hat{\sigma}^2_\eta 
\quad
\text{with}
\quad
\hat{\sigma}^2_\eta = \frac{ \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2} {n - |\eta|}
$$

Hence:
$$
FPE(\hat{\mathbf{y}}_\eta) = \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2 
\left(1 + \frac{2 |\eta|} {n - |\eta|}\right) 
$$

## AIC vs Log FPE {.build}

$$
\begin{aligned}
\log \frac1n FPE(\hat{\mathbf{y}}_\eta) 
&= \log 
\left(\frac1n \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2 
\left(1 + \frac{2 |\eta|} {n - |\eta|}\right) \right)\\
&= \log 
\left(\frac1n \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2 \right)
+ \log \left(1 + \frac{2 |\eta|} {n - |\eta|}\right)
\end{aligned}
$$
But:
$$
n \log\left(\frac{ \|\mathbf{y} - \hat{\mathbf{y}}_\eta \|^2}{n}\right) 
= 
- 2\log \hat{L}_\eta
- n \log\left(2\pi\right) 
- n
$$

## AIC vs Log FPE {.build}

Hence:
$$
n\log \frac1n FPE(\hat{\mathbf{y}}_\eta) + n \log\left(2\pi\right) + n
\qquad \qquad \qquad \qquad\\
\qquad \qquad \qquad \qquad = - 2\log \hat{L}_\eta + n \log \left(1 + \frac{2 |\eta|} {n - |\eta|}\right)\\
$$

When $n$ is large:
$$
n \log \left(1 + \frac{2 |\eta|} {n - |\eta|}\right)
\approx
n \frac{2 |\eta|} {n - |\eta|}
\approx
2 |\eta|
$$

## AIC vs Log FPE {.build}

Hence:
$$
\begin{aligned}
n\log \frac1n FPE(\hat{\mathbf{y}}_\eta) + n \log\left(2\pi\right) + n 
&\approx - 2\log \hat{L}_\eta + 2 |\eta| \\
&\approx AIC(\hat{\mathbf{y}}_\eta)
\end{aligned}
$$

The log FPE and the AIC criteria select for the same models (asymptotically).

<!-- ************************************************************************ -->
# BIC and Laplace Approximation
<!-- ************************************************************************ -->

## Bayesian Model Selection {.build}

In a **Bayesian setting**,  
we have priors on the models $\eta$ and parameters $\boldsymbol{\theta}_\eta = (\boldsymbol{\beta}_\eta, \sigma^2_\eta)$:
$$
\left\{
\begin{aligned}
\eta &\sim \pi_{\eta} \\
\boldsymbol{\theta}_\eta & \sim p(\boldsymbol{\theta} | \eta)
\end{aligned}
\right.
$$

The **likelihood** given some parameters and a model is:
$p(\mathbf{y} | \boldsymbol{\theta}, \eta)$

the **marginal likelihood** given a model is:
$$
p(\mathbf{y} | \eta)
= 
\int p(\mathbf{y} | \boldsymbol{\theta}, \eta) p(\boldsymbol{\theta} | \eta) d\boldsymbol{\theta}
$$

and the **posterior** probability of the models are:
$$
p(\eta | \mathbf{y}).
$$

## Bayesian Model Selection {.build}

**Goal**: find the model with the *highest* posterior probability:
$$
\hat{\eta} 
= \underset{\eta}{\operatorname{argmax}} 
p(\eta | \mathbf{y})
$$

Assume that all models have the same prior probability:
$$
\pi_{\eta} \equiv \pi.
$$

Then, from Bayes rule:
$$
p(\eta | \mathbf{y})
=
p(\mathbf{y} | \eta)\pi_{\eta} / p(\mathbf{y})
=
p(\mathbf{y} | \eta) \times \pi / p(\mathbf{y})
$$

And:
$$
\hat{\eta} 
= \underset{\eta}{\operatorname{argmax}} 
p(\eta | \mathbf{y})
= \underset{\eta}{\operatorname{argmax}} 
p(\mathbf{y} | \eta).
$$

## Bayesian Model Selection {.build}

**Goal**: find the model with the *highest* posterior probability:
$$
\begin{aligned}
\hat{\eta} 
& = \underset{\eta}{\operatorname{argmax}} 
p(\eta | \mathbf{y}) 
 = \underset{\eta}{\operatorname{argmax}} 
p(\mathbf{y} | \eta) \\
& = \underset{\eta}{\operatorname{argmax}} 
\int p(\mathbf{y} | \boldsymbol{\theta}, \eta) p(\boldsymbol{\theta} | \eta) d\boldsymbol{\theta}
\end{aligned}
$$

**Idea**:  
Assume that 
$$p(\mathbf{y} | \boldsymbol{\theta}, \eta) p(\boldsymbol{\theta} | \eta)$$
is concentrated around its maximum 
$$p(\mathbf{y} | \boldsymbol{\theta}_\eta^*, \eta) p(\boldsymbol{\theta}_\eta^* | \eta)$$
and apply Laplace Approximation.

## Laplace Approximation {.build}

Let $L: \mathbb{R}^q \to \mathbb{R}$ be a three times differentiable function,
with a unique maximum $\mathbf{u}^*$. Then:
$$
\int_{\mathbb{R}^q} e^{n L(\mathbf{u})} d \mathbf{u}
=
\left(\frac{2\pi}{n}\right)^{q/2} |-L''(\mathbf{u}^*)|^{-1/2}e^{nL(\mathbf{u}^*)}
+ o(n^{-1})
$$

Applied to:
$$
L(\boldsymbol{\theta}) 
= \frac1n\log p(\mathbf{y} | \boldsymbol{\theta}, \eta) + \frac1n\log p(\boldsymbol{\theta} | \eta)
$$
and assuming 
$\log p(\mathbf{y} | \boldsymbol{\theta}^*, \eta) \approx \log p(\mathbf{y} | \hat{\boldsymbol{\theta}}, \eta)$,
we get:
$$
\log \int p(\mathbf{y} | \boldsymbol{\theta}, \eta) p(\boldsymbol{\theta} | \eta) d\boldsymbol{\theta}
\approx
\log p(\mathbf{y} | \hat{\boldsymbol{\theta}}, \eta) - \frac{|\eta|}{2} \log(n).
$$

## BIC

Hence:
$$
\begin{aligned}
\hat{\eta} 
& = \underset{\eta}{\operatorname{argmax}} 
p(\eta | \mathbf{y}) 
 = \underset{\eta}{\operatorname{argmax}} 
p(\mathbf{y} | \eta) \\
& \approx \underset{\eta}{\operatorname{argmax}} 
\left\{\log p(\mathbf{y} | \hat{\boldsymbol{\theta}}, \eta) - \frac{|\eta|}{2} \log(n)\right\}\\
& \approx \underset{\eta}{\operatorname{argmin}} 
\left\{-2\log p(\mathbf{y} | \hat{\boldsymbol{\theta}}, \eta) + |\eta| \log(n) \right\}\\
&\approx \underset{\eta}{\operatorname{argmax}}
\left\{- 2\log \hat{L}_\eta + |\eta| \log(n)\right\}
= \underset{\eta}{\operatorname{argmax}}
 BIC(\hat{\mathbf{y}}_\eta)
\end{aligned}
$$

BIC is an approximation of the marginal likelihood.