---
title: 'Régression Multiple - Moindres Carrés'
author: "Paul Bastide - Benjamin Charlier"
date: "01/02/2022"
output:
  ioslides_presentation:
    fig_width: 7
    fig_height: 4
  self_contained: true
---

<!-- ************************************************************************ -->
# Advertising Data
<!-- ************************************************************************ -->

## Advertising - Data

```{r ad1, message = FALSE}
library(here)
ad <- read.csv(here("data", "Advertising.csv"), row.names = "X")
head(ad)
```

* `TV`, `radio`, `newspaper`: advertising budgets (thousands of $)
* `sales`: number of sales (thousands of units)

## Advertising - Data

```{r ad2}
attach(ad)
par(mfrow = c(1, 3))
plot(TV, sales); plot(radio, sales); plot(newspaper, sales)
```

## Advertising - Questions

* Previous chapters: use only **one** medium to predict `sales`  
    `sales` $\approx \beta_0^{TV} + \beta_1^{TV}$ `TV`  
    **or**  
    `sales` $\approx \beta_0^{radio} + \beta_1^{radio}$ `radio`  
    **or**  
    `sales` $\approx \beta_0^{newspaper} + \beta_1^{newspaper}$ `newspaper`

>* Can we use **all** variables at once ?  
    `sales` $\approx \beta_0 + \beta_1$ `TV` $+ \beta_2$ `radio` $+ \beta_3$ `newspaper`

>* Makes it possible to answer new questions:  
>   - Is the global fit better ?
>   - Which media contribute to sales ?
>   - Is there synergy among the media ?


<!-- ************************************************************************ -->
# From Simple to Multiple Regression
<!-- ************************************************************************ -->

## Model - Simple Regression

$$y_i = \beta_0 + \beta_1 x_i + \epsilon_i, \quad \forall 1 \leq i \leq n$$

* $y_i$: quantitative response for $i$ (`sales`)
* $x_i$: quantitative predicting variable for $i$ (`TV`)
* $\epsilon_i$: "error" for $i$
  * **random variable**
  * (H1) $\mathbb{E}[\epsilon_i] = 0$ for all $i$ [(centered)]{style="float:right"}
  * (H2) $\mathbb{Var}[\epsilon_i] = \sigma^2$ for all $i$ [(identical variance)]{style="float:right"}
  * (H3) $\mathbb{Cov}[\epsilon_i; \epsilon_j] = 0$ for all $i \neq j$ [(independent)]{style="float:right"}


## Model - Multiple Regression

$$y_i = \beta_0 + \beta_1 x_{i1} + \dotsb + \beta_p x_{ip} + \epsilon_i, \quad \forall 1 \leq i \leq n$$

* $y_i$: quantitative response for $i$ (`sales`)
* $x_{ij}$: quantitative predicting variable $j$ for observation $i$  
(`TV`, `radio` and `newspaper`, i.e. $p=3$)
* $\epsilon_i$: "error" for $i$
  * **random variable**
  * (H1) $\mathbb{E}[\epsilon_i] = 0$ for all $i$ [(centered)]{style="float:right"}
  * (H2) $\mathbb{Var}[\epsilon_i] = \sigma^2$ for all $i$ [(identical variance)]{style="float:right"}
  * (H3) $\mathbb{Cov}[\epsilon_i; \epsilon_j] = 0$ for all $i \neq j$ [(independent)]{style="float:right"}
  
## Simple Regression - Vectorial notation {.build}

$$
\begin{pmatrix} y_1 \\ \vdots \\ y_n \end{pmatrix} 
= \beta_0 \begin{pmatrix} 1 \\ \vdots \\ 1 \end{pmatrix} 
+ \beta_1 \begin{pmatrix} x_1 \\ \vdots \\ x_n \end{pmatrix} 
+ \begin{pmatrix} \epsilon_1 \\ \vdots \\ \epsilon_n \end{pmatrix} 
$$
i.e.
$$\mathbf{y} = \beta_0 \mathbb{1} + \beta_1 \mathbf{x} + \boldsymbol{\epsilon}$$

* $\mathbf{y} = (y_1, \dotsc, y_n)^T$ **random** vector of responses
* $\mathbb{1} = (1, \dotsc, 1)^T$ vector of ones
* $\mathbf{x} = (x_1, \dotsc, x_n)^T$ **non random** vector of predictors
* $\boldsymbol{\epsilon} = (\epsilon_1, \dotsc, \epsilon_n)^T$ **random** vector of errors
* $\beta_0$, $\beta_1$ **non random, unknown** coefficients

## Multiple Regression - Matricial notation {.build}

$$y_i = \beta_0 + \beta_1 x_{i1} + \dotsb + \beta_p x_{ip} + \epsilon_i, \quad \forall 1 \leq i \leq n$$

$$
\begin{pmatrix} y_1 \\ \vdots \\ y_n \end{pmatrix} 
= \beta_0 \begin{pmatrix} 1 \\ \vdots \\ 1 \end{pmatrix} 
+ \beta_1 \begin{pmatrix} x_{11} \\ \vdots \\ x_{n1} \end{pmatrix} 
+ \dotsb
+ \beta_p \begin{pmatrix} x_{1p} \\ \vdots \\ x_{np} \end{pmatrix} 
+ \begin{pmatrix} \epsilon_1 \\ \vdots \\ \epsilon_n \end{pmatrix} 
$$

$$
\begin{pmatrix} y_1 \\ \vdots \\ y_n \end{pmatrix} 
= \begin{pmatrix} 1 & x_{11} & \dotsc & x_{1p}\\ \vdots \\ 1 &  x_{n1} & \dotsc & x_{np} \end{pmatrix} 
\begin{pmatrix} \beta_0 \\ \beta_1 \\ \vdots \\ \beta_p \end{pmatrix} 
+ \begin{pmatrix} \epsilon_1 \\ \vdots \\ \epsilon_n \end{pmatrix} 
$$

$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
$$

## Multiple Regression - Matricial notation

$$y_i = \beta_0 + \sum_{k = 1}^p \beta_k x_{ik} + \epsilon_i, \quad \forall 1 \leq i \leq n$$

i.e
$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
$$

* $\mathbf{y}$ **random** vector of responses
* $\mathbf{X}$ **non random** matrix of predictors
* $\boldsymbol{\epsilon}$ **random** vector of errors
* $\boldsymbol{\beta}$ **non random, unknown** vector of coefficients

## Multiple Regression - Intercept

With the intercept, the model is written as:
$$
\begin{pmatrix} y_1 \\ \vdots \\ y_n \end{pmatrix} 
= \begin{pmatrix} 1 & x_{11} & \dotsc & x_{1p}\\ \vdots \\ 1 &  x_{n1} & \dotsc & x_{np} \end{pmatrix} 
\begin{pmatrix} \beta_0 \\ \beta_1 \\ \vdots \\ \beta_p \end{pmatrix} 
+ \begin{pmatrix} \epsilon_1 \\ \vdots \\ \epsilon_n \end{pmatrix} 
$$
Re-numbering, we get: 
$$
\begin{pmatrix} y_1 \\ \vdots \\ y_n \end{pmatrix} 
= \begin{pmatrix} x'_{11} & x'_{12} & \dotsc & x'_{1(p+1)}\\ \vdots \\ x'_{n1} &  x'_{n2} & \dotsc & x'_{n(p+1)} \end{pmatrix} 
\begin{pmatrix} \beta'_1 \\ \beta'_2 \\ \vdots \\ \beta'_{p+1} \end{pmatrix} 
+ \begin{pmatrix} \epsilon_1 \\ \vdots \\ \epsilon_n \end{pmatrix} 
$$
So that we see the intercept $\mathbb{1} = (1, \dotsc, 1)^T$ as a predictor.
<!-- $$ -->
<!-- x'_{ij} = -->
<!-- \begin{cases} -->
<!-- 1 & \text{if } j = 1 \\ -->
<!-- x_{ij-1} &\text{if } j > 1 -->
<!-- \end{cases} -->
<!-- \quad \text{and} \quad -->
<!-- \beta'_{j} = \beta_{j-1} -->
<!-- \quad \forall 1 \leq j \leq p+1 -->
<!-- $$ -->

<!-- $$ -->
<!-- \begin{pmatrix} y_1 \\ \vdots \\ y_n \end{pmatrix}  -->
<!-- = \beta_1' \begin{pmatrix} x'_{11} \\ \vdots \\ x'_{11} \end{pmatrix}  -->
<!-- + \beta_2' \begin{pmatrix} x'_{12} \\ \vdots \\ x'_{n2} \end{pmatrix}  -->
<!-- + \dotsb -->
<!-- + \beta_p' \begin{pmatrix} x'_{1p+1} \\ \vdots \\ x'_{np+1} \end{pmatrix}  -->
<!-- + \begin{pmatrix} \epsilon_1 \\ \vdots \\ \epsilon_n \end{pmatrix}  -->
<!-- $$ -->

## Multiple Regression - Intercept

Without loss of generality, we write:

$$
\begin{pmatrix} y_1 \\ \vdots \\ y_n \end{pmatrix} 
= \begin{pmatrix} x_{11} & \dotsc & x_{1p}\\ \vdots \\ x_{n1} & \dotsc & x_{np} \end{pmatrix} 
\begin{pmatrix} \beta_1 \\ \vdots \\ \beta_p \end{pmatrix} 
+ \begin{pmatrix} \epsilon_1 \\ \vdots \\ \epsilon_n \end{pmatrix} 
$$

We use this convention in the rest of the text.

<!-- ************************************************************************ -->
# Model
<!-- ************************************************************************ -->

## Multiple Regression - Model

### Model:
$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
$$

* $\mathbf{y}$ **random** vector of $n$ responses
* $\mathbf{X}$ **non random** $n\times p$ matrix of predictors
* $\boldsymbol{\epsilon}$ **random** vector of $n$ errors
* $\boldsymbol{\beta}$ **non random, unknown** vector of $p$ coefficients

### Assumptions:
  * (H1) $rg(\mathbf{X}) = p$
  * (H2) $\mathbb{E}[\boldsymbol{\epsilon}] = 0$ and $\mathbb{Var}[\boldsymbol{\epsilon}] = \sigma^2 \mathbf{I}_n$ 

## Notations

The column and row vectors of $\mathbf{X}$ are written as:

$$
\mathbf{X} = 
\begin{pmatrix} \mathbf{x}_{1} & \cdots & \mathbf{x}_{p} \end{pmatrix} 
=
\begin{pmatrix} \mathbf{x}^{1} \\ \vdots \\ \mathbf{x}^{n} \end{pmatrix} 
$$

For $1\leq i \leq n$:
$$
y_i = [\mathbf{X}\boldsymbol{\beta}]_i + \epsilon_i = \mathbf{x}^{i} \boldsymbol{\beta} + \epsilon_i
$$

And:
$$
\mathbf{y} = \sum_{k = 1}^p \beta_k \mathbf{x}_k + \boldsymbol{\epsilon}
$$

## Degrees of freedom

For the model
$$
\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
\quad
\text{with}
\quad
rg(\mathbf{X}) = p
$$

there are **$n-p$ degrees of freedom**.

## Degrees of freedom - simple regression

In a simple regression, we have:
$$
\mathbf{y} 
= \beta_0 \mathbb{1} + \beta_1 \mathbf{x} + \boldsymbol{\epsilon}
= \begin{pmatrix} 1 & x_{1}\\ \vdots & \vdots \\  1 & x_{n} \end{pmatrix} 
\begin{pmatrix} \beta_0 \\ \beta_1 \end{pmatrix} 
+ \begin{pmatrix} \epsilon_1 \\ \vdots \\ \epsilon_n \end{pmatrix} 
$$

* There are $n - 2$ degrees of freedom.

* But only *one* "explanatory variables" $\mathbf{x}$.

## Degrees of freedom - with intercept

If we write:
$$
\begin{pmatrix} y_1 \\ \vdots \\ y_n \end{pmatrix} 
= \begin{pmatrix} 1 & x_{11} & \dotsc & x_{1p}\\ \vdots \\ 1 &  x_{n1} & \dotsc & x_{np} \end{pmatrix} 
\begin{pmatrix} \beta_0 \\ \beta_1 \\ \vdots \\ \beta_p \end{pmatrix} 
+ \begin{pmatrix} \epsilon_1 \\ \vdots \\ \epsilon_n \end{pmatrix} 
$$

* There are $n - (p + 1) = n - p - 1$ degrees of freedom.

* But only $p$ "explanatory variables" $\mathbf{x}_1, \cdots, \mathbf{x}_p$.

## Degrees of freedom - general case

If we write:
$$
\begin{pmatrix} y_1 \\ \vdots \\ y_n \end{pmatrix} 
= \begin{pmatrix} x_{11} & \dotsc & x_{1p}\\ \vdots \\  x_{n1} & \dotsc & x_{np} \end{pmatrix} 
\begin{pmatrix} \beta_1 \\ \vdots \\ \beta_p \end{pmatrix} 
+ \begin{pmatrix} \epsilon_1 \\ \vdots \\ \epsilon_n \end{pmatrix} 
$$

* There are $n - p$ degrees of freedom.

* If $\mathbf{x}_1 = \mathbb{1}$ is the intercept,  
then there are $p - 1$ "explanatory variables" $\mathbf{x}_2, \cdots, \mathbf{x}_p$.

<!-- ************************************************************************ -->
# Reminders: Orthogonal Projection
<!-- ************************************************************************ -->

## Reminder - Euclidean Geometry 1/2 {.build}

Let $\mathbf{x} = (x_1, \dotsc, x_n)^T$ and $\mathbf{y} = (y_1, \dotsc, y_n)^T$ be two vectors of $\mathbb{R}^n$.

The squared euclidean **norm** of $\mathbf{x}$ is given by:
$$
\|\mathbf{x}\|^2 = \sum_{i = 1}^n x_i^2 = \mathbf{x}^T\mathbf{x}
$$

The euclidean **scalar product** between $\mathbf{x}$ and $\mathbf{y}$ is given by:
$$
\langle \mathbf{x} ; \mathbf{y}\rangle = \sum_{i = 1}^n x_iy_i = \mathbf{x}^T\mathbf{y} = \mathbf{y}^T\mathbf{x}
$$

## Reminder - Euclidean Geometry 2/2 

We have the following formulas:

$$
\|\mathbf{x} + \mathbf{y}\|^2 = \|\mathbf{x}\|^2 + \|\mathbf{y}\|^2 + 2\langle \mathbf{x} ; \mathbf{y}\rangle
$$

$$
\|\mathbf{x} - \mathbf{y}\|^2 = \|\mathbf{x}\|^2 + \|\mathbf{y}\|^2 - 2\langle \mathbf{x} ; \mathbf{y}\rangle
$$

$$
\langle \mathbf{x} ; \mathbf{y}\rangle 
= \frac{1}{4} \left(\|\mathbf{x} + \mathbf{y}\|^2 - \|\mathbf{x} - \mathbf{y}\|^2 \right)
$$

$$
\langle \mathbf{x} ; \mathbf{y}\rangle 
\leq \|\mathbf{x}\|\|\mathbf{y}\| \qquad \text{(Cauchy-Schwarz)}
$$

$$
\|\mathbf{x} + \mathbf{y}\|
\leq \|\mathbf{x}\| + \|\mathbf{y}\| \qquad \text{(Triangle Inequality)}
$$

## Reminder: Orthogonal Projection - 1/3 {.build}

Let $\mathbf{y} \in \mathbb{R}^n$,  
$(\mathbf{x}_1, \dotsc, \mathbf{x}_p) \in \mathbb{R}^n \times \dotsb \times \mathbb{R}^n$
$p$ linearly independent vectors,  
and $\mathcal{M}(\mathbf{X}) = \text{span}\{\mathbf{x}_1, \dotsc, \mathbf{x}_p\}$.

The *orthogonal projection* of $\mathbf{y}$ on $\mathcal{M}(\mathbf{X})$
is defined as the unique vector $\text{Proj}_{\mathcal{M}(\mathbf{X})}(\mathbf{y}) \in \mathcal{M}(\mathbf{X})$ such that:
$$
\text{Proj}_{\mathcal{M}(\mathbf{X})}(\mathbf{y})
 = \underset{\tilde{\mathbf{y}} = \mathbf{X}\boldsymbol{\beta}\\ \boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\|\mathbf{y} - \tilde{\mathbf{y}} \|^2.
$$

It is also the unique vector such that $\mathbf{y} - \text{Proj}_{\mathcal{M}(\mathbf{X})}(\mathbf{y})$
is orthogonal to $\mathcal{M}(\mathbf{X})$, 
i.e. orthogonal to $\mathbf{x}_k$ for all $1 \leq k \leq p$.

There is a unique matrix $\mathbf{P}^\mathbf{X}$ such that 
$\text{Proj}_{\mathcal{M}(\mathbf{X})}(\mathbf{y}) = \mathbf{P}^\mathbf{X} \mathbf{y}$.

## Reminder: Orthogonal Projection - 2/4

```{r proj0, echo=FALSE, fig.height=5, fig.width=5, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
yv <- c(0.8, 1.0)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
yhatv <- c(0.8, 0.3)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "red", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(y)), col = "red", pos = 2)
# X
arrows(x0 = or[1], y0 = or[2], x1 = xv[1], y1 = xv[2],
       length = 0.1, col = "black", lwd = 2)
text(xv[1], xv[2], labels = expression(bold(X)[1]), col = "black", pos = 1)
# one
arrows(x0 = or[1], y0 = or[2], x1 = onev[1], y1 = onev[2],
       length = 0.1, col = "black", lwd = 2)
text(onev[1], onev[2], labels = expression(bold(X)[2]), col = "black", pos = 2)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "blue", lwd = 2)
text(yhatv[1], yhatv[2] + 0.05,
     labels = expression(paste(Proj[bold(X)], bold(y))),
     col = "blue", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "black", lwd = 1, lty = 2)
st <- 0.03
segments(x0 = yhatv[1], y0 = yhatv[2] + st,
         x1 = yhatv[1] - st*5/sqrt(26), y1 = yhatv[2] + st - st/sqrt(26),
         col = "black", lwd = 1)
segments(x0 = yhatv[1] - st*5/sqrt(26), y0 = yhatv[2] - st/sqrt(26),
         x1 = yhatv[1] - st*5/sqrt(26), y1 = yhatv[2] + st - st/sqrt(26),
         col = "black", lwd = 1)
# hat epsilon
text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
     labels = expression(bold(y) - paste(Proj[bold(X)], bold(y))), col = "blue", pos = 4)
```

## Reminder: Orthogonal Projection - 3/4

**Definition**:  

* An $n$ square matrix $\mathbf{P}$ is a *projection* matrix if
$\mathbf{P}^2 = \mathbf{P}$.

>* If $\mathbf{P}^2 = \mathbf{P}$ and $\mathbf{P}$ is symmetric,  
then $\mathbf{P}$ is an *orthogonal projection* matrix.

>* For any $\mathbf{x} \in \mathbb{R}^n$, 
$\mathbf{P}\mathbf{x}$ is the projection on $\text{Im}(\mathbf{P})$,
parallel to $\text{Ker}(\mathbf{P})$.

>* If $\mathbf{P}$ is a projection matrix, then $\mathbf{P}$ is an orthogonal projection matrix
iff for any
$\mathbf{x} \in \mathbb{R}^n$, we have:
$$
\mathbf{x} = \mathbf{P}\mathbf{x} + (\mathbf{I}_n - \mathbf{P})\mathbf{x}
$$
with
$\mathbf{P}\mathbf{x}$ and $(\mathbf{I}_n - \mathbf{P})\mathbf{x}$ orthogonal.

## Reminder: Matrix of Projection - 4/4 {.build}

**Definition**:  
An $n \times n$ matrix is said to be *orthogonal* if
$\mathbf{U}\mathbf{U}^T = \mathbf{I}_n$.  
The columns of $\mathbf{U}$ are then an orthogonal basis of $\mathbb{R}^n$.

**Property**:  
If $\mathbf{P}$ is an orthogonal projection matrix,
then there exists one orthogonal matrix $\mathbf{U}$
and a diagonal matrix $\mathbf{\Delta}$ such that:
$$
\mathbf{P} = \mathbf{U}\mathbf{\Delta}\mathbf{U}^T
$$
with:
$$
\mathbf{\Delta} =
\begin{pmatrix}
\mathbf{I}_p & \mathbf{0} \\
\mathbf{0} & \mathbf{0} \\
\end{pmatrix}
\quad \text{and} \quad
p = \dim(\text{Im}(\mathbf{P})) = \text{rg}(\mathbf{P})
$$

<!-- ************************************************************************ -->
# Ordinary Least Squares (OLS)
<!-- ************************************************************************ -->

## OLS - Definition

The OLS estimator $\hat{\boldsymbol{\beta}}$ is given by:

$$ 
\hat{\boldsymbol{\beta}} 
= \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\left\{
\sum_{i = 1}^n \left(y_i - \sum_{j = 1}^p \beta_j x_{ij}\right)^2
\right\}
$$
$~$

**Goal**: Minimize the squared errors between: 

  * the prediction of the model $\beta_1 x_{i1} + \dotsb + \beta_p x_{ip}$ and
  * the actual observed value $y_i$.
  
## OLS - Projection

$$ 
\begin{aligned}
\hat{\boldsymbol{\beta}} 
 & = \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\left\{
\sum_{i = 1}^n \left(y_i - \sum_{j = 1}^p \beta_j x_{ij}\right)^2
\right\}\\
 & = \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\left\{
\sum_{i = 1}^n \left(y_i - (\mathbf{X}\boldsymbol{\beta})_i\right)^2
\right\}\\
 & = \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\|\mathbf{y} - \mathbf{X}\boldsymbol{\beta} \|^2
\end{aligned}
$$

## OLS - Projection

$$ 
\hat{\boldsymbol{\beta}} 
 = \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\left\{
\sum_{i = 1}^n \left(y_i - \sum_{j = 1}^p \beta_j x_{ij}\right)^2
\right\}
 = \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\|\mathbf{y} - \mathbf{X}\boldsymbol{\beta} \|^2
$$

hence

$$ 
\hat{\mathbf{y}} = \mathbf{X}\hat{\boldsymbol{\beta}} 
 = \underset{\tilde{\mathbf{y}} = \mathbf{X}\boldsymbol{\beta}\\ \boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\|\mathbf{y} - \tilde{\mathbf{y}} \|^2 
 = \text{Proj}_{\mathcal{M}(\mathbf{X})}(\mathbf{y})
 = \mathbf{P}^{\mathbf{X}}\mathbf{y} 
$$

$\hat{\mathbf{y}}$ is the **orthogonal projection** of $\mathbf{y}$ on 
$$\mathcal{M}(\mathbf{X}) = \text{span}\{\mathbf{x}_1, \dotsc, \mathbf{x}_p\}$$
the plan spanned by the columns of $\mathbf{X}$.

## Geometrical Interpretation

```{r proj, echo=FALSE, fig.height=5, fig.width=5, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
yv <- c(0.8, 1.0)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
yhatv <- c(0.8, 0.3)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "red", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(Y)), col = "red", pos = 2)
# X
arrows(x0 = or[1], y0 = or[2], x1 = xv[1], y1 = xv[2],
       length = 0.1, col = "black", lwd = 2)
text(xv[1], xv[2], labels = expression(bold(X)[1]), col = "black", pos = 1)
# one
arrows(x0 = or[1], y0 = or[2], x1 = onev[1], y1 = onev[2],
       length = 0.1, col = "black", lwd = 2)
text(onev[1], onev[2], labels = expression(bold(X)[2]), col = "black", pos = 2)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "blue", lwd = 2)
text(yhatv[1], yhatv[2] + 0.05,
     labels = expression(hat(bold(Y)) == paste(bold(X), hat(bold(beta)))),
     col = "blue", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "black", lwd = 1, lty = 2)
st <- 0.03
segments(x0 = yhatv[1], y0 = yhatv[2] + st,
         x1 = yhatv[1] - st*5/sqrt(26), y1 = yhatv[2] + st - st/sqrt(26),
         col = "black", lwd = 1)
segments(x0 = yhatv[1] - st*5/sqrt(26), y0 = yhatv[2] - st/sqrt(26),
         x1 = yhatv[1] - st*5/sqrt(26), y1 = yhatv[2] + st - st/sqrt(26),
         col = "black", lwd = 1)
# hat epsilon
text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
     labels = expression(hat(bold(epsilon))), col = "blue", pos = 4)
# beta x
segments(x0 = betaonev[1], y0 = betaonev[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "blue", lwd = 1)
text((betaonev[1] + yhatv[1])/2, (betaonev[2] + yhatv[2])/2,
     labels = expression(paste(hat(beta)[1], bold(X)[1])), col = "blue", pos = 3)
# beta one
segments(x0 = betaxv[1], y0 = betaxv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "blue", lwd = 1)
text((betaxv[1] + yhatv[1])/2, (betaxv[2] + yhatv[2])/2,
     labels = expression(paste(hat(beta)[2], bold(X)[2])), col = "blue", pos = 4)
segments(x0 = or[1], y0 = or[2], x1 = betaxv[1], y1 = betaxv[2],
         col = "black", lwd = 1, lty = 2)
```

## OLS - Projection

$$ 
\hat{\boldsymbol{\beta}} 
 = \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\left\{
\sum_{i = 1}^n \left(y_i - \sum_{j = 1}^p \beta_j x_{ij}\right)^2
\right\}
 = \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\|\mathbf{y} - \mathbf{X}\boldsymbol{\beta} \|^2
$$


$$ 
\hat{\mathbf{y}} = \mathbf{X}\hat{\boldsymbol{\beta}} 
 = \underset{\tilde{\mathbf{y}} = \mathbf{X}\boldsymbol{\beta}\\ \boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\|\mathbf{y} - \tilde{\mathbf{y}} \|^2 
 = \text{Proj}_{\mathcal{M}(\mathbf{X})}(\mathbf{y})
 = \mathbf{P}^{\mathbf{X}}\mathbf{y} 
$$

$\mathbf{P}^{\mathbf{X}}$ is the *orthogonal projection matrix* on $\mathcal{M}(\mathbf{X})$.

$$
\mathbf{y} = \mathbf{P}^{\mathbf{X}}\mathbf{y} + (\mathbf{I}_n - \mathbf{P}^{\mathbf{X}})\mathbf{y}
$$

with
$\mathbf{P}^{\mathbf{X}}\mathbf{y}$ and $(\mathbf{I}_n - \mathbf{P}^{\mathbf{X}})\mathbf{y}$ orthogonal.

<!-- ************************************************************************ -->
# OLS - Expression
<!-- ************************************************************************ -->

## OLS - Expression {.build}

$$ 
\hat{\boldsymbol{\beta}} 
 = \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{argmin}} 
\|\mathbf{y} - \mathbf{X}\boldsymbol{\beta} \|^2
\quad \text{and} \quad
\hat{\mathbf{y}} = \mathbf{X}\hat{\boldsymbol{\beta}} 
 = \mathbf{P}^{\mathbf{X}}\mathbf{y} 
$$

The OLS estimator is given by:
$$ 
\hat{\boldsymbol{\beta}} = (\mathbf{X}^{T}\mathbf{X})^{-1}\mathbf{X}^{T}\mathbf{y}
$$

And the matrix of orthogonal projection on $\mathcal{M}(\mathbf{X})$ is:
$$ 
\mathbf{P}^{\mathbf{X}} = \mathbf{X}(\mathbf{X}^{T}\mathbf{X})^{-1}\mathbf{X}^{T}
$$

## OLS - Proof (Geometrical) - 1/2 {.build}

$\hat{\mathbf{y}} = \mathbf{X}\hat{\boldsymbol{\beta}}$ is the orthogonal projection
of $\mathbf{y}$ on $\mathcal{M}(\mathbf{X})$.

It is the only vector such that $\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}}$ is orthogonal to
$\mathcal{M}(\mathbf{X})$.

I.e. $\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}}$ is orthogonal to all $\mathbf{x}_k$:
$$
\langle \mathbf{x}_k,  \mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}} \rangle
= \mathbf{x}_k^T (\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}}) = 0
\qquad 
1 \leq k \leq p
$$

$$
i.e. \quad 
(\mathbf{x}_1 \dotsc \mathbf{x}_p)^T (\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}})
= 
\begin{pmatrix}
\mathbf{x}_1^T (\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}})\\
\vdots\\
\mathbf{x}_p^T (\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}})
\end{pmatrix}
=
\mathbf{0}
$$

$$
and \quad 
\mathbf{X}^T (\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}}) = 0.
$$

## OLS - Proof (Geometrical) - 2/2 {.build}

$$
\mathbf{X}^T (\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}}) = 0.
$$

$$
\mathbf{X}^T\mathbf{X}\hat{\boldsymbol{\beta}} = \mathbf{X}^T\mathbf{y}
$$

As $rg(\mathbf{X}) = p$, $\mathbf{X}^T\mathbf{X}$ is invertible, and:

$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}
$$

Then:
$$
\hat{\mathbf{y}} 
= \mathbf{P}^{\mathbf{X}}\mathbf{y} 
= \mathbf{X}\hat{\boldsymbol{\beta}} 
= \mathbf{X} (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}
$$

As this equality is true for any $\mathbf{y}$, we get:

$$
\mathbf{P}^{\mathbf{X}}
= \mathbf{X} (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T.
$$


## OLS - Proof (Analytical) {.build}

The minimum of quadratic function $S$ is obtained at the point where the gradient is zero:

$$ 
S(\boldsymbol{\beta})
= \|\mathbf{y} - \mathbf{X}\boldsymbol{\beta} \|^2
= \boldsymbol{\beta}^T\mathbf{X}^T\mathbf{X}\boldsymbol{\beta}
- 2 \boldsymbol{\beta}^T\mathbf{X}^T\mathbf{y}
+ \mathbf{y}^T \mathbf{y}
$$

$$ 
\nabla S(\hat{\boldsymbol{\beta}}) = \cdots = 0
$$

$$ 
\nabla S(\hat{\boldsymbol{\beta}})
= 2 \mathbf{X}^T\mathbf{X}\hat{\boldsymbol{\beta}}
- 2 \mathbf{X}^T\mathbf{y}
= 0
$$

$$ 
\mathbf{X}^T\mathbf{X}\hat{\boldsymbol{\beta}} = \mathbf{X}^T\mathbf{y}
$$

Same conclusion as before.

<!-- ************************************************************************ -->
# Reminder - Expectation and Variance of random vectors
<!-- ************************************************************************ -->

## Expectation and Variance 

Let $\mathbf{z}$ be a random vector of dimension $n$.

* The *expectation* of $\mathbf{z}$ is an $n$ vector:
$$
\mathbb{E}[\mathbf{z}] = (\mathbb{E}[z_1], \cdots, \mathbb{E}[z_n])^T
$$

* The *variance* of $\mathbf{z}$ is an $n \times n$ **matrix**:
$$
\begin{aligned}
\mathbb{Var}[\mathbf{z}] 
&= \left[\mathbb{Cov}[z_i, z_j]\right]_{1 \leq i, j \leq n}\\
&= \mathbb{E}\left[(\mathbf{z} - \mathbb{E}[\mathbf{z}])(\mathbf{z} - \mathbb{E}[\mathbf{z}])^T\right]\\
&= \mathbb{E}[\mathbf{z}\mathbf{z}^T] - \mathbb{E}[\mathbf{z}]\mathbb{E}[\mathbf{z}]^T
\end{aligned}
$$

## Expectation and Variance - Properties

Let $\mathbf{A}$ be a $m \times n$ deterministic matrix,
and $\mathbf{b}$ be a $m$ vector.
Then:

$$
\mathbb{E}[\mathbf{A}\mathbf{z} + \mathbf{b}] = \mathbf{A}\mathbb{E}[\mathbf{z}] + \mathbf{b}
$$

$$
\mathbb{Var}[\mathbf{A}\mathbf{z} + \mathbf{b}] = \mathbf{A}\mathbb{Var}[\mathbf{z}]\mathbf{A}^T.
$$

**Attention**: Transpose is on the *right* (variance is a matrix).

<!-- ************************************************************************ -->
# The OLS Estimator is unbiased
<!-- ************************************************************************ -->

## OLS estimator is unbiased
The OLS estimator
$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}
$$
is unbiased:
$$
\mathbb{E}[\hat{\boldsymbol{\beta}}] = \boldsymbol{\beta}.
$$

## OLS estimator is unbiased - Proof - 1/3

$$
\mathbb{E}[\hat{\boldsymbol{\beta}}]
= \mathbb{E}[(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}]
= \cdots
$$

## OLS estimator is unbiased - Proof - 2/3

$$
\mathbb{E}[\hat{\boldsymbol{\beta}}]
= \mathbb{E}[(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}]
= (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbb{E}[\mathbf{y}]
$$

And:
$$
\mathbb{E}[\mathbf{y}] 
= \mathbb{E}[\mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}] 
= \cdots
$$

## OLS estimator is unbiased - Proof - 3/3

$$
\mathbb{E}[\hat{\boldsymbol{\beta}}]
= \mathbb{E}[(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}]
= (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbb{E}[\mathbf{y}]
$$

And:
$$
\mathbb{E}[\mathbf{y}] 
= \mathbb{E}[\mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}] 
= \mathbf{X}\boldsymbol{\beta} + \mathbb{E}[\boldsymbol{\epsilon}] 
= \mathbf{X}\boldsymbol{\beta}
$$

Hence:
$$
\mathbb{E}[\hat{\boldsymbol{\beta}}]
= (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{X}\boldsymbol{\beta}
= \boldsymbol{\beta}.
$$

<!-- ************************************************************************ -->
# Variance of the OLS Estimator
<!-- ************************************************************************ -->

## Variance of the OLS Estimator

The OLS estimator
$$
\hat{\boldsymbol{\beta}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}
$$
has variance:
$$
\mathbb{Var}[\hat{\boldsymbol{\beta}}] = \sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}.
$$

## Variance of OLS Estimator - Proof - 1/3

$$
\mathbb{Var}[\hat{\boldsymbol{\beta}}]
= \mathbb{Var}[(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}]
= \cdots
$$

## Variance of OLS Estimator - Proof - 2/3

$$
\begin{aligned}
\mathbb{Var}[\hat{\boldsymbol{\beta}}]
& = \mathbb{Var}[(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}]\\
& = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T \mathbb{Var}[\mathbf{y}] \left[(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\right]^T \\
& = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T \mathbb{Var}[\mathbf{y}] \mathbf{X} (\mathbf{X}^T\mathbf{X})^{-1}
\end{aligned}
$$

And:
$$
\mathbb{Var}[\mathbf{y}] 
= \mathbb{Var}[\mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}] 
= \cdots
$$

## Variance of OLS Estimator - Proof - 3/3

$$
\begin{aligned}
\mathbb{Var}[\hat{\boldsymbol{\beta}}]
& = \mathbb{Var}[(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}]\\
& = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T \mathbb{Var}[\mathbf{y}] \left[(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\right]^T \\
& = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T \mathbb{Var}[\mathbf{y}] \mathbf{X} (\mathbf{X}^T\mathbf{X})^{-1}
\end{aligned}
$$

And:
$$
\begin{aligned}
\mathbb{Var}[\mathbf{y}] 
= \mathbb{Var}[\mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}] 
= \mathbb{Var}[\boldsymbol{\epsilon}]
= \sigma^2 \mathbf{I}_n
\end{aligned}
$$

Hence:
$$
\begin{aligned}
\mathbb{Var}[\hat{\boldsymbol{\beta}}]
& = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T \left[\sigma^2 \mathbf{I}_n\right] \mathbf{X} (\mathbf{X}^T\mathbf{X})^{-1}\\
&= \sigma^2 (\mathbf{X}^T\mathbf{X})^{-1} (\mathbf{X}^T\mathbf{X}) (\mathbf{X}^T\mathbf{X})^{-1}\\
&= \sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}.
\end{aligned}
$$

<!-- ************************************************************************ -->
# Gauss - Markov Theorem
<!-- ************************************************************************ -->

## Partial order on Symmetric matrices

**Definition** Let $\mathbf{S}_1$ and $\mathbf{S}_2$ two real $n\times n$ symmetric matrices.  
We say that $\mathbf{S}_1$ is smaller than $\mathbf{S}_2$, and write
$$\mathbf{S}_1 \leq \mathbf{S}_2$$
iif $\mathbf{S}_2 - \mathbf{S}_1$ is
positive, semi-definite, i.e.:
$$
\mathbf{z}^T(\mathbf{S}_2 - \mathbf{S}_1)\mathbf{z} \geq 0 \qquad \forall \mathbf{z} \in \mathbb{R}^n,
$$

or, equivalently:
$$
\mathbf{z}^T \mathbf{S}_1\mathbf{z} \leq \mathbf{z}^T \mathbf{S}_2\mathbf{z} \qquad \forall \mathbf{z} \in \mathbb{R}^n.
$$

## Gauss-Markov theorem {.build}

The OLS estimator $\hat{\boldsymbol{\beta}}$ is the **BLUE**  
(Best Linear Unbiased Estimator):  
it is the linear unbiased estimator with the minimal variance.

**Remark**:
The OLS estimator
$$
\hat{\boldsymbol{\beta}} 
= (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}
= \mathbf{A}\mathbf{y}
$$
is indeed *linear*, and it is unbiased.

## Reminder - Variance of the Sum

Let $\mathbf{u}$ and $\mathbf{v}$ two random vectors of size $n$.
Then:
$$
\mathbb{Var}[\mathbf{u} + \mathbf{v}]
=
\mathbb{Var}[\mathbf{u}] + \mathbb{Var}[\mathbf{v}]
+ \mathbb{Cov}[\mathbf{u}; \mathbf{v}]
+ \mathbb{Cov}[\mathbf{v}; \mathbf{u}]
$$

where:
$$
\mathbb{Cov}[\mathbf{u}; \mathbf{v}] 
= 
\mathbb{E}[\mathbf{u}\mathbf{v}^T] - \mathbb{E}[\mathbf{u}]\mathbb{E}[\mathbf{v}]^T
=
\mathbb{Cov}[\mathbf{v}; \mathbf{u}]^T 
$$

## Gauss-Markov - Proof - 1/6

Let $\bar{\boldsymbol{\beta}}$ a linear unbiased estimator of $\boldsymbol{\beta}$.  
Let's show that $\mathbb{Var}[\hat{\boldsymbol{\beta}}] \leq \mathbb{Var}[\bar{\boldsymbol{\beta}}]$.

$$
\mathbb{Var}[\bar{\boldsymbol{\beta}}]
=
\mathbb{Var}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}} + \hat{\boldsymbol{\beta}}]
= \cdots
$$

## Gauss-Markov - Proof - 2/6

Let $\bar{\boldsymbol{\beta}}$ a linear unbiased estimator of $\boldsymbol{\beta}$.  
Let's show that $\mathbb{Var}[\hat{\boldsymbol{\beta}}] \leq \mathbb{Var}[\bar{\boldsymbol{\beta}}]$.

$$
\begin{aligned}
\mathbb{Var}[\bar{\boldsymbol{\beta}}]
& = \mathbb{Var}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}} + \hat{\boldsymbol{\beta}}]\\
& = 
\mathbb{Var}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}]
+ \mathbb{Var}[\hat{\boldsymbol{\beta}}]\\
& \qquad + \mathbb{Cov}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}] 
+ \mathbb{Cov}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}]^T
\end{aligned}
$$

i.e.
$$
\begin{aligned}
\mathbb{Var}[\bar{\boldsymbol{\beta}}] - \mathbb{Var}[\hat{\boldsymbol{\beta}}]
& = 
\mathbb{Var}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}] \\
& \qquad + \mathbb{Cov}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}] 
+ \mathbb{Cov}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}]^T
\end{aligned}
$$

We need to prove: $\mathbb{Var}[\bar{\boldsymbol{\beta}}] - \mathbb{Var}[\hat{\boldsymbol{\beta}}]$
is positive, semi-definite.

## Gauss-Markov - Proof - 3/6

$$
\begin{aligned}
\mathbb{Var}[\bar{\boldsymbol{\beta}}] - \mathbb{Var}[\hat{\boldsymbol{\beta}}]
& = 
\mathbb{Var}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}] \\
& \qquad + \mathbb{Cov}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}] 
+ \mathbb{Cov}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}]^T
\end{aligned}
$$

We need to prove: $\mathbb{Var}[\bar{\boldsymbol{\beta}}] - \mathbb{Var}[\hat{\boldsymbol{\beta}}]$
is positive, semi-definite.

As $\mathbb{Var}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}]$ is positive, semi-definite,
we just need to prove:
$$
\mathbb{Cov}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}] = 0.
$$

## Gauss-Markov - Proof - 3/6 {.build}

$\bar{\boldsymbol{\beta}}$ is a linear unbiased estimator of $\boldsymbol{\beta}$.  
Hence:
$$
\bar{\boldsymbol{\beta}} = \mathbf{B} \mathbf{y}
$$

and:
$$
\begin{aligned}
\mathbb{E}[\bar{\boldsymbol{\beta}}] 
&= \mathbb{E}[\mathbf{B} \mathbf{y}] 
= \mathbf{B} \mathbb{E}[\mathbf{y}] 
= \mathbf{B} \mathbb{E}[\mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}] 
= \mathbf{B} \mathbf{X}\boldsymbol{\beta}\\
&= \boldsymbol{\beta}
\end{aligned}
$$
for all $\boldsymbol{\beta}$, hence:
$$
\mathbf{B} \mathbf{X} = \mathbf{I}_n.
$$

## Gauss-Markov - Proof - 4/6 

Let's show that:
$$
\mathbb{Cov}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}] = 0.
$$

$$
\begin{aligned}
\mathbb{Cov}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}]
& = \mathbb{Cov}[\bar{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}] - \mathbb{Var}[\hat{\boldsymbol{\beta}}]\\
& = \cdots
\end{aligned}
$$

## Gauss-Markov - Proof - 5/6 {.build}

$$
\begin{aligned}
\mathbb{Cov}[\bar{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}]
& = \mathbb{E}[\bar{\boldsymbol{\beta}} \hat{\boldsymbol{\beta}}^T] 
- \mathbb{E}[\bar{\boldsymbol{\beta}}]\mathbb{E}[\hat{\boldsymbol{\beta}}]^T \\
& = \mathbb{E}[\mathbf{B} \mathbf{y} [(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}]^T] 
- \boldsymbol{\beta}\boldsymbol{\beta}^T \\
& = \mathbf{B} \mathbb{E}[\mathbf{y}\mathbf{y}^T]\mathbf{X}(\mathbf{X}^T\mathbf{X})^{-1}
- \boldsymbol{\beta}\boldsymbol{\beta}^T \\
\end{aligned}
$$


and:
$$
\mathbb{E}[\mathbf{y}\mathbf{y}^T]
=
\mathbb{Var}[\mathbf{y}] + 
\mathbb{E}[\mathbf{y}]\mathbb{E}[\mathbf{y}]^T
=
\sigma^2 \mathbf{I}_n + \mathbf{X}\boldsymbol{\beta}\boldsymbol{\beta}^T\mathbf{X}^T
$$


hence:
$$
\begin{aligned}
\mathbb{Cov}[\bar{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}]
 = \sigma^2 \mathbf{B}\mathbf{X}(\mathbf{X}^T\mathbf{X})^{-1}
+ \mathbf{B}\mathbf{X}\boldsymbol{\beta}\boldsymbol{\beta}^T\mathbf{X}^T \mathbf{X}(\mathbf{X}^T\mathbf{X})^{-1}
- \boldsymbol{\beta}\boldsymbol{\beta}^T \\
\end{aligned}
$$


as $\mathbf{B} \mathbf{X} = \mathbf{I}_n$:
$$
\begin{aligned}
\mathbb{Cov}[\bar{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}]
 = \sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}
+ \boldsymbol{\beta}\boldsymbol{\beta}^T
- \boldsymbol{\beta}\boldsymbol{\beta}^T
 = \sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}
\end{aligned}
$$

## Gauss-Markov - Proof - 6/6 

Finally:
$$
\begin{aligned}
\mathbb{Cov}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}]
& = \mathbb{Cov}[\bar{\boldsymbol{\beta}}; \hat{\boldsymbol{\beta}}] - \mathbb{Var}[\hat{\boldsymbol{\beta}}]\\
& = \sigma^2 (\mathbf{X}^T\mathbf{X})^{-1} - \sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}\\
&= 0.
\end{aligned}
$$

and:
$$
\mathbb{Var}[\bar{\boldsymbol{\beta}}] - \mathbb{Var}[\hat{\boldsymbol{\beta}}]
= \mathbb{Var}[\bar{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}]
$$

is positive, semi-definite, i.e.:
$$
\mathbb{Var}[\bar{\boldsymbol{\beta}}] \geq \mathbb{Var}[\hat{\boldsymbol{\beta}}].
$$
Which ends the proof.

<!-- ************************************************************************ -->
# Residuals
<!-- ************************************************************************ -->

## Geometrical Interpretation

```{r proj_res, echo=FALSE, fig.height=5, fig.width=5, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
yv <- c(0.8, 1.0)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
yhatv <- c(0.8, 0.3)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "red", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(Y)), col = "red", pos = 2)
# X
arrows(x0 = or[1], y0 = or[2], x1 = xv[1], y1 = xv[2],
       length = 0.1, col = "black", lwd = 2)
text(xv[1], xv[2], labels = expression(bold(X)[1]), col = "black", pos = 1)
# one
arrows(x0 = or[1], y0 = or[2], x1 = onev[1], y1 = onev[2],
       length = 0.1, col = "black", lwd = 2)
text(onev[1], onev[2], labels = expression(bold(X)[2]), col = "black", pos = 2)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "blue", lwd = 2)
text(yhatv[1], yhatv[2] + 0.05,
     labels = expression(hat(bold(Y)) == paste(bold(X), hat(bold(beta)))),
     col = "blue", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "black", lwd = 1, lty = 2)
st <- 0.03
segments(x0 = yhatv[1], y0 = yhatv[2] + st,
         x1 = yhatv[1] - st*5/sqrt(26), y1 = yhatv[2] + st - st/sqrt(26),
         col = "black", lwd = 1)
segments(x0 = yhatv[1] - st*5/sqrt(26), y0 = yhatv[2] - st/sqrt(26),
         x1 = yhatv[1] - st*5/sqrt(26), y1 = yhatv[2] + st - st/sqrt(26),
         col = "black", lwd = 1)
# hat epsilon
text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
     labels = expression(hat(bold(epsilon))), col = "blue", pos = 4)
# beta x
segments(x0 = betaonev[1], y0 = betaonev[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "blue", lwd = 1)
text((betaonev[1] + yhatv[1])/2, (betaonev[2] + yhatv[2])/2,
     labels = expression(paste(hat(beta)[1], bold(X)[1])), col = "blue", pos = 3)
# beta one
segments(x0 = betaxv[1], y0 = betaxv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "blue", lwd = 1)
text((betaxv[1] + yhatv[1])/2, (betaxv[2] + yhatv[2])/2,
     labels = expression(paste(hat(beta)[2], bold(X)[2])), col = "blue", pos = 4)
segments(x0 = or[1], y0 = or[2], x1 = betaxv[1], y1 = betaxv[2],
         col = "black", lwd = 1, lty = 2)
```

## Residuals - Definitions

$$
\begin{aligned}
\hat{\boldsymbol{\epsilon}}
& = \mathbf{y} - \hat{\mathbf{y}} \\
& = (\mathbf{I}_n - \mathbf{P}^{\mathbf{X}})\mathbf{y}\\
& = \mathbf{P}^{\mathbf{X}^\bot}\mathbf{y}\\
& = \mathbf{P}^{\mathbf{X}^\bot}(\mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon})\\
& = \mathbf{P}^{\mathbf{X}^\bot}\boldsymbol{\epsilon}
\end{aligned}
$$

## Residuals - Bias and Variance

$$
\begin{aligned}
\mathbb{E}[\hat{\boldsymbol{\epsilon}}] & = \mathbf{0} \\
\mathbb{Var}[\hat{\boldsymbol{\epsilon}}] & = \sigma^2 \mathbf{P}^{\mathbf{X}^\bot}
\end{aligned}
$$

## Residuals - Bias and Variance - Proof {.build}

Bias:
$$
\mathbb{E}[\hat{\boldsymbol{\epsilon}}] 
= \mathbb{E}[\mathbf{P}^{\mathbf{X}^\bot}\boldsymbol{\epsilon}] = \cdots
$$

$$
\mathbb{E}[\hat{\boldsymbol{\epsilon}}] 
= \mathbf{P}^{\mathbf{X}^\bot}\mathbb{E}[\boldsymbol{\epsilon}]
= \mathbf{0}.
$$

Variance:
$$
\mathbb{Var}[\hat{\boldsymbol{\epsilon}}] 
= \mathbb{Var}[\mathbf{P}^{\mathbf{X}^\bot}\boldsymbol{\epsilon}] = \cdots
$$


$$
\mathbb{Var}[\hat{\boldsymbol{\epsilon}}] 
= \mathbf{P}^{\mathbf{X}^\bot}\mathbb{Var}[\boldsymbol{\epsilon}][\mathbf{P}^{\mathbf{X}^\bot}]^T
= \sigma^2 \mathbf{P}^{\mathbf{X}^\bot}[\mathbf{P}^{\mathbf{X}^\bot}]^T
= \cdots
$$


$$
\mathbb{Var}[\hat{\boldsymbol{\epsilon}}] 
= \sigma^2 \mathbf{P}^{\mathbf{X}^\bot}\mathbf{P}^{\mathbf{X}^\bot}
= \sigma^2 \mathbf{P}^{\mathbf{X}^\bot}.
$$

## $\hat{\mathbf{y}}$ - Bias and Variance

$$
\begin{aligned}
\mathbb{E}[\hat{\mathbf{y}}] & = \mathbf{X}\boldsymbol{\beta} \\
\mathbb{Var}[\hat{\mathbf{y}}] & = \sigma^2 \mathbf{P}^{\mathbf{X}}
\end{aligned}
$$

## $\hat{\mathbf{y}}$ - Bias and Variance - Proof {.build}

Bias:
$$
\mathbb{E}[\hat{\mathbf{y}}] 
= \mathbb{E}[\mathbf{X}\hat{\boldsymbol{\beta}}] = \cdots
$$

$$
\mathbb{E}[\hat{\mathbf{y}}] 
= \mathbf{X}\mathbb{E}[\hat{\boldsymbol{\beta}}]
= \mathbf{X}\boldsymbol{\beta}
$$

Variance:
$$
\mathbb{Var}[\hat{\mathbf{y}}] 
= \mathbb{Var}[\mathbf{X}\hat{\boldsymbol{\beta}}] = \cdots
$$


$$
\mathbb{Var}[\hat{\mathbf{y}}] 
= \mathbf{X}\mathbb{Var}[\hat{\boldsymbol{\beta}}]\mathbf{X}^T
= \mathbf{X}[\sigma^2 (\mathbf{X}^T\mathbf{X})^{-1}]\mathbf{X}^T
= \cdots
$$


$$
\mathbb{Var}[\hat{\mathbf{y}}] 
= \sigma^2 \mathbf{X}(\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T
= \sigma^2 \mathbf{P}^{\mathbf{X}}.
$$

## Covariance of $\hat{\boldsymbol{\epsilon}}$ and $\hat{\mathbf{y}}$

$$
\mathbb{Cov}[\hat{\boldsymbol{\epsilon}}; \hat{\mathbf{y}}] = \mathbf{0}.
$$

## Covariance of $\hat{\boldsymbol{\epsilon}}$ and $\hat{\mathbf{y}}$ - Proof {.build}

$$
\mathbb{Cov}[\hat{\boldsymbol{\epsilon}}; \hat{\mathbf{y}}] 
= \mathbb{Cov}[\hat{\boldsymbol{\epsilon}}; \mathbf{y} - \hat{\boldsymbol{\epsilon}}] 
= \cdots
$$

$$
\mathbb{Cov}[\hat{\boldsymbol{\epsilon}}; \hat{\mathbf{y}}] 
= \mathbb{Cov}[\hat{\boldsymbol{\epsilon}}; \mathbf{y}] - \mathbb{Var}[\hat{\boldsymbol{\epsilon}}]
= \cdots
$$

$$
\mathbb{Cov}[\hat{\boldsymbol{\epsilon}}; \hat{\mathbf{y}}] 
= \mathbb{Cov}[\mathbf{P}^{\mathbf{X}^\bot}\mathbf{y}; \mathbf{y}] - \sigma^2\mathbf{P}^{\mathbf{X}^\bot}
= \cdots
$$

$$
\mathbb{Cov}[\hat{\boldsymbol{\epsilon}}; \hat{\mathbf{y}}] 
= \mathbf{P}^{\mathbf{X}^\bot}\mathbb{Var}[\mathbf{y}] - \sigma^2\mathbf{P}^{\mathbf{X}^\bot}
= \cdots
$$

$$
\mathbb{Cov}[\hat{\boldsymbol{\epsilon}}; \hat{\mathbf{y}}] 
= \mathbf{P}^{\mathbf{X}^\bot}[\sigma^2\mathbf{I}_n] - \sigma^2\mathbf{P}^{\mathbf{X}^\bot}
= \cdots
$$

$$
\mathbb{Cov}[\hat{\boldsymbol{\epsilon}}; \hat{\mathbf{y}}] 
= \sigma^2 \mathbf{P}^{\mathbf{X}^\bot}- \sigma^2\mathbf{P}^{\mathbf{X}^\bot}
= \mathbf{0}.
$$

<!-- ************************************************************************ -->
# Variance Estimation
<!-- ************************************************************************ -->

## Unbiased Variance Estimator

$$
\hat{\sigma}^2 = \frac{1}{n - p} \| \hat{\epsilon} \|^2 = \frac{1}{n - p} RSS
$$

is an unbiased estimator of $\sigma^2$.

**Note:** `p` parameters 
$\to$
n - `p` degrees of freedom.

## Unbiased Var Estimator - Proof - 1/2

$$
\mathbb{E}[\hat{\sigma}^2] 
= \frac{1}{n - p} \mathbb{E}[\| \hat{\epsilon} \|^2]
= \cdots
$$

**Magic trick**:
$$
\| \hat{\epsilon} \|^2
= Tr(\| \hat{\epsilon} \|^2)
= Tr(\hat{\epsilon}^T \hat{\epsilon})
= Tr(\hat{\epsilon} \hat{\epsilon}^T)
$$

## Unbiased Var Estimator - Proof - 2/2

We get:
$$
\begin{aligned}
\mathbb{E}[\| \hat{\epsilon} \|^2]
&= \mathbb{E}[Tr(\hat{\epsilon} \hat{\epsilon}^T)] \\
&= Tr(\mathbb{E}[\hat{\epsilon} \hat{\epsilon}^T]) & [\text{Trace is linear}]\\
&= Tr(\mathbb{Var}[\hat{\epsilon}]) & [\mathbb{E}[\hat{\epsilon}] = 0] \\
&= Tr(\sigma^2 \mathbf{P}^{\mathbf{X}^\bot}) \\
&= \sigma^2  Tr(\mathbf{P}^{\mathbf{X}^\bot}) \\
&= \sigma^2  (n - p)
\end{aligned}
$$

as $\mathbf{P}^{\mathbf{X}^\bot}$ is the projection matrix on a space of dimension $n-p$.

<!-- ************************************************************************ -->
# Prediction
<!-- ************************************************************************ -->

## Predict a new point

>* We fitted $\hat{\beta}_1, \dots, \hat{\beta}_p$ on $((y_1, \mathbf{x}^1), \dotsc, (y_n, \mathbf{x}^n))$

>* A new line of predictors $\mathbf{x}^{n+1}$ comes along. How can we guess $y_{n+1}$ ?

>* We use the same model:
$$
y_{n+1} = \mathbf{x}^{n+1}\boldsymbol{\beta} + \epsilon_{n+1}
$$
with
$\mathbb{E}[\epsilon_{n+1}] = 0$,
$\mathbb{Var}[\epsilon_{n+1}] = \sigma^2$
and 
$\mathbb{Cov}[\epsilon_{n+1}; \epsilon_i] = 0$.

>* We predict $y_{n+1}$ with:
$$
\hat{y}_{n+1} = \mathbf{x}^{n+1}\hat{\boldsymbol{\beta}}
$$

>* **Question**: What is the error 
$\hat{\epsilon}_{n+1} = y_{n+1} - \hat{y}_{n+1}$ ?

## Prediction Error {.build}

The prediction error $\hat{\epsilon}_{n+1} = y_{n+1} - \hat{y}_{n+1}$
is such that:
$$
\begin{aligned}
\mathbb{E}[\hat{\epsilon}_{n+1}] &= 0\\
\mathbb{Var}[\hat{\epsilon}_{n+1}] &= \sigma^2 \left(1 + \mathbf{x}^{n+1} (\mathbf{X}^T\mathbf{X})^{-1}(\mathbf{x}^{n+1})^T\right)\\
\end{aligned}
$$

**Remarks**:  

- $\mathbf{x}^{n+1}$ is a *line-vector* of dimension $1 \times p$.

- $\mathbf{X}$ is a *matrix* of dimension $n \times p$.

- $\mathbf{X}^T\mathbf{X}$ is a *matrix* of dimension $p \times p$.

- $\mathbf{x}^{n+1} (\mathbf{X}^T\mathbf{X})^{-1}(\mathbf{x}^{n+1})^T$ is a *scalar*.

## Prediction Error - Proof 1/2 {.build}

$$
\begin{aligned}
\mathbb{E}[\hat{\epsilon}_{n+1}] 
&= \mathbb{E}[y_{n+1} - \mathbf{x}^{n+1}\hat{\boldsymbol{\beta}}]\\
&= \cdots
\end{aligned}
$$

$$
\begin{aligned}
\mathbb{E}[\hat{\epsilon}_{n+1}] 
&= \mathbb{E}[y_{n+1} - \mathbf{x}^{n+1}\hat{\boldsymbol{\beta}}]\\
&= \mathbb{E}[y_{n+1}] - \mathbf{x}^{n+1} \mathbb{E}[\hat{\boldsymbol{\beta}}]\\
&= \mathbf{x}^{n+1}\boldsymbol{\beta} -  \mathbf{x}^{n+1}\boldsymbol{\beta}\\
&= 0
\end{aligned}
$$

## Prediction Error - Proof 2/2 {.build}

Because $\hat{y}_{n+1}$ *does not depend* on $\epsilon_{n+1}$:

$$
\begin{aligned}
\mathbb{Var}[\hat{\epsilon}_{n+1}] 
&= \mathbb{Var}[y_{n+1} - \hat{y}_{n+1}]\\
&= \mathbb{Var}[y_{n+1}] + \mathbb{Var}[\hat{y}_{n+1}]\\
\end{aligned}
$$

Hence:
$$
\begin{aligned}
\mathbb{Var}[\hat{\epsilon}_{n+1}] 
&= \sigma^2 + \mathbb{Var}[\mathbf{x}^{n+1}\hat{\boldsymbol{\beta}}]\\
&= \sigma^2 + \mathbf{x}^{n+1} \mathbb{Var}[\hat{\boldsymbol{\beta}}](\mathbf{x}^{n+1})^T\\
&= \sigma^2 + \mathbf{x}^{n+1} \sigma^2 (\mathbf{X}^T\mathbf{X})^{-1} (\mathbf{x}^{n+1})^T\\
\end{aligned}
$$

<!-- ************************************************************************ -->
# Geometrical Interpretation
<!-- ************************************************************************ -->

## Variance Decomposition

Assuming $\mathbb{1}$ is in the model:

```{r proj2, echo=FALSE, fig.height=3, fig.width=3, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
yv <- c(0.8, 1.0)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
ybaronev <- c(0.7, 0.2)
yhatv <- c(0.8, 0.3)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "red", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(y)), col = "red", pos = 2)
# X
arrows(x0 = or[1], y0 = or[2], x1 = xv[1], y1 = xv[2],
       length = 0.1, col = "black", lwd = 2)
text(xv[1], xv[2], labels = expression(bold(x)), col = "black", pos = 1)
# one
arrows(x0 = or[1], y0 = or[2], x1 = onev[1], y1 = onev[2],
       length = 0.1, col = "black", lwd = 2)
text(onev[1], onev[2], labels = expression(bold("1")), col = "black", pos = 2)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "blue", lwd = 2)
text(yhatv[1], yhatv[2],
     labels = expression(hat(bold(y))),
     col = "blue", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "black", lwd = 1, lty = 2)
st <- 0.03
segments(x0 = yhatv[1], y0 = yhatv[2] + st,
         x1 = yhatv[1] - st*5/sqrt(26), y1 = yhatv[2] + st - st/sqrt(26),
         col = "black", lwd = 1)
segments(x0 = yhatv[1] - st*5/sqrt(26), y0 = yhatv[2] - st/sqrt(26),
         x1 = yhatv[1] - st*5/sqrt(26), y1 = yhatv[2] + st - st/sqrt(26),
         col = "black", lwd = 1)
# hat epsilon
text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
     labels = expression(hat(bold(epsilon))), col = "blue", pos = 4)
# beta x
# segments(x0 = betaonev[1], y0 = betaonev[2], x1 = yhatv[1], y1 = yhatv[2],
         # col = "blue", lwd = 1)
# text((betaonev[1] + yhatv[1])/2, (betaonev[2] + yhatv[2])/2,
     # labels = expression(paste(hat(beta)[1], bold(x))), col = "blue", pos = 3)
# beta one
arrows(x0 = yhatv[1], y0 = yhatv[2], x1 = ybaronev[1], y1 = ybaronev[2],
       length = 0.1, col = "blue", lwd = 2)
text(ybaronev[1], ybaronev[2],
     labels = expression(-paste(bar(y), bold("1"))), col = "blue", pos = 2)
# segments(x0 = or[1], y0 = or[2], x1 = betaxv[1], y1 = betaxv[2],
#          col = "black", lwd = 1, lty = 2)
```

Using Pythagore:

$$
\begin{aligned}
\|\mathbf{y} - \bar{y} \mathbb{1}\|^2 
&= \|\hat{\mathbf{y}} - \bar{y} \mathbb{1} + \mathbf{y} - \hat{\mathbf{y}}\|^2 
= \|\hat{\mathbf{y}} - \bar{y} \mathbb{1} + \hat{\boldsymbol{\epsilon}}\|^2 \\
&= \|\hat{\mathbf{y}} - \bar{y} \mathbb{1}\|^2  + \|\hat{\boldsymbol{\epsilon}}\|^2 
\end{aligned}
$$

## Variance Decomposition

$$
\begin{aligned}
&\|\mathbf{y} - \bar{y} \mathbb{1}\|^2 
&=&&& \|\hat{\mathbf{y}} - \bar{y} \mathbb{1}\|^2  &&+& \|\hat{\boldsymbol{\epsilon}}\|^2 \\
&TSS &=&&& ESS &&+& RSS
\end{aligned}
$$

* **TSS**: Total Sum of Squares  
$\to$ Amount of variability in the response $\mathbf{y}$ *before* the regression is performed.
* **RSS**: Residual Sum of Squares  
$\to$ Amount of variability that is left *after* performing the regression.
* **ESS**: Explained Sum of Squares  
$\to$ Amount of variability that is *explained* by the regression.

## $R^2$ Statistic

$$
R^2 = \frac{ESS}{TSS} 
= \frac{\|\hat{\mathbf{y}} - \bar{y} \mathbb{1}\|^2}{\|\mathbf{y} - \bar{y} \mathbb{1}\|^2}
= 1 - \frac{\|\hat{\epsilon}\|^2}{\|\mathbf{y} - \bar{y} \mathbb{1}\|^2}
= 1 - \frac{RSS}{TSS}
$$

* $R^2$ is the **proportion** of variability in $\mathbf{y}$ that can be explained by the regression.

>* $0 \leq R^2 \leq 1$

>* $R^2$ = 1: the regression is perfect.  
$\to$ $\mathbf{y} = \hat{\mathbf{y}}$ and $\mathbf{y}$ is in $\mathcal{M}(\mathbf{x})$.

>* $R^2$ = 0: the regression is useless.  
$\to$ $\hat{\mathbf{y}} = \bar{y}\mathbb{1}$, the empirical mean is sufficient.

## $R^2$ Statistic

$$
R^2 = \frac{ESS}{TSS} 
= \frac{\|\hat{\mathbf{y}} - \bar{y} \mathbb{1}\|^2}{\|\mathbf{y} - \bar{y} \mathbb{1}\|^2}
= \cos^2(\theta)
$$

```{r proj3, echo=FALSE, fig.height=4, fig.width=4, fig.align='center'}
par(mar = c(0, 0, 0, 0) + 0.1)
plot.new()
plot.window(c(0, 1.1), c(0, 1.1))
polygon(x = c(0, 0.6, 1, 0.4), y = c(0.2, 0, 0.4, 0.6), col = "gray90")
or <- c(0.3, 0.2)
ybaronev <- c(0.5, 0.4)
yv <- c(0.8, 1.0)
xv <- c(0.5, 0.1 + 0.1/3)
betaxv <- c(0.6, 0.1)
onev <- c(0.6, 0.5)
betaonev <- c(0.5, 0.4)
# ybaronev <- c(0.7, 0.2)
yhatv <- c(0.8, 0.3)
# Y
arrows(x0 = or[1], y0 = or[2], x1 = yv[1], y1 = yv[2],
       length = 0.1, col = "red", lwd = 2)
text(yv[1], yv[2], labels = expression(bold(y)), col = "red", pos = 2)
# X
arrows(x0 = or[1], y0 = or[2], x1 = xv[1], y1 = xv[2],
       length = 0.1, col = "black", lwd = 2)
text(xv[1], xv[2], labels = expression(bold(x)), col = "black", pos = 1)
# one
arrows(x0 = or[1], y0 = or[2], x1 = onev[1], y1 = onev[2],
       length = 0.1, col = "black", lwd = 2)
text(onev[1], onev[2], labels = expression(bold("1")), col = "black", pos = 4)
# yhat
arrows(x0 = or[1], y0 = or[2], x1 = yhatv[1], y1 = yhatv[2],
       length = 0.1, col = "blue", lwd = 2)
text(yhatv[1], yhatv[2],
     labels = expression(hat(bold(y))),
     col = "blue", pos = 4)
# y - yhat
segments(x0 = yv[1], y0 = yv[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "black", lwd = 1, lty = 2)
st <- 0.03
segments(x0 = yhatv[1], y0 = yhatv[2] + st,
         x1 = yhatv[1] - st*5/sqrt(26), y1 = yhatv[2] + st - st/sqrt(26),
         col = "black", lwd = 1)
segments(x0 = yhatv[1] - st*5/sqrt(26), y0 = yhatv[2] - st/sqrt(26),
         x1 = yhatv[1] - st*5/sqrt(26), y1 = yhatv[2] + st - st/sqrt(26),
         col = "black", lwd = 1)
# hat epsilon
text((yv[1] + yhatv[1])/2, (yv[2] + yhatv[2])/2,
     labels = expression(hat(bold(epsilon))), col = "blue", pos = 4)
# beta x
# segments(x0 = betaonev[1], y0 = betaonev[2], x1 = yhatv[1], y1 = yhatv[2],
         # col = "blue", lwd = 1)
# text((betaonev[1] + yhatv[1])/2, (betaonev[2] + yhatv[2])/2,
     # labels = expression(paste(hat(beta)[1], bold(x))), col = "blue", pos = 3)
# beta one
arrows(x0 = or[1], y0 = or[2], x1 = ybaronev[1], y1 = ybaronev[2],
       length = 0.1, col = "blue", lwd = 2)
text(ybaronev[1], ybaronev[2],
     labels = expression(paste(bar(y), bold("1"))), col = "blue", pos = 2)
# theta
segments(x0 = yv[1], y0 = yv[2], x1 = ybaronev[1], y1 = ybaronev[2],
         col = "darkblue", lwd = 3, lty = 3)
segments(x0 = ybaronev[1], y0 = ybaronev[2], x1 = yhatv[1], y1 = yhatv[2],
         col = "darkblue", lwd = 3, lty = 3)
segments(x0 = ybaronev[1] + (xv[1] - or[1])/5, y0 = ybaronev[2] + (xv[2] - or[2])/5,
         x1 =  ybaronev[1] + (yv[1] - ybaronev[1])/5, y1 = ybaronev[2] + (yv[2] - ybaronev[2])/5,
         col = "darkblue", lwd = 3, lty = 3)
text(0.61, 0.43,
     labels = expression(bold(theta)), col = "darkblue", pos = 2)
```

## $R^2$ Statistic

$$
\begin{aligned}
TSS &= \|\mathbf{y} - \bar{y} \mathbb{1}\|^2 = \sum_{i = 1}^n (y_i - \bar{y})^2 & \text{"total inertia"}\\
RSS &= \|\mathbf{y} - \hat{\mathbf{y}}\|^2 = \sum_{i = 1}^n (y_i - \hat{y}_i)^2 & \text{"intra-class inertia"}\\
ESS &= \|\hat{\mathbf{y}} - \bar{y} \mathbb{1}\|^2 = \sum_{i = 1}^n (\hat{y}_i - \bar{y})^2 & \text{"inter-class inertia"}
\end{aligned}
$$
$$
~
$$

$$
R^2 = \frac{ESS}{TSS} = 1 - \frac{RSS}{TSS}
$$


## $R^2$ Statistic - Issue - 1/3

$$
y_i = -2 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

```{r rsq1, echo=TRUE}
set.seed(12890926)

## Predictors
n <- 100
x_1 <- runif(n, min = -2, max = 2)
x_2 <- runif(n, min = 0, max = 4)

## Noise
eps <- rnorm(n, mean = 0, sd = 5)

## Model sim
beta_0 <- -2; beta_1 <- 3; beta_2 <- -1
y_sim <- beta_0 + beta_1 * x_1 + beta_2 * x_2 + eps
```

## $R^2$ Statistic - Issue - 2/3 {.build}

$$
y_i = -2 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

```{r rsq2, echo=TRUE}
fit <- lm(y_sim ~ x_1 + x_2)
summary(fit)$r.squared
```

```{r rsq3, echo=TRUE}
## unrelated noise variable x_3
x_3 <- runif(n, min = -4, max = 0)
## Fit
fit2 <- lm(y_sim ~ x_1 + x_2 + x_3)
summary(fit2)$r.squared
```

**The more there are predictors, the better $R^2$ is !**

## $R^2$ Statistic - Issue - 3/3

$$
R^2 = 1 - \frac{RSS}{TSS}
= 1 - \frac{\|\hat{\epsilon}\|^2}{\|\mathbf{y} - \bar{y} \mathbb{1}\|^2}
= 1 - \frac{\|\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}}\|^2}{\|\mathbf{y} - \bar{y} \mathbb{1}\|^2}
$$

If $\mathbf{X}' = (\mathbf{X} ~ \mathbf{x}_{p+1})$ has one more row, then:
$$
\begin{aligned}
\|\mathbf{y} - \mathbf{X}'\hat{\boldsymbol{\beta}}'\|^2
& = \underset{\boldsymbol{\beta}' \in \mathbb{R}^{p+1}}{\operatorname{min}} 
\|\mathbf{y} - \mathbf{X}'\boldsymbol{\beta}' \|^2\\
& = \underset{\boldsymbol{\beta} \in \mathbb{R}^p\\ \beta_{p+1} \in \mathbb{R}}{\operatorname{min}} 
\|\mathbf{y} - \mathbf{X}\boldsymbol{\beta} - \mathbf{x}_{p+1}\beta_{p+1} \|^2\\
& \leq \underset{\boldsymbol{\beta} \in \mathbb{R}^p}{\operatorname{min}} 
\|\mathbf{y} - \mathbf{X}\boldsymbol{\beta}\|^2
\end{aligned}
$$

$$
\text{Hence:}\quad
\|\mathbf{y} - \mathbf{X}'\hat{\boldsymbol{\beta}}'\|^2 \leq \|\mathbf{y} - \mathbf{X}\hat{\boldsymbol{\beta}}\|^2
$$


## Adjusted $R^2$ Statistic

$$
\begin{aligned}
R_a^2 
&= 1 - \frac{RSS / (n-p)}{TSS / (n-1)} 
= 1 - \frac{(n-1) RSS}{(n-p) TSS} \\
&= 1 - \frac{n-1}{n-p} (1 - R^2)
\end{aligned}
$$

* **Penalize** for the number of predictors $p$.

* **Attention** $p$ includes the intercept (rank of matrix $\mathbf{X}$).

## Adjusted $R^2$ Statistic

$$
y_i = -1 + 3 x_{i1} - x_{i2} + \epsilon_i
$$

```{r rsq4, echo=TRUE}
fit <- lm(y_sim ~ x_1 + x_2)
summary(fit)$adj.r.squared
```

```{r rsq5, echo=TRUE}
## unrelated noise variable x_3 and x_4
x_3 <- runif(n, min = -4, max = 0)
## Fit
fit2 <- lm(y_sim ~ x_1 + x_2 + x_3)
summary(fit2)$adj.r.squared
```

<!-- ************************************************************************ -->
# Advertising
<!-- ************************************************************************ -->

## Advertising {.smaller}

```{r ad3}
fit_all <- lm(sales ~ TV + radio + newspaper, data = ad)
summary(fit_all)
```

<!-- ************************************************************************ -->
# To be continued
<!-- ************************************************************************ -->

## Questions

* Confidence interval for $(\hat{\boldsymbol{\beta}}, \hat{\sigma}^2)$ ?

* Can we test $\hat{\boldsymbol{\beta}} = 0$ (i.e. no linear trend) ?

>* Assumptions on the moments are not enough.  
$\hookrightarrow$ We need assumptions on the *specific distribution* of the $\epsilon_i$.

>* Most common assumption: $\epsilon_i$ are **Gaussian**.
